---
title: "JUDGE channel — I. Descriptives & response processes"
subtitle: "Implicit personality theory in 9 LLMs, read as a rating instrument"
author: "ecb"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 9
    fig_height: 5.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 110)
source("_common.R")
```

## What this data is, and the stance I take on it

The **JUDGE** channel (rgb's `tom_likely`) asks each model, for every ordered pair of
523 person-descriptive adjectives *(i, j)*, *"how strongly would a person described as
**i** also be **j**?"* and reads the answer as an **expected rating on a 1–7 Likert
scale** from the full log-prob distribution over the rating tokens (not the argmax).
Each model therefore produces a `523 × 523` matrix **B** of conditional-endorsement
ratings (row-conditioned, *not* symmetric, diagonal not collected), plus a matching
matrix **Hent** of the Shannon entropy of each rating distribution — a per-judgment
uncertainty.

rgb's own analyses of this data (W15–W16) are **geometric and mechanistic**: the
read/write dissociation, a Helmholtz–Hodge decomposition of the asymmetry, factor
congruence to the human 525-PDA. This report series takes the **psychometric** view
the geometry brackets: *if we treat these judgments as measurements, are they sound?*
The unit I adopt (per ecb's steer) is **models as raters** — here **9 judges** rating the
same 273,006 ordered pairs. (This reanalysis is built on rgb's **full 7-category raw
distributions**, `judge_dists_full` batch 1, which cover 9 of the 12-model cohort; the raw
data for the other three — `FalconMamba`, `Gemma-4`, `Qwen-32` — are pending in batch 2,
so those raters drop out here. Earlier versions of these reports used a 12-model EV-only
release; the raw distributions supersede it — see the caveat below.) Reliability, validity
and response-process questions all follow from the models-as-raters frame.

This first report is the descriptive floor: **how does each rater use the scale?**
Everything downstream (what to do about asymmetry; whether the raters agree; whether the
structure is interpretable) is conditioned by the response processes documented here.

> **Data provenance, stated once — and it shapes the method here.** This reanalysis works
> from the **full 7-category rating distributions** `dists` (523×523×7 per model), from
> which I recompute `B` (the expected rating) and `Hent` (the entropy) in float64 so the
> two moments are internally consistent (the entropy envelope of §1B is respected by
> construction). These raw distributions **supersede** the earlier EV-only release, whose
> bf16 numerics had drifted (mean |ΔEV| 0.02–0.09, with bimodal cells' EV flipping by up
> to ~4). Because §§1–5 are framed on the two moments (`B`, `Hent`) — the summary a rater's
> *location* and *decisiveness* are captured by — I keep that framing throughout for
> continuity and because it is the right altitude for the response-style story; the
> **spread index** (§1B) places each judgment's entropy between the minimum and maximum
> attainable at its mean, a mean-detrended proxy for how diffuse the underlying
> distribution is. But two moments cannot see *shape* (an EV of 4 from "all mass on 4" and
> from "half on 1, half on 7" are opposite response processes that round identically), and
> **now we can**: **§6 uses the full 523² raw distributions** to give the honest category
> usage, floor/ceiling, and a shape census — and it **corrects the earlier medoid-based
> verdict** that the two-moment proxy was faithful for 10/12 models. `Phi4` and `Aya` are
> *not* summarized by EV alone; the earlier error was one of *metric* (reading faithfulness
> off strict bimodality) rather than the medoid sample, which in fact tracked off-mode mass
> reasonably well.

```{r load}
summ <- read_csv_("model_summary.csv") %>% as_tibble() %>%
  mutate(display = factor(display, levels = disp_ord))
marg <- read_csv_("marginals.csv") %>% as_tibble()
dir_ev <- fread_gz("dir_ev.csv.gz")     # i,j, <9 models>  (273006 directional EVs)
dir_h  <- fread_gz("dir_hent.csv.gz")
# long form for distributional plots
ev_long <- melt(dir_ev, id.vars = c("i","j"), variable.name = "model", value.name = "ev")
ev_long[, display := factor(DISPLAY[as.character(model)], levels = disp_ord)]
ev_long[, family  := FAMILY[as.character(model)]]
h_long <- melt(dir_h, id.vars = c("i","j"), variable.name = "model", value.name = "hent")
```

## 1. Scale usage: some raters never leave the middle

The single most consequential descriptive fact is that the 9 models **do not use the
1–7 scale in remotely the same way.** The table sorts by mean rating (leniency); note
especially the spread column (`ev_sd`) and the EV tail fractions (share of judgments with
an expected rating in each band).

```{r usage-table}
tab <- summ %>% arrange(ev_mean) %>%
  transmute(Model = display, Family = family, `Params (B)` = params_b,
            `Mean` = round(ev_mean,2), `SD` = round(ev_sd,2),
            `Skew` = round(ev_skew,2), `Excess kurt.` = round(ev_kurtosis,2),
            `% ≤2` = round(100*frac_le2,1), `% 3–5` = round(100*frac_mid_3to5,1),
            `% ≥6` = round(100*frac_ge6,1))
knitr::kable(tab, caption = "Per-rater location, spread and tail usage of the 1–7 scale.")
```

### 1A. The distribution of the expected rating

Rather than round `B` into pseudo-categories, plot its actual distribution per rater —
the density of the expected-rating statistic itself. This shows leniency (location) and
range restriction (spread) directly and honestly.

```{r ev-density, fig.height=5}
ggplot(ev_long, aes(ev)) +
  geom_histogram(aes(y = after_stat(density), fill = family), bins = 60) +
  facet_wrap(~display, ncol = 4) +
  scale_fill_manual(values = FAM_PAL, guide = "none") +
  scale_x_continuous(breaks = 1:7, limits = c(1,7)) +
  labs(title = "Distribution of the expected rating B, per rater",
       x = "expected rating (1–7)", y = "density") +
  theme_judge()
```

Read as a **response-style census**:

- **Range-restricted / central responder.** `Llama3.2-3B`'s expected ratings collapse
  onto a narrow 4–5 band (SD = 0.46) — whatever the pair, the answer is "middling". It is
  the sole flat rater in this batch (`FalconMamba`, the other range-restricted model in the
  full cohort, is in the pending batch 2).
- **Full-range responders with ceiling/floor mass.** `Aya-expanse-8B` and `Phi4-mini`
  spread expected ratings across the whole scale with heavy mass toward the ceiling
  (SD ≈ 1.7–1.9) — §6 shows these two are the models whose raw distributions are genuinely
  diffuse (real mass at both ends), not just wide-ranging means.
- **Gemma family** is central-to-high and, on the raw data (§6), the most *concentrated*.
- **Qwen family** sits modal around 3–4 (the most parsimonious raters, mean 3.6–3.7) with
  intermediate spread.

### 1B. Spread index: are the *underlying* distributions concentrated or diffuse?

The expected rating hides the shape of the distribution it came from. Since we also hold
each judgment's entropy, we can recover *how concentrated* the underlying rating
distribution is, **detrended for its mean** — the raw entropy mechanically falls near the
scale ends, so a bare entropy comparison confounds concentration with leniency. The
**spread index** places each judgment's entropy between the minimum and maximum entropy
attainable at its mean (0 = as concentrated as any distribution with that mean can be;
1 = maximally diffuse). It is the closest honest proxy to the raw response shape.

```{r spread, fig.height=5}
ev_long[, spread := spread_index(ev, h_long$hent)]
sp_summ <- ev_long[, .(mean_spread = mean(spread, na.rm = TRUE)), by = display]
ggplot(ev_long[!is.na(spread)][sample(.N, 120000)], aes(spread, y = display, fill = after_stat(x))) +
  ggridges::geom_density_ridges_gradient(scale = 2.2, rel_min_height = 0.01, linewidth = 0.2) +
  scale_fill_viridis_c(name = "spread", option = "C") +
  labs(title = "Spread index by rater (mean-detrended concentration of the rating distribution)",
       subtitle = "left = concentrated / decisive · right = diffuse / hedged",
       x = "spread index", y = NULL) +
  theme_judge() + theme(legend.position = "right")
knitr::kable(
  sp_summ %>% arrange(mean_spread) %>% transmute(Model = display, `Mean spread` = round(mean_spread,3)),
  caption = "Mean spread index per rater — a mean-independent response-style axis.")
```

This **re-sorts the cohort on a genuinely different axis than leniency or SD**: the
Gemma family produces the most *concentrated* underlying distributions (mean spread
0.06–0.11 — it commits to sharp judgments) while `Phi4-mini` is by far the most *diffuse*
(0.94 — its judgments are broad even where the expected rating is mid-scale), with
`Llama3.1-8B` and `Aya` next (0.57–0.65). Note this is not the same as the SD of `B`: a rater can produce a wide range
of expected ratings out of individually *diffuse* distributions (`Phi4`) or a narrow
range out of *sharp* ones. Concentration and differentiation are distinct response-style
dimensions — the point §4 formalizes.

This is exactly the heterogeneity that makes cross-model *magnitude* comparison unsafe
(rgb's calibration caveat) — but stated in response-style terms it is sharper: **these
are different rating styles, not just different means.** Range restriction limits how
much *differentiated information* a rater can carry (it caps the achievable
between-pair variance) — though, importantly, Report III shows it does **not** make these
raters disagree with the crowd: a flat rater is low-information, not noisy.

## 2. Location and spread are only loosely coupled

```{r loc-spread}
ggplot(summ, aes(ev_mean, ev_sd, colour = family, label = display)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20, seed = 1) +
  scale_colour_manual(values = FAM_PAL, name = NULL) +
  labs(title = "Rater leniency (mean) vs differentiation (SD)",
       x = "mean rating (leniency →)", y = "SD of ratings (differentiation →)") +
  theme_judge()
```

Leniency (mean 3.64 → 4.83) and differentiation (SD 0.46 → 1.95) vary independently:
`Llama3.2-3B` is *lenient but flat* (high mean, tiny SD — it agrees warmly with almost
everything), whereas `Aya` is *lenient and spread*. A high mean does not buy you
usable variance. For the measurement, **differentiation (SD) is the resource** — a rater
with SD 0.47 can encode at most a sliver of true between-pair signal.

## 3. Decisiveness: the entropy channel is a second, partly independent axis

Each judgment carries an entropy. Averaged per rater, this is a **decisiveness** index
that is *not* the same thing as scale spread.

```{r entropy-violin, fig.height=5}
ggplot(h_long, aes(x = factor(DISPLAY[as.character(model)], levels = disp_ord), y = hent,
                   fill = FAMILY[as.character(model)])) +
  geom_violin(scale = "width", linewidth = 0.2) +
  scale_fill_manual(values = FAM_PAL, name = NULL) +
  coord_flip() +
  labs(title = "Per-judgment entropy (uncertainty) by rater",
       x = NULL, y = "Shannon entropy of the 1–7 rating distribution (nats)") +
  theme_judge()
```

```{r ev-entropy-curve, fig.height=4.5}
# entropy as a function of the rating level: the response-process signature
set.seed(1)
samp <- ev_long[sample(.N, 60000)][, .(i,j,model,ev)]
samp <- merge(samp, h_long[, .(i,j,model,hent)], by = c("i","j","model"))
samp[, display := factor(DISPLAY[as.character(model)], levels = disp_ord)]
ggplot(samp, aes(ev, hent)) +
  geom_hex(bins = 40) + scale_fill_viridis_c(trans = "log10", name = "pairs") +
  facet_wrap(~display, ncol = 4, scales = "free_y") +
  labs(title = "Entropy vs expected rating — the graded-in-the-middle signature",
       x = "expected rating", y = "entropy") +
  theme_judge() + theme(legend.position = "right")
```

Two things:

1. **The inverted-U** (entropy peaks at mid-scale, collapses at the extremes) is present
   for the discriminating raters — a *sensible* response process: the model is certain
   when it judges a pair clearly (dis)similar and hedges in between.
2. **The flat rater breaks it.** `Llama3.2-3B` sits at moderate entropy (mean ≈ 1.09 nats)
   almost everywhere and is essentially *never* decisive — only 0.002% of its judgments
   have entropy < 0.1, versus 49–67% for the Gemma family. Its narrow EV in §1 is now
   explained: it is not compressing a confident signal, it is **reporting broadish
   distributions whose means all sit mid-scale.** (`FalconMamba`, the cohort's other
   near-uniform responder, is in the pending batch 2.) Note the two *highest*-entropy
   raters here are `Phi4-mini` and `Llama3.1-8B` (≈ 1.26 nats) — but these *do*
   differentiate (high EV-SD), so their diffuseness is a wide-then-hedged style, not
   non-response; §6 confirms `Phi4` in particular carries real off-mode mass.

## 4. A response-process typology: two kinds of "uncertain"

Crossing scale differentiation (SD of `B`) against concentration (mean spread index,
§1B — the mean-detrended decisiveness) separates raters that look similar on either axis
alone.

```{r typology}
typ <- summ %>% left_join(sp_summ, by = "display")
ggplot(typ, aes(mean_spread, ev_sd, colour = family, label = display)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20, seed = 2) +
  scale_colour_manual(values = FAM_PAL, name = NULL) +
  geom_vline(xintercept = median(typ$mean_spread), linetype = 3, colour = "grey60") +
  geom_hline(yintercept = median(typ$ev_sd), linetype = 3, colour = "grey60") +
  labs(title = "Response-process typology",
       subtitle = "up-left = confident differentiator · down-right = diffuse non-discriminator",
       x = "mean spread index (← concentrated/decisive | diffuse →)",
       y = "SD of expected ratings (↑ differentiated)") +
  theme_judge()
```

- **Confident differentiators** (low spread, high SD): the Gemma family, esp.
  `Gemma3-27B`/`Gemma3-12B`. They commit to sharp, well-separated judgments — §6 confirms
  they are ≥98% unimodal on the raw data.
- **Low-information corner** (low SD): `Llama3.2-3B` — little between-pair signal, and only
  mid-spread (its distributions are broadish, not sharp). It is the sole occupant now that
  `FalconMamba` is in batch 2.
- **Spread-but-hedging** (high spread, high SD): `Aya`, `Phi4-mini` — they use the whole
  scale yet keep each individual distribution diffuse. *(§6, on the full raw distributions,
  shows these two are exactly where the two-moment proxy fails: `Phi4` puts ≥0.25 off-mode
  mass in **56%** of cells and `Aya` in **26%**, and `Aya` makes genuine no-OR-yes
  **mixtures** in ~15% of judgments — invisible to EV+entropy.)*

This typology is the load-bearing setup for reliability — but with a twist that
Report III makes precise and that is easy to get wrong: a **diffuse non-discriminator
carries little *differentiated* information** (its between-pair variance is small), yet
it still **agrees with the consensus** and does not inject noise (its judgments are a
smooth, prevalence-tracking, compressed image of the crowd's). So the flat raters are
*low-information*, not *low-quality*; the raters that actually pull away from the
consensus turn out to be the high-variance *idiosyncratic* models, not the flat ones.
Weighting/trimming therefore buys much less than the typology first suggests (Report III
§3, Report V).

## 5. Marginal structure: base rates (prevalence) and anchor generosity

Because the diagonal (self-pairs) was not collected, each rater's matrix has two
informative margins:

- **column mean** = how strongly adjective *j* is ascribed *given any anchor* — a
  **base rate / prevalence** ("Awake", "Normal" are ascribed to almost anyone).
- **row mean** = how generously anchor *i* ascribes *other* traits — anchor-level
  acquiescence.

```{r marginals, fig.height=4.5}
mm <- marg %>% mutate(display = factor(DISPLAY[model], levels = disp_ord))
ggplot(mm, aes(col_mean, fill = FAMILY[model])) +
  geom_histogram(bins = 40) + facet_wrap(~display, ncol = 4, scales = "free_y") +
  scale_fill_manual(values = FAM_PAL, guide = "none") +
  labs(title = "Target base-rate (column-mean) distribution per rater",
       x = "mean ascription of a target across all anchors (prevalence)", y = "count") +
  theme_judge()
```

```{r prevalence-extremes}
cohort_prev <- mm %>% group_by(adjective) %>% summarise(prev = mean(col_mean)) %>%
  arrange(desc(prev))
knitr::kable(
  bind_cols(
    cohort_prev %>% slice_head(n = 12) %>% transmute(`Most universally ascribed` = adjective, `μ` = round(prev,2)),
    cohort_prev %>% slice_tail(n = 12) %>% arrange(prev) %>% transmute(`Least ascribed` = adjective, ` μ` = round(prev,2))
  ), caption = "Cohort-mean target base rates (prevalence marginals).")
```

The prevalence marginal is large and structured — and it is precisely what a
marginal-free human *correlation* matrix does not contain, which is why the reliability
and validity reports **double-center** (strip additive row/column marginals) before
treating `B` as a similarity. It is also the primary axis of the asymmetry (Report II).

## 6. The full raw distributions: what the two moments actually hide

Sections 1–5 read each judgment through its two moments (`B`, `Hent`). We now hold the
**full 523×523×7 rating distributions** for all 9 batch-1 raters, so we can replace the
proxy with ground truth over the *entire* pair space — honest category usage, floor and
ceiling mass, and a distribution-shape census — and settle, at cohort scale, what the two
moments summarize well and what they miss.

This **corrects an earlier verdict** — but not in the way the internal handoff note
anticipated, and the difference is itself the finding. A prior version of this section used
a 35-adjective *facet-medoid* sample and concluded the `(EV, entropy)` proxy was "faithful
for 10/12 models." The natural hypothesis was that medoids — facet *exemplars*, i.e.
central prototypical cases — under-represent the diffuse judgments in the messy middle of
the 523² space, so the full set would look far more spread. **The data only partly bears
that out** (§6.1), and the real error turns out to be one of *metric*, not *sample* (§6.3).

```{r full-load}
shp  <- read_csv_("full_shape.csv") %>% as_tibble() %>%
  mutate(display = factor(DISPLAY[model], levels = disp_ord))
cm   <- read_csv_("full_catmass.csv") %>% as_tibble() %>%
  mutate(display = factor(display, levels = disp_ord), cat = factor(cat))
```

### 6.1 The medoid sample was a *reasonable* proxy for off-mode mass

Recomputing rgb's **off-mode-mass** metric (fraction of off-diagonal cells that place ≥ 0.25
probability ≥ 2 categories away from the modal rating) on the 35-medoid sample vs the full
523² set:

```{r medoid-vs-full}
offmode_frac <- function(P) {           # P: matrix rows = 7-cat pmfs
  m <- max.col(P, ties.method = "first")
  far <- abs(col(P) - m) >= 2
  mean(rowSums(P * far) >= 0.25)
}
md <- fread_gz("medoids_dists.csv.gz"); pcols <- paste0("p", 1:7)
md_off <- md[model %in% MODELS, .(medoid = offmode_frac(as.matrix(.SD))),
             by = model, .SDcols = pcols]
comp <- shp %>% transmute(model, full = offmode_frac) %>%
  left_join(md_off, by = "model") %>%
  transmute(Model = DISPLAY[model],
            `Off-mode % (35-medoid)` = round(100*medoid,1),
            `Off-mode % (full 523²)` = round(100*full,1),
            `Δ` = round(100*(full-medoid),1)) %>%
  arrange(desc(`Off-mode % (full 523²)`))
knitr::kable(comp, caption = "rgb's off-mode-mass metric, medoid sample vs full 523². The medoids track it well; the drift is a few points (the Llamas roughly double off a low base).")
```

Contrary to the "medoids badly under-represent spread" hypothesis, the medoid sample
**tracked off-mode mass reasonably**: `Phi4` reads 50.5% on the medoids vs 56.1% on the
full set, `Aya` 26.6% vs 26.2%. The exemplars do drift *slightly* low (both Llamas roughly
double, 2→7% and 5→10%, off small bases — consistent with central cases being modestly less
diffuse) but the **ranking and the qualitative story are preserved**. So the medoid sample
was not the problem. The blanket "faithful for 10/12" verdict is nonetheless **wrong for
`Phi4` and `Aya`** — and §6.3 shows *why* the earlier analysis missed it: it read
faithfulness off *strict bimodality* alone, when the operative failure is *off-mode mass*.

### 6.2 True category usage (the honest version of §1), cohort-wide

Actual mean probability mass per rating category over all 273,006 off-diagonal judgments —
no rounding, no proxy:

```{r full-catusage, fig.height=5}
ggplot(cm, aes(cat, mean_mass, fill = cat)) +
  geom_col() + facet_wrap(~display, ncol = 3) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1, guide = "none") +
  labs(title = "True mean category mass (full 523², 9 raters)",
       x = "rating category 1–7", y = "mean probability") + theme_judge()
```

The *location* story of §1 holds (Qwen is floor-leaning, Gemma central-to-high, Llama3.2
hugs 4–5), but the honest picture also shows how much end-mass `Aya` and `Phi4` carry
(mass on category 1 ≈ 0.18 / 0.15; on category 7 ≈ 0.39 / 0.37) — the raw signature of
diffuse, often U-shaped judgments that no expected rating can convey.

### 6.3 Two different questions: *off-mode mass* vs *genuine bimodality*

"Diffuse" and "bimodal" are not the same claim, and the full data lets us separate them —
which resolves what the off-mode metric can and can't say:

- **Off-mode mass** (rgb's metric, §6.1): how much probability sits ≥ 2 categories from the
  mode. High = the EV is a poor summary, but the distribution could be *broad-unimodal*.
- **Prominence bimodality** (`shape_class`, requiring a real valley < 0.6× the smaller of
  two peaks ≥ 0.15): a genuine two-mode "no OR yes" **semantic** mixture, distinguished
  from a **digit-notch** (a dip at a dispreferred digit 2/6 — a rating-channel artifact,
  echoing the Llama-2 category-2 suppression in the soft-evidence track).

```{r full-shape-census}
knitr::kable(
  shp %>% arrange(desc(semantic)) %>%
    transmute(Model = DISPLAY[model],
              `Off-mode %` = round(100*offmode_frac,1),
              `unimodal %` = round(100*unimodal,1),
              `semantic %` = round(100*semantic,1),
              `digit-notch %` = round(100*`digit-notch`,1),
              `adjacent %` = round(100*adjacent,1)),
  caption = "Off-mode mass vs a strict shape census over all 523². High off-mode need NOT mean bimodal.")
```

The census sharpens §6.1 into the real finding:

1. **`Phi4` is diffuse but *not* strongly bimodal.** Despite 56% off-mode mass it is **93%
   unimodal** by the prominence test — its judgments are *broad, flat-topped* distributions,
   not clean two-mode mixtures. Its EV genuinely mis-summarizes it, but the shape is
   spread, not split.
2. **`Aya` is the one genuinely bimodal rater** — **15% semantic** mixtures (empty-middle
   "almost certainly no, but sometimes yes" distributions), on top of its 26% off-mode
   mass. This is the case the two moments materially misread as a mid/high mean.
3. **The Gemma family and `Llama3.1-8B` are ≥ 98% unimodal** — for them EV + entropy lose
   essentially nothing, off-mode mass and all.
4. **Digit-2/6 suppression is real but minor**, concentrated in the Qwen family (`Qwen2.5-7B`
   4.4% digit-notch) and `Aya` (8%) — a channel artifact to keep in mind, not substantive
   structure.

```{r full-examples, fig.height=3.4}
ex <- read_csv_("full_examples.csv") %>% as_tibble()
exl <- ex %>%
  mutate(who = sprintf("%s: %s→%s  (EV=%.1f, %s)", display, adj_i, adj_j, ev, shape)) %>%
  select(who, shape, p1:p7) %>%
  pivot_longer(p1:p7, names_to = "cat", values_to = "p") %>%
  mutate(cat = as.integer(sub("p", "", cat)),
         who = factor(who, levels = unique(who)))
ggplot(exl, aes(factor(cat), p, fill = shape)) +
  geom_col(show.legend = FALSE) + facet_wrap(~who, ncol = 2) +
  scale_fill_manual(values = c(semantic = "#8172B3", unimodal = "#DD8452")) +
  labs(title = "Same expected rating, opposite response process",
       subtitle = "EV+entropy (and the spread index) cannot tell these apart; only the raw distribution can",
       x = "rating category", y = "probability") + theme_judge()
```

Each `Phi4`/`Aya` semantic mixture is paired with a matched-EV *unimodal* judgment from the
sharp `Gemma3-27B`: identical expected rating, opposite underlying process — the exact
failure mode the two-moment summary cannot see.

### 6.4 Verdict — and what it enables

The full distributions **correct the medoid-era conclusion**: the `(EV, entropy)` summary is
adequate for **7 of the 9 batch-1 raters** (the Gemmas, both Llamas, both Qwens), but **not
for `Phi4` or `Aya`**. Crucially, the spread index of §1B *did* flag both as diffuse — it
was doing its job — but a two-moment index **cannot distinguish broad-unimodal (`Phi4`)
from genuinely bimodal (`Aya`)**, and that distinction matters for what a judgment *means*.
The earlier verdict erred not because the 35-medoid sample was unrepresentative (§6.1 shows
it tracked off-mode mass well) but because it certified faithfulness off *strict bimodality*
alone — and `Phi4`, 93% unimodal yet 56% off-mode, slips exactly through that gap. The
lesson is methodological as much as empirical: pick the diffuseness metric that matches the
question.

Two forward implications:

- **The remaining raters (`FalconMamba`, `Gemma-4`, `Qwen-32`) are batch-2 pending**, so a
  cohort-complete shape census waits on that data. `FalconMamba` in particular — flagged as
  diffuse on the medoids — should be re-checked here: on the medoids its breadth read as a
  digit-notch, not bimodality, but that too may not survive the full set.
- **The full 7-way distributions open a soft-evidence treatment** (bridging ecb's earlier
  soft-evidence IRT work): distributional similarity (e.g. earth-mover distance between two
  judgment distributions) as an alternative to EV-cosine, and genuine confidence weighting
  from real distributions rather than an entropy proxy — most valuable exactly for the two
  raters (`Aya`, `Phi4`) the moments fail on.

## Takeaways for the measurement program

1. **Scale use is a rater property, not a constant.** Leniency, differentiation and
   decisiveness each vary ~4× across the cohort and are only loosely coupled — three
   separate response-style nuisances to model, not one calibration knob.
2. **`Llama3.2-3B` is range-restricted / low-information** (the sole flat rater in batch 1;
   `FalconMamba` is pending). It carries little differentiated signal — but Report III shows
   it still *agrees* with the consensus and does not harm it, so exclusion is **not**
   warranted; the ensemble is robust. (This corrects the tempting "down-weight the flat
   raters" intuition.)
3. **The entropy channel is real signal.** The inverted-U response process is sensible
   for discriminating raters; where it flattens, the rater is guessing. Entropy gives us
   a per-judgment reliability weight we will use.
4. **Big additive marginals (prevalence + anchor generosity)** must be removed before
   `B` is read as a similarity — motivating the symmetrization + double-centering in
   Reports II–IV.
5. **On the full raw distributions (§6), the two-moment summary is adequate for 7 of the 9
   batch-1 raters but fails for `Phi4` and `Aya`** — correcting the earlier "faithful for
   10/12" verdict, which erred by certifying faithfulness off *strict bimodality* alone
   (the 35-medoid sample actually tracked off-mode mass fine — `Phi4` 50%→56%). The two
   failures are *different*: `Phi4` is diffuse but
   **broad-unimodal** (56% off-mode, 93% unimodal), while `Aya` is **genuinely bimodal**
   (~15% empty-middle semantic mixtures). The spread index correctly flags both as diffuse
   but cannot tell those two shapes apart — which is exactly why the full distributions
   matter, and where a soft-evidence treatment should start. A minor digit-2/6 suppression
   artifact runs through the Qwen family and `Aya`.
```
