---
title: "JUDGE channel — I. Descriptives & response processes"
subtitle: "Implicit personality theory in 12 LLMs, read as a rating instrument"
author: "ecb"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 9
    fig_height: 5.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 110)
source("_common.R")
```

## What this data is, and the stance I take on it

The **JUDGE** channel (rgb's `tom_likely`) asks each model, for every ordered pair of
523 person-descriptive adjectives *(i, j)*, *"how strongly would a person described as
**i** also be **j**?"* and reads the answer as an **expected rating on a 1–7 Likert
scale** from the full log-prob distribution over the rating tokens (not the argmax).
Each model therefore produces a `523 × 523` matrix **B** of conditional-endorsement
ratings (row-conditioned, *not* symmetric, diagonal not collected), plus a matching
matrix **Hent** of the Shannon entropy of each rating distribution — a per-judgment
uncertainty.

rgb's own analyses of this data (W15–W16) are **geometric and mechanistic**: the
read/write dissociation, a Helmholtz–Hodge decomposition of the asymmetry, factor
congruence to the human 525-PDA. This report series takes the **psychometric** view
the geometry brackets: *if we treat these judgments as measurements, are they sound?*
The unit I adopt (per ecb's steer) is **models as raters** — 12 judges rating the same
273,006 ordered pairs. Reliability, validity and response-process questions all follow
from that frame.

This first report is the descriptive floor: **how does each rater use the scale?**
Everything downstream (what to do about asymmetry; whether the raters agree; whether the
structure is interpretable) is conditioned by the response processes documented here.

> **Data caveat, stated once — and it shapes the method here.** We hold the *expected
> value* `B` and the *entropy* `Hent` of each judgment's rating distribution — **not** the
> raw 7-category probabilities. rgb's generator computes the full distribution but
> persists only these two moments (`adjective_introspection.py`), so a true category
> histogram is not recoverable, and *rounding* `B` to a nearest category would be
> actively misleading (an EV of 4 from "all mass on 4" and from "half on 1, half on 7"
> are opposite response processes that round identically). I therefore characterize each
> judgment by **both moments jointly**: the distribution of the expected rating itself
> (§1A, an honest quantity), and a **spread index** (§1B) that places each judgment's
> entropy between the minimum and maximum attainable at its mean — the closest proxy to
> underlying-distribution *shape* that two moments allow. *rgb has since shared a
> 35-adjective sample of the raw distributions; **§6** uses it to validate this proxy
> against ground truth and to characterize exactly what it hides (genuine mixtures vs a
> digit artifact).* (Aside: ~0.5% of judgments have `Hent` fractionally below the
> theoretical minimum for their `B` — §6.1 shows this is a numerical/rounding residual,
> not a substantive support mismatch.)

```{r load}
summ <- read_csv_("model_summary.csv") %>% as_tibble() %>%
  mutate(display = factor(display, levels = disp_ord))
marg <- read_csv_("marginals.csv") %>% as_tibble()
dir_ev <- fread_gz("dir_ev.csv.gz")     # i,j, <12 models>  (273006 directional EVs)
dir_h  <- fread_gz("dir_hent.csv.gz")
# long form for distributional plots
ev_long <- melt(dir_ev, id.vars = c("i","j"), variable.name = "model", value.name = "ev")
ev_long[, display := factor(DISPLAY[as.character(model)], levels = disp_ord)]
ev_long[, family  := FAMILY[as.character(model)]]
h_long <- melt(dir_h, id.vars = c("i","j"), variable.name = "model", value.name = "hent")
```

## 1. Scale usage: some raters never leave the middle

The single most consequential descriptive fact is that the 12 models **do not use the
1–7 scale in remotely the same way.** The table sorts by mean rating (leniency); note
especially the spread column (`ev_sd`) and the EV tail fractions (share of judgments with
an expected rating in each band).

```{r usage-table}
tab <- summ %>% arrange(ev_mean) %>%
  transmute(Model = display, Family = family, `Params (B)` = params_b,
            `Mean` = round(ev_mean,2), `SD` = round(ev_sd,2),
            `Skew` = round(ev_skew,2), `Excess kurt.` = round(ev_kurtosis,2),
            `% ≤2` = round(100*frac_le2,1), `% 3–5` = round(100*frac_mid_3to5,1),
            `% ≥6` = round(100*frac_ge6,1))
knitr::kable(tab, caption = "Per-rater location, spread and tail usage of the 1–7 scale.")
```

### 1A. The distribution of the expected rating

Rather than round `B` into pseudo-categories, plot its actual distribution per rater —
the density of the expected-rating statistic itself. This shows leniency (location) and
range restriction (spread) directly and honestly.

```{r ev-density, fig.height=5}
ggplot(ev_long, aes(ev)) +
  geom_histogram(aes(y = after_stat(density), fill = family), bins = 60) +
  facet_wrap(~display, ncol = 4) +
  scale_fill_manual(values = FAM_PAL, guide = "none") +
  scale_x_continuous(breaks = 1:7, limits = c(1,7)) +
  labs(title = "Distribution of the expected rating B, per rater",
       x = "expected rating (1–7)", y = "density") +
  theme_judge()
```

Read as a **response-style census**:

- **Range-restricted / central responders.** `Llama3.2-3B`'s expected ratings collapse
  onto a narrow 4–5 band (SD = 0.47) — whatever the pair, the answer is "middling".
  `FalconMamba-7B` is similarly compressed into 3–5 (SD = 0.75).
- **Full-range / bimodal responders.** `Aya-expanse-8B` and `Phi4-mini` spread expected
  ratings across the whole scale with heavy mass toward the ceiling (SD ≈ 1.7–1.9); the
  Gemma family is visibly multi-modal.
- **Qwen family** sits modal around 3–4 with intermediate spread.

### 1B. Spread index: are the *underlying* distributions concentrated or diffuse?

The expected rating hides the shape of the distribution it came from. Since we also hold
each judgment's entropy, we can recover *how concentrated* the underlying rating
distribution is, **detrended for its mean** — the raw entropy mechanically falls near the
scale ends, so a bare entropy comparison confounds concentration with leniency. The
**spread index** places each judgment's entropy between the minimum and maximum entropy
attainable at its mean (0 = as concentrated as any distribution with that mean can be;
1 = maximally diffuse). It is the closest honest proxy to the raw response shape.

```{r spread, fig.height=5}
ev_long[, spread := spread_index(ev, h_long$hent)]
sp_summ <- ev_long[, .(mean_spread = mean(spread, na.rm = TRUE)), by = display]
ggplot(ev_long[!is.na(spread)][sample(.N, 120000)], aes(spread, y = display, fill = after_stat(x))) +
  ggridges::geom_density_ridges_gradient(scale = 2.2, rel_min_height = 0.01, linewidth = 0.2) +
  scale_fill_viridis_c(name = "spread", option = "C") +
  labs(title = "Spread index by rater (mean-detrended concentration of the rating distribution)",
       subtitle = "left = concentrated / decisive · right = diffuse / hedged",
       x = "spread index", y = NULL) +
  theme_judge() + theme(legend.position = "right")
knitr::kable(
  sp_summ %>% arrange(mean_spread) %>% transmute(Model = display, `Mean spread` = round(mean_spread,3)),
  caption = "Mean spread index per rater — a mean-independent response-style axis.")
```

This **re-sorts the cohort on a genuinely different axis than leniency or SD**: the
Gemma family produces the most *concentrated* underlying distributions (mean spread
0.05–0.11 — it commits to sharp judgments) while `Phi4-mini` and `FalconMamba-7B` are the
most *diffuse* (0.87–0.94 — their judgments are broad even where the expected rating is
mid-scale). Note this is not the same as the SD of `B`: a rater can produce a wide range
of expected ratings out of individually *diffuse* distributions (`Phi4`) or a narrow
range out of *sharp* ones. Concentration and differentiation are distinct response-style
dimensions — the point §4 formalizes.

This is exactly the heterogeneity that makes cross-model *magnitude* comparison unsafe
(rgb's calibration caveat) — but stated in response-style terms it is sharper: **these
are different rating styles, not just different means.** Range restriction limits how
much *differentiated information* a rater can carry (it caps the achievable
between-pair variance) — though, importantly, Report III shows it does **not** make these
raters disagree with the crowd: a flat rater is low-information, not noisy.

## 2. Location and spread are only loosely coupled

```{r loc-spread}
ggplot(summ, aes(ev_mean, ev_sd, colour = family, label = display)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20, seed = 1) +
  scale_colour_manual(values = FAM_PAL, name = NULL) +
  labs(title = "Rater leniency (mean) vs differentiation (SD)",
       x = "mean rating (leniency →)", y = "SD of ratings (differentiation →)") +
  theme_judge()
```

Leniency (mean 3.63 → 4.83) and differentiation (SD 0.47 → 1.95) vary independently:
`Llama3.2-3B` is *lenient but flat* (high mean, tiny SD — it agrees warmly with almost
everything), whereas `Aya` is *lenient and spread*. A high mean does not buy you
usable variance. For the measurement, **differentiation (SD) is the resource** — a rater
with SD 0.47 can encode at most a sliver of true between-pair signal.

## 3. Decisiveness: the entropy channel is a second, partly independent axis

Each judgment carries an entropy. Averaged per rater, this is a **decisiveness** index
that is *not* the same thing as scale spread.

```{r entropy-violin, fig.height=5}
ggplot(h_long, aes(x = factor(DISPLAY[as.character(model)], levels = disp_ord), y = hent,
                   fill = FAMILY[as.character(model)])) +
  geom_violin(scale = "width", linewidth = 0.2) +
  scale_fill_manual(values = FAM_PAL, name = NULL) +
  coord_flip() +
  labs(title = "Per-judgment entropy (uncertainty) by rater",
       x = NULL, y = "Shannon entropy of the 1–7 rating distribution (nats)") +
  theme_judge()
```

```{r ev-entropy-curve, fig.height=4.5}
# entropy as a function of the rating level: the response-process signature
set.seed(1)
samp <- ev_long[sample(.N, 60000)][, .(i,j,model,ev)]
samp <- merge(samp, h_long[, .(i,j,model,hent)], by = c("i","j","model"))
samp[, display := factor(DISPLAY[as.character(model)], levels = disp_ord)]
ggplot(samp, aes(ev, hent)) +
  geom_hex(bins = 40) + scale_fill_viridis_c(trans = "log10", name = "pairs") +
  facet_wrap(~display, ncol = 4, scales = "free_y") +
  labs(title = "Entropy vs expected rating — the graded-in-the-middle signature",
       x = "expected rating", y = "entropy") +
  theme_judge() + theme(legend.position = "right")
```

Two things:

1. **The inverted-U** (entropy peaks at mid-scale, collapses at the extremes) is present
   for the discriminating raters — a *sensible* response process: the model is certain
   when it judges a pair clearly (dis)similar and hedges in between.
2. **The non-responsive raters break it.** `FalconMamba` sits at entropy ≈ 1.66 nats
   almost everywhere (near the `log(7)=1.95` uniform ceiling) — it is *maximally
   uncertain on every pair*. `Llama3.2-3B` is similar (never decisive: only 0.002% of
   judgments have entropy < 0.1). Their narrow EV in §1 is now explained: they are not
   compressing a confident signal, they are **reporting near-flat distributions whose
   means happen to sit mid-scale.**

## 4. A response-process typology: two kinds of "uncertain"

Crossing scale differentiation (SD of `B`) against concentration (mean spread index,
§1B — the mean-detrended decisiveness) separates raters that look similar on either axis
alone.

```{r typology}
typ <- summ %>% left_join(sp_summ, by = "display")
ggplot(typ, aes(mean_spread, ev_sd, colour = family, label = display)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20, seed = 2) +
  scale_colour_manual(values = FAM_PAL, name = NULL) +
  geom_vline(xintercept = median(typ$mean_spread), linetype = 3, colour = "grey60") +
  geom_hline(yintercept = median(typ$ev_sd), linetype = 3, colour = "grey60") +
  labs(title = "Response-process typology",
       subtitle = "up-left = confident differentiator · down-right = diffuse non-discriminator",
       x = "mean spread index (← concentrated/decisive | diffuse →)",
       y = "SD of expected ratings (↑ differentiated)") +
  theme_judge()
```

- **Confident differentiators** (low spread, high SD): the Gemma family, esp.
  `Gemma-4-31B`/`Gemma3-27B`. They commit to sharp, well-separated judgments.
- **Diffuse non-discriminators** (high spread, low SD): `FalconMamba-7B` most clearly —
  broad distributions *and* little between-pair signal. `Llama3.2-3B` is the low-SD /
  low-information corner but only mid-spread. *(§6, on the raw distributions, shows
  FalconMamba's breadth is a smooth digit-notched **unimodal**, not bimodality — the
  spread index cannot see the difference.)*
- **Spread-but-hedging** (high spread, high SD): `Aya`, `Phi4-mini` — they use the whole
  scale yet keep each individual distribution diffuse. *(§6 shows `Aya` is special: ~11%
  of its judgments are genuine no-OR-yes **mixtures**, the one case the two-moment proxy
  materially misreads.)*

This typology is the load-bearing setup for reliability — but with a twist that
Report III makes precise and that is easy to get wrong: a **diffuse non-discriminator
carries little *differentiated* information** (its between-pair variance is small), yet
it still **agrees with the consensus** and does not inject noise (its judgments are a
smooth, prevalence-tracking, compressed image of the crowd's). So the flat raters are
*low-information*, not *low-quality*; the raters that actually pull away from the
consensus turn out to be the high-variance *idiosyncratic* models, not the flat ones.
Weighting/trimming therefore buys much less than the typology first suggests (Report III
§3, Report V).

## 5. Marginal structure: base rates (prevalence) and anchor generosity

Because the diagonal (self-pairs) was not collected, each rater's matrix has two
informative margins:

- **column mean** = how strongly adjective *j* is ascribed *given any anchor* — a
  **base rate / prevalence** ("Awake", "Normal" are ascribed to almost anyone).
- **row mean** = how generously anchor *i* ascribes *other* traits — anchor-level
  acquiescence.

```{r marginals, fig.height=4.5}
mm <- marg %>% mutate(display = factor(DISPLAY[model], levels = disp_ord))
ggplot(mm, aes(col_mean, fill = FAMILY[model])) +
  geom_histogram(bins = 40) + facet_wrap(~display, ncol = 4, scales = "free_y") +
  scale_fill_manual(values = FAM_PAL, guide = "none") +
  labs(title = "Target base-rate (column-mean) distribution per rater",
       x = "mean ascription of a target across all anchors (prevalence)", y = "count") +
  theme_judge()
```

```{r prevalence-extremes}
cohort_prev <- mm %>% group_by(adjective) %>% summarise(prev = mean(col_mean)) %>%
  arrange(desc(prev))
knitr::kable(
  bind_cols(
    cohort_prev %>% slice_head(n = 12) %>% transmute(`Most universally ascribed` = adjective, `μ` = round(prev,2)),
    cohort_prev %>% slice_tail(n = 12) %>% arrange(prev) %>% transmute(`Least ascribed` = adjective, ` μ` = round(prev,2))
  ), caption = "Cohort-mean target base rates (prevalence marginals).")
```

The prevalence marginal is large and structured — and it is precisely what a
marginal-free human *correlation* matrix does not contain, which is why the reliability
and validity reports **double-center** (strip additive row/column marginals) before
treating `B` as a similarity. It is also the primary axis of the asymmetry (Report II).

## 6. Raw-distribution validation (35-medoid sample)

Everything above works from two moments per judgment (`B`, `Hent`) because the raw
7-category distributions aren't in the main release. rgb has since shared a **sample of the
full distributions** — the complete `(35, 35, 7)` probability tensors for all 12 models on
35 human-derived **facet-medoid** adjectives, generated with the *exact* `tom_likely`
prompt. This lets us (a) check the spread-index machinery on ground truth and (b) see
directly what the two moments hide. *Caveats: it is a 35-adjective sample (facet
exemplars, not a random draw), and bf16 numerics drifted 0.01–0.08 mean-|ΔEV| from the
June full-matrix release — treat these `dists` as their own canonical object, not a
cell-by-cell match to the 523² `B`.*

```{r medoids-load}
md <- fread_gz("medoids_dists.csv.gz")
pcols <- paste0("p", 1:7); P <- as.matrix(md[, ..pcols])
md[, disp := DISPLAY[model]]
md[, `:=`(spread = spread_index(ev, entropy), shape = shape_class(P))]
```

### 6.1 The entropy envelope holds on internally-consistent data

Here `ev` and `entropy` are recomputed from the same `dists`, so they should be mutually
consistent up to numerical precision — and they are: the spread-index envelope (§1B) is
respected but for a small boundary residual.

```{r medoids-envelope}
md[, hmin := h_min_mean(ev)]
cat(sprintf("Envelope violations (entropy < theoretical min for its EV): %.2f%%\n",
            100 * mean(md$entropy < md$hmin - 1e-6)))
```

This **validates the spread-index construction**: the residual (~0.4%) sits at
near-integer means where float32 precision bites. Notably it is the *same magnitude* as
the main release's ~0.5% (§1 note) — so that inconsistency, too, is most likely numerical
(rgb's documented bf16 drift plus rounding), not a substantive `B`/`Hent` mismatch. Either
way, it does not affect the spread index materially.

### 6.2 True category usage (the honest version of §1)

With the raw distributions we can plot *actual* mean probability mass per category — no
rounding, no proxy — for this sample:

```{r medoids-catusage, fig.height=5}
cat_long <- md[, lapply(.SD, mean), by = disp, .SDcols = pcols] |>
  melt(id.vars = "disp", variable.name = "cat", value.name = "mass")
cat_long[, cat := factor(as.integer(gsub("p", "", cat)))]
cat_long[, disp := factor(disp, levels = disp_ord)]
ggplot(cat_long, aes(cat, mass, fill = cat)) +
  geom_col() + facet_wrap(~disp, ncol = 4) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1, guide = "none") +
  labs(title = "True mean category mass (35-medoid sample)",
       x = "rating category 1–7", y = "mean probability") + theme_judge()
```

The picture is consistent with the expected-rating story of §1 (Aya/Phi4 put real mass at
the extremes; the Gemma family is central-to-high; Llama3.2 hugs 4–5) — reassuring that
`B`'s distribution wasn't misleading us about *location*. What it could not tell us is
*shape*, which is §6.3.

### 6.3 What the two moments hide: genuine mixtures vs a digit artifact

Classifying each raw distribution's shape (a prominent second mode with a real valley
between peaks) separates two things `B`+`Hent` — and the spread index — cannot:

```{r medoids-shape}
shape_tab <- md[, .(
  `unimodal %`   = round(100*mean(shape == "unimodal"),1),
  `semantic %`   = round(100*mean(shape == "semantic"),1),
  `digit-notch %`= round(100*mean(shape == "digit-notch"),1),
  `adjacent %`   = round(100*mean(shape == "adjacent"),1)),
  by = .(disp)][order(-`semantic %`)]
knitr::kable(shape_tab, caption = "Distribution-shape census on the 35-medoid sample. 'semantic' = a true no-OR-yes mixture (peaks ≥3 apart); 'digit-notch' = a dip at a dispreferred digit (2 or 6).")
```

Three findings, and one **correction to §1/§3/§4**:

1. **Genuine bimodality is rare and concentrated in one model.** Only `Aya` shows
   meaningful *semantic* mixtures (~11% — clean "almost certainly no, but sometimes yes"
   distributions with an empty middle). `phi4` has a little; every other model is ≥97%
   unimodal. So for **10 of 12 models the `(EV, entropy)` proxy loses essentially
   nothing** — the spread index is a faithful summary.
2. **`FalconMamba`'s apparent "diffuseness" is a smooth distribution with a digit-6
   notch, not bimodality.** Most of its flagged second modes are `digit-notch`: the model
   under-uses the token "6", leaving a dip between 5 and 7. This is the same tokenizer
   phenomenon as the category-2 suppression in the Llama-2 soft-evidence work — an
   *artifact of the rating channel*, not a two-mode judgment. **This corrects the earlier
   reading** (§4) that grouped FalconMamba as a diffuse *responder*: it is diffuse but
   *unimodal-with-a-notch*.
3. **A digit-2/6 suppression artifact is real but minor** across Aya, FalconMamba and the
   Qwen family (~1–12%) — worth knowing before treating any single distribution's shape as
   substantive.

```{r medoids-examples, fig.height=3.4}
# an Aya genuine mixture vs a matched-EV unimodal judgment: same EV, opposite shape
aya_bi <- md[model == "Aya" & shape == "semantic"][order(-(p1 + p7))][1]
uni <- md[model == "Gemma12" & shape == "unimodal"][which.min(abs(ev - aya_bi$ev))]
ex <- rbind(
  data.table(who = sprintf("Aya: %s→%s  (EV=%.1f, bimodal)", aya_bi$adj_i, aya_bi$adj_j, aya_bi$ev),
             cat = 1:7, p = as.numeric(aya_bi[, ..pcols])),
  data.table(who = sprintf("Gemma3-12B: %s→%s  (EV=%.1f, unimodal)", uni$adj_i, uni$adj_j, uni$ev),
             cat = 1:7, p = as.numeric(uni[, ..pcols])))
ggplot(ex, aes(factor(cat), p, fill = who)) +
  geom_col(show.legend = FALSE) + facet_wrap(~who) +
  scale_fill_manual(values = c("#8172B3", "#DD8452")) +
  labs(title = "Same expected rating, opposite response process",
       subtitle = "the spread index and EV+entropy cannot tell these apart; only the raw distribution can",
       x = "rating category", y = "probability") + theme_judge()
```

### 6.4 Verdict

The sample **validates the response-process pipeline** and sharpens it: the `(EV, entropy)`
spread proxy is adequate for 10/12 models, with **`Aya` the one model whose genuine
mixture judgments (~11%) require the raw distributions**, and a minor **digit-2/6
suppression** artifact to keep in mind cohort-wide. It also corrects the FalconMamba
characterization: its breadth is a notched-unimodal, not a bimodal, response. For the full
523², rgb notes any adjective subset is a cheap re-run — so the raw distributions are
obtainable without an independent re-collection if the Aya mixtures (or the digit artifact)
warrant it downstream.

## Takeaways for the measurement program

1. **Scale use is a rater property, not a constant.** Leniency, differentiation and
   decisiveness each vary ~4× across the cohort and are only loosely coupled — three
   separate response-style nuisances to model, not one calibration knob.
2. **Two raters (`Llama3.2-3B`, `FalconMamba-7B`) are range-restricted / low-information.**
   They carry little differentiated signal — but Report III shows they still *agree* with
   the consensus and do not harm it, so exclusion is **not** warranted; the ensemble is
   robust. (This corrects the tempting "down-weight the flat raters" intuition.)
3. **The entropy channel is real signal.** The inverted-U response process is sensible
   for discriminating raters; where it flattens, the rater is guessing. Entropy gives us
   a per-judgment reliability weight we will use.
4. **Big additive marginals (prevalence + anchor generosity)** must be removed before
   `B` is read as a similarity — motivating the symmetrization + double-centering in
   Reports II–IV.
5. **The two-moment proxy is validated against the raw distributions** (§6, 35-medoid
   sample): faithful for 10/12 models. The exceptions are specific and now known — `Aya`
   makes genuine no-OR-yes *mixtures* (~11%, invisible to EV+entropy), and a minor
   digit-2/6 *suppression* artifact runs through Aya/FalconMamba/Qwen. FalconMamba's
   "diffuseness" is a notched-unimodal, not bimodality.
```
