---
title: "JUDGE channel — I. Descriptives & response processes"
subtitle: "Implicit personality theory in 12 LLMs, read as a rating instrument"
author: "ecb"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 9
    fig_height: 5.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 110)
source("_common.R")
```

## What this data is, and the stance I take on it

The **JUDGE** channel (rgb's `tom_likely`) asks each model, for every ordered pair of
523 person-descriptive adjectives *(i, j)*, *"how strongly would a person described as
**i** also be **j**?"* and reads the answer as an **expected rating on a 1–7 Likert
scale** from the full log-prob distribution over the rating tokens (not the argmax).
Each model therefore produces a `523 × 523` matrix **B** of conditional-endorsement
ratings (row-conditioned, *not* symmetric, diagonal not collected), plus a matching
matrix **Hent** of the Shannon entropy of each rating distribution — a per-judgment
uncertainty.

rgb's own analyses of this data (W15–W16) are **geometric and mechanistic**: the
read/write dissociation, a Helmholtz–Hodge decomposition of the asymmetry, factor
congruence to the human 525-PDA. This report series takes the **psychometric** view
the geometry brackets: *if we treat these judgments as measurements, are they sound?*
The unit I adopt (per ecb's steer) is **models as raters** — 12 judges rating the same
273,006 ordered pairs. Reliability, validity and response-process questions all follow
from that frame.

This first report is the descriptive floor: **how does each rater use the scale?**
Everything downstream (what to do about asymmetry; whether the raters agree; whether the
structure is interpretable) is conditioned by the response processes documented here.

> **Data caveat, stated once.** We hold the *expected value* `B` and the *entropy*
> `Hent` of each rating distribution — **not** the raw 7-category histogram. So
> "category usage" below is the distribution of **rounded expected values** (an
> *effective* category), not the true modal-response frequency. EV+entropy still pin
> the response process tightly (entropy≈0 ⇒ all mass on one category ⇒ EV≈that integer),
> but claims about specific categories are claims about the effective rating.

```{r load}
summ <- read_csv_("model_summary.csv") %>% as_tibble() %>%
  mutate(display = factor(display, levels = disp_ord))
marg <- read_csv_("marginals.csv") %>% as_tibble()
dir_ev <- fread_gz("dir_ev.csv.gz")     # i,j, <12 models>  (273006 directional EVs)
dir_h  <- fread_gz("dir_hent.csv.gz")
# long form for distributional plots
ev_long <- melt(dir_ev, id.vars = c("i","j"), variable.name = "model", value.name = "ev")
ev_long[, display := factor(DISPLAY[as.character(model)], levels = disp_ord)]
ev_long[, family  := FAMILY[as.character(model)]]
h_long <- melt(dir_h, id.vars = c("i","j"), variable.name = "model", value.name = "hent")
```

## 1. Scale usage: some raters never leave the middle

The single most consequential descriptive fact is that the 12 models **do not use the
1–7 scale in remotely the same way.** The table sorts by mean rating (leniency); note
especially the spread column (`ev_sd`) and the extreme-category fractions.

```{r usage-table}
tab <- summ %>% arrange(ev_mean) %>%
  transmute(Model = display, Family = family, `Params (B)` = params_b,
            `Mean` = round(ev_mean,2), `SD` = round(ev_sd,2),
            `Skew` = round(ev_skew,2), `Excess kurt.` = round(ev_kurtosis,2),
            `% ≤2` = round(100*frac_le2,1), `% 3–5` = round(100*frac_mid_3to5,1),
            `% ≥6` = round(100*frac_ge6,1))
knitr::kable(tab, caption = "Per-rater location, spread and tail usage of the 1–7 scale.")
```

```{r cat-usage, fig.height=5}
catcols <- paste0("cat", 1:7)
cat_long <- summ %>% select(display, all_of(catcols)) %>%
  pivot_longer(all_of(catcols), names_to = "cat", values_to = "frac") %>%
  mutate(cat = factor(as.integer(gsub("cat","",cat))))
ggplot(cat_long, aes(cat, frac, fill = cat)) +
  geom_col() + facet_wrap(~display, ncol = 4) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1, guide = "none") +
  labs(title = "Effective category usage (fraction of pairs by rounded expected rating)",
       x = "rounded expected rating (1–7)", y = "fraction of pairs") +
  theme_judge()
```

Read this as a **response-style census**:

- **Range-restricted / central responders.** `Llama3.2-3B` puts ~96% of all 273k
  judgments in categories {4,5} and **never** effectively uses 1, 2, 6 or 7
  (SD = 0.47). `FalconMamba-7B` lives in {3,4,5} (SD = 0.75). As raters they are close
  to *non-discriminating*: whatever the pair, the answer is "middling".
- **Full-range / extreme responders.** `Aya-expanse-8B` and `Phi4-mini` spread across
  all seven categories and put ~24% of judgments at the ceiling (7) — SD ≈ 1.7–1.9.
- **Everything in between**, with the Qwen family modal around 3–4 and the Gemma family
  running higher and more bimodal.

This is exactly the heterogeneity that makes cross-model *magnitude* comparison unsafe
(rgb's calibration caveat) — but stated in response-style terms it is sharper: **these
are different rating styles, not just different means.** Range restriction limits how
much *differentiated information* a rater can carry (it caps the achievable
between-pair variance) — though, importantly, Report III shows it does **not** make these
raters disagree with the crowd: a flat rater is low-information, not noisy.

## 2. Location and spread are only loosely coupled

```{r loc-spread}
ggplot(summ, aes(ev_mean, ev_sd, colour = family, label = display)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20, seed = 1) +
  scale_colour_manual(values = FAM_PAL, name = NULL) +
  labs(title = "Rater leniency (mean) vs differentiation (SD)",
       x = "mean rating (leniency →)", y = "SD of ratings (differentiation →)") +
  theme_judge()
```

Leniency (mean 3.63 → 4.83) and differentiation (SD 0.47 → 1.95) vary independently:
`Llama3.2-3B` is *lenient but flat* (high mean, tiny SD — it agrees warmly with almost
everything), whereas `Aya` is *lenient and spread*. A high mean does not buy you
usable variance. For the measurement, **differentiation (SD) is the resource** — a rater
with SD 0.47 can encode at most a sliver of true between-pair signal.

## 3. Decisiveness: the entropy channel is a second, partly independent axis

Each judgment carries an entropy. Averaged per rater, this is a **decisiveness** index
that is *not* the same thing as scale spread.

```{r entropy-violin, fig.height=5}
ggplot(h_long, aes(x = factor(DISPLAY[as.character(model)], levels = disp_ord), y = hent,
                   fill = FAMILY[as.character(model)])) +
  geom_violin(scale = "width", linewidth = 0.2) +
  scale_fill_manual(values = FAM_PAL, name = NULL) +
  coord_flip() +
  labs(title = "Per-judgment entropy (uncertainty) by rater",
       x = NULL, y = "Shannon entropy of the 1–7 rating distribution (nats)") +
  theme_judge()
```

```{r ev-entropy-curve, fig.height=4.5}
# entropy as a function of the rating level: the response-process signature
set.seed(1)
samp <- ev_long[sample(.N, 60000)][, .(i,j,model,ev)]
samp <- merge(samp, h_long[, .(i,j,model,hent)], by = c("i","j","model"))
samp[, display := factor(DISPLAY[as.character(model)], levels = disp_ord)]
ggplot(samp, aes(ev, hent)) +
  geom_hex(bins = 40) + scale_fill_viridis_c(trans = "log10", name = "pairs") +
  facet_wrap(~display, ncol = 4, scales = "free_y") +
  labs(title = "Entropy vs expected rating — the graded-in-the-middle signature",
       x = "expected rating", y = "entropy") +
  theme_judge() + theme(legend.position = "right")
```

Two things:

1. **The inverted-U** (entropy peaks at mid-scale, collapses at the extremes) is present
   for the discriminating raters — a *sensible* response process: the model is certain
   when it judges a pair clearly (dis)similar and hedges in between.
2. **The non-responsive raters break it.** `FalconMamba` sits at entropy ≈ 1.66 nats
   almost everywhere (near the `log(7)=1.95` uniform ceiling) — it is *maximally
   uncertain on every pair*. `Llama3.2-3B` is similar (never decisive: only 0.002% of
   judgments have entropy < 0.1). Their narrow EV in §1 is now explained: they are not
   compressing a confident signal, they are **reporting near-flat distributions whose
   means happen to sit mid-scale.**

## 4. A response-process typology: two kinds of "uncertain"

Crossing scale differentiation (SD of EV) against decisiveness (mean entropy) separates
raters that look similar on either axis alone.

```{r typology}
ggplot(summ, aes(hent_mean, ev_sd, colour = family, label = display)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20, seed = 2) +
  scale_colour_manual(values = FAM_PAL, name = NULL) +
  geom_vline(xintercept = median(summ$hent_mean), linetype = 3, colour = "grey60") +
  geom_hline(yintercept = median(summ$ev_sd), linetype = 3, colour = "grey60") +
  labs(title = "Response-process typology",
       subtitle = "up-left = confident differentiator · down-right = diffuse non-discriminator",
       x = "mean per-judgment entropy (← decisive | diffuse →)",
       y = "SD of expected ratings (↑ differentiated)") +
  theme_judge()
```

- **Confident differentiators** (low entropy, high SD): the Gemma family, esp.
  `Gemma-4-31B`/`Gemma3-27B`. They commit to graded, spread judgments.
- **Diffuse non-discriminators** (high entropy, low SD): `FalconMamba-7B`,
  `Llama3.2-3B`. Little between-pair signal *and* little within-judgment confidence.
- **Spread-but-hedging** (high entropy, high SD): `Aya`, `Phi4` — they use the whole
  scale but keep meaningful mass off the modal category.

This typology is the load-bearing setup for reliability — but with a twist that
Report III makes precise and that is easy to get wrong: a **diffuse non-discriminator
carries little *differentiated* information** (its between-pair variance is small), yet
it still **agrees with the consensus** and does not inject noise (its judgments are a
smooth, prevalence-tracking, compressed image of the crowd's). So the flat raters are
*low-information*, not *low-quality*; the raters that actually pull away from the
consensus turn out to be the high-variance *idiosyncratic* models, not the flat ones.
Weighting/trimming therefore buys much less than the typology first suggests (Report III
§3, Report V).

## 5. Marginal structure: base rates (prevalence) and anchor generosity

Because the diagonal (self-pairs) was not collected, each rater's matrix has two
informative margins:

- **column mean** = how strongly adjective *j* is ascribed *given any anchor* — a
  **base rate / prevalence** ("Awake", "Normal" are ascribed to almost anyone).
- **row mean** = how generously anchor *i* ascribes *other* traits — anchor-level
  acquiescence.

```{r marginals, fig.height=4.5}
mm <- marg %>% mutate(display = factor(DISPLAY[model], levels = disp_ord))
ggplot(mm, aes(col_mean, fill = FAMILY[model])) +
  geom_histogram(bins = 40) + facet_wrap(~display, ncol = 4, scales = "free_y") +
  scale_fill_manual(values = FAM_PAL, guide = "none") +
  labs(title = "Target base-rate (column-mean) distribution per rater",
       x = "mean ascription of a target across all anchors (prevalence)", y = "count") +
  theme_judge()
```

```{r prevalence-extremes}
cohort_prev <- mm %>% group_by(adjective) %>% summarise(prev = mean(col_mean)) %>%
  arrange(desc(prev))
knitr::kable(
  bind_cols(
    cohort_prev %>% slice_head(n = 12) %>% transmute(`Most universally ascribed` = adjective, `μ` = round(prev,2)),
    cohort_prev %>% slice_tail(n = 12) %>% arrange(prev) %>% transmute(`Least ascribed` = adjective, ` μ` = round(prev,2))
  ), caption = "Cohort-mean target base rates (prevalence marginals).")
```

The prevalence marginal is large and structured — and it is precisely what a
marginal-free human *correlation* matrix does not contain, which is why the reliability
and validity reports **double-center** (strip additive row/column marginals) before
treating `B` as a similarity. It is also the primary axis of the asymmetry (Report II).

## Takeaways for the measurement program

1. **Scale use is a rater property, not a constant.** Leniency, differentiation and
   decisiveness each vary ~4× across the cohort and are only loosely coupled — three
   separate response-style nuisances to model, not one calibration knob.
2. **Two raters (`Llama3.2-3B`, `FalconMamba-7B`) are range-restricted / low-information.**
   They carry little differentiated signal — but Report III shows they still *agree* with
   the consensus and do not harm it, so exclusion is **not** warranted; the ensemble is
   robust. (This corrects the tempting "down-weight the flat raters" intuition.)
3. **The entropy channel is real signal.** The inverted-U response process is sensible
   for discriminating raters; where it flattens, the rater is guessing. Entropy gives us
   a per-judgment reliability weight we will use.
4. **Big additive marginals (prevalence + anchor generosity)** must be removed before
   `B` is read as a similarity — motivating the symmetrization + double-centering in
   Reports II–IV.
```
