---
title: "JUDGE channel — II. Asymmetry as a measurement problem"
subtitle: "Is B[i,j] ≠ B[j,i] signal or nuisance, and what do we do about it?"
author: "ecb"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 9
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 110)
source("_common.R")
```

## The problem

Almost every psychometric use of `B` — as an inter-adjective *similarity*, a proximity
for factoring or MDS, a covariance analogue — assumes it is **symmetric**. But the raw
judgment is conditional and directed: `B[i,j]` ("given **i**, how strongly also **j**")
need not equal `B[j,i]`. Before we symmetrize (as Reports III–IV will), we owe the data
three questions:

1. **How much** asymmetry is there?
2. **What is it** — a removable marginal artefact, or substantive directional structure?
3. **Is it reliable** across raters — reproducible signal, or per-model noise?

The answers decide whether symmetrizing is a free simplification or throws away a
construct. (rgb's W16 §5 gave this a full Helmholtz–Hodge treatment aimed at *what the
asymmetry means*; here I ask the narrower measurement question *what to do with it*.)

```{r load}
summ <- read_csv_("model_summary.csv") %>% as_tibble() %>%
  mutate(display = factor(display, levels = disp_ord))
marg <- read_csv_("marginals.csv") %>% as_tibble()
asy  <- fread_gz("asym_ev.csv.gz")     # i<j : B[i,j]-B[j,i]
sym  <- fread_gz("sym_ev.csv.gz")      # i<j : (B[i,j]+B[j,i])/2
n <- nrow(adjectives)
```

## 1. Magnitude: small against the grand mean, but ~1/5 of the *meaningful* signal

Decomposing each rater's off-diagonal matrix into symmetric + antisymmetric parts, the
share of squared magnitude in the antisymmetric part depends entirely on the reference —
and this is a trap worth defusing up front. Measured against the **mean-inclusive**
sum-of-squares, asymmetry looks negligible; measured against the **centered** variance
(the part that actually carries similarity information, and all that survives the
double-centering every downstream use applies), it is substantial.

```{r magnitude}
dir <- fread_gz("dir_ev.csv.gz")
frac <- sapply(MODELS, function(m){
  B <- build_matrix(dir, m, "dir"); off <- !is.na(B); A <- (B - t(B))/2
  mu <- mean(B[off])
  c(meaninc = sum(A[off]^2)/sum(B[off]^2),
    centered = sum(A[off]^2)/sum((B[off]-mu)^2))
})
mag <- tibble(Model = DISPLAY[MODELS], Family = FAMILY[MODELS],
              `Asym % of mean-incl. SS` = round(100*frac["meaninc",],2),
              `Asym % of *centered* variance` = round(100*frac["centered",],1),
              `SD of ratings` = round(summ$ev_sd[match(MODELS,summ$model)],2)) %>%
  arrange(`Asym % of *centered* variance`)
knitr::kable(mag, caption = "Antisymmetric share against two references. The centered figure is the honest one.")
```

Against the grand mean (~3.6 Likert), asymmetry is 0.2–3.5% — but that denominator is
dominated by the mean, so the number is almost meaningless. Against the **centered**
signal it is **18–26% (median ~20%)**: roughly a fifth of each rater's *differentiated*
judgment variance is directional. **Symmetrizing is therefore not a free rounding-off** —
it sets aside a real slice of signal. Note too that the flat rater `Llama3.2-3B` has the
*smallest* mean-inclusive asymmetry (0.25%) yet the *largest* centered one (24.7%): with
almost no differentiated variance to begin with, most of what little it has is
direction-inconsistent — a first sign that for the flat raters the asymmetry is mostly
noise (§3). The real questions are now sharper: **what is that ~20%, and does it
replicate?**

## 2. Anatomy: half of the asymmetry is a prevalence gradient

An antisymmetric matrix on a complete graph decomposes (discrete Helmholtz–Hodge) into a
**gradient** part — a single transitive potential `g`, `D_grad[i,j] = g[i] − g[j]`,
curl-free and carrying all the "flow" — plus a divergence-free **curl** (genuine
cycles). The least-squares potential is `g[i] = mean_j D[i,j] = rowmean_i(B) −
colmean_i(B)` = anchor generosity minus target prevalence.

```{r hodge}
hodge <- map_dfr(MODELS, function(m) {
  D <- build_matrix(asy, m, "asym")          # full antisymmetric, NA diag
  D[is.na(D)] <- 0
  g <- rowMeans(D)
  Dg <- outer(g, g, "-")
  data.frame(model = m,
             grad_frac = sum(Dg^2) / sum(D^2),
             # correlate the potential with the prevalence marginal
             cor_g_prev = cor(g, -marg$col_mean[marg$model == m]),
             cor_g_gen  = cor(g,  marg$row_mean[marg$model == m]))
}) %>% mutate(display = factor(DISPLAY[model], levels = disp_ord))

knitr::kable(
  hodge %>% arrange(desc(grad_frac)) %>%
    transmute(Model = display,
              `Gradient % of asym.` = round(100*grad_frac,1),
              `corr(potential, −prevalence)` = round(cor_g_prev,2),
              `corr(potential, generosity)`  = round(cor_g_gen,2)),
  caption = "Gradient (transitive potential) share of the antisymmetric part, and what the potential is."
)
cat(sprintf("Mean gradient fraction: %.2f\n", mean(hodge$grad_frac)))
```

About **half** (mean 0.51) of the already-small asymmetry is a pure gradient, and that
gradient is essentially the **base-rate / prevalence** structure from Report I §5: the
flow runs toward universally-ascribed traits (everyone "also is" Awake/Normal; few
"also are" some rare trait). This half is the *trivial Bayes direction* — a marginal we
strip anyway by double-centering. The other half is curl: genuine directional
person-perception cycles (rgb's warmth×morality "charming villain", Dunning–Kruger,
etc.). So:

> Asymmetry = (≈50% removable prevalence marginal) + (≈50% substantive but small curl).

## 3. Is the asymmetry reliable across raters?

The decisive measurement question. If the antisymmetric part does not replicate across
models it is rater noise and symmetrizing is pure gain. Mean pairwise Pearson
correlation across the 9 raters, symmetric vs antisymmetric parts:

```{r reliab-asym}
mean_pairwise <- function(dt, models) {
  X <- as.matrix(dt[, ..models]); C <- cor(X)
  mean(C[upper.tri(C)])
}
disc <- setdiff(MODELS, "Llama")            # drop the one flat rater (FalconMamba pending)
tab <- tibble(
  Part = c("Symmetric  (i,j)+(j,i)", "Antisymmetric (i,j)−(j,i)"),
  `r̄ across all 9` = c(mean_pairwise(sym, MODELS), mean_pairwise(asy, MODELS)),
  `r̄ across 8 discriminating` = c(mean_pairwise(sym, disc), mean_pairwise(asy, disc))
) %>% mutate(across(where(is.numeric), ~round(.,3)))
knitr::kable(tab, caption = "Inter-rater agreement: the symmetric channel is far more reproducible.")
```

The symmetric similarity is **strongly shared** across raters (r̄ = 0.77); the
antisymmetric part is **reproducible but much weaker** (r̄ = 0.43). It is *not* noise —
0.43 is well above zero and survives dropping the flat rater — but it carries
roughly half the cross-rater signal of the symmetric channel. Splitting the asymmetry
into its gradient and curl:

```{r reliab-split}
# build per-rater gradient-potential vectors and curl (i<j) vectors, correlate across raters
G <- sapply(MODELS, function(m){ D<-build_matrix(asy,m,"asym"); D[is.na(D)]<-0; rowMeans(D) })
# curl = asym minus gradient reconstruction, evaluated on i<j pairs
ij <- as.matrix(asy[, .(i,j)]) + 1L
curlmat <- sapply(MODELS, function(m){
  D<-build_matrix(asy,m,"asym"); D[is.na(D)]<-0; g<-rowMeans(D)
  (D - outer(g,g,"-"))[ij]
})
r_prev <- { C<-cor(G[, disc]); mean(C[upper.tri(C)]) }
r_curl <- { C<-cor(curlmat[, disc]); mean(C[upper.tri(C)]) }
knitr::kable(tibble(
  Component = c("Prevalence gradient (potential g)", "Curl (directional cycles)"),
  `r̄ across 8 discriminating raters` = round(c(r_prev, r_curl), 3)),
  caption = "Where the directional reliability lives.")
```

As expected, the reliability of the asymmetry is carried by its **prevalence gradient**
(highly reproducible — base rates are a stable property of the cohort), while the
**curl** is only weakly shared. So the reproducible part of the asymmetry is precisely
the part we *remove* as a marginal; the substantive directional theory (curl) is real
but faint and idiosyncratic.

## 4. What the directional signal looks like (cohort consensus)

The most directionally-asymmetric pairs, averaged over the 9 raters — read as
"**anchor ⇒ target** is endorsed much more than **target ⇒ anchor**":

```{r top-asym}
asy_long <- melt(asy, id.vars = c("i","j"), variable.name="model", value.name="d")
cons <- asy_long[, .(d = mean(d)), by = .(i,j)]
cons[, `:=`(a_i = ADJ[i+1L], a_j = ADJ[j+1L])]
top <- cons[order(-abs(d))][1:15]
top[, `:=`(strong = ifelse(d>0, paste(a_i,"⇒",a_j), paste(a_j,"⇒",a_i)),
           weak   = ifelse(d>0, paste(a_j,"⇒",a_i), paste(a_i,"⇒",a_j)),
           gap = round(abs(d),2))]
knitr::kable(top[, .(`Strongly endorsed direction`=strong,
                     `Weak reverse`=weak, `Δ (Likert pts)`=gap)],
             caption = "Largest cohort-consensus directional gaps.")
```

These are interpretable one-way implications — a specific trait implies a general/
prevalent one but not vice versa — consistent with the prevalence account (§2) more than
a valence one. (rgb's curl-plane analysis extracts the subtler person-perception cycles
from the residual; that is a *substantive* follow-up, not a measurement prerequisite.)

## Measurement verdict

**Symmetrize** — but *not* because the asymmetry is negligible (it is ~20% of the
centered signal). For the core JUDGE *similarity* (Reports III–IV) the symmetric part
`(B[i,j]+B[j,i])/2` is the right object on three sounder grounds:

1. **Reliability** — the antisymmetric part is ~2× less reproducible across raters
   (r̄ 0.43 vs 0.77), i.e. lower-quality signal per unit variance.
2. **Half is a removable marginal** — ~50% of it is a prevalence gradient (§2) that
   double-centering strips anyway, and that gradient is where most of its reliability
   sits; the reproducible remainder (curl) is small.
3. **It is a different construct** — directionality (A⇒B ≠ B⇒A) is not a *similarity*;
   it does not belong in a symmetric proximity regardless of its size.

**But do not call the asymmetry "error."** It is ~20% of the centered signal — not
negligible — and its curl is a weakly-reproducible but non-zero directional construct:
the model's *implicit-personality-theory* directionality (A implies B more than B implies
A). It is a legitimate secondary channel for a directional analysis (and the natural
bridge to rgb's person-perception cycles), just not part of the symmetric *similarity* we
take forward. Downstream reports therefore use **symmetrized, double-centered** `B` for
the similarity measurement and treat directionality as a separate, flagged channel worth
its own study — not a nuisance to be rounded away.
```
