---
title: "JUDGE channel — III. Reliability & generalizability"
subtitle: "12 LLMs as raters of adjective-pair judgments"
author: "ecb"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 9
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 110)
source("_common.R")
```

## Frame

Per ecb's steer, the unit is **models as raters**: 12 judges each rate the same 136,503
unordered adjective pairs (symmetric similarity `S = (B[i,j]+B[j,i])/2`, justified in
Report II). Reliability here is **inter-rater generalizability** — if we call the JUDGE
similarity a *measurement*, how much of it is shared true score vs rater idiosyncrasy,
and how many raters does a dependable measurement need? This is the psychometric spine of
rgb's "wisdom-of-crowds" observation (W16 §8b), made into variance components and a
D-study.

Two measurement objects are analysed:

- **raw pair ratings** `S` — agreement on the endorsement level of each pair (includes
  the shared prevalence marginal from Report I);
- **double-centered** `S̃` — the marginal-free *similarity structure* (prevalence and
  anchor generosity swept out), a stricter "do raters agree on the personality
  covariance shape" test.

```{r load}
sym <- fread_gz("sym_ev.csv.gz")
X <- as.matrix(sym[, ..MODELS])                 # pairs x raters (raw symmetric)
summ <- read_csv_("model_summary.csv") %>% as_tibble()
np <- nrow(X); nr <- ncol(X)
```

## 1. Variance components (G-theory, crossed pair × rater)

The design is fully crossed and balanced (every rater rates every pair once), so the
random-effects variance components come straight from the ANOVA mean squares.

```{r vc}
gtheory <- function(X) {
  np<-nrow(X); nr<-ncol(X); grand<-mean(X)
  SSp<-nr*sum((rowMeans(X)-grand)^2); SSr<-np*sum((colMeans(X)-grand)^2)
  SSe<-sum((X-grand)^2)-SSp-SSr
  MSp<-SSp/(np-1); MSr<-SSr/(nr-1); MSe<-SSe/((np-1)*(nr-1))
  list(vp=(MSp-MSe)/nr, vr=(MSr-MSe)/np, ve=MSe)
}
vc <- gtheory(X)
tibble(Component=c("Pair (universe / true score)","Rater (leniency main effect)",
                   "Pair×Rater + error (residual)"),
       Variance=round(c(vc$vp,vc$vr,vc$ve),3),
       `% of total`=round(100*c(vc$vp,vc$vr,vc$ve)/sum(unlist(vc)),1)) %>%
  knitr::kable(caption="Generalizability variance components for the raw symmetric similarity.")
```

The **pair** facet dominates (≈61% of variance) — there is a large shared true-score
signal every rater is tracking. The **rater/leniency** main effect (≈11%) is the
response-style heterogeneity of Report I, now as a variance term; it depresses *absolute*
agreement but is removed by design in a *relative* (consistency) coefficient. The
residual (≈29%) is pair×rater interaction + measurement error — genuine disagreement
about *which* pairs are similar.

```{r icc}
icc_from_vc <- function(vc, k) c(
  `ICC(2,1) abs, 1 rater` = vc$vp/(vc$vp+vc$vr+vc$ve),
  `ICC(2,k) abs, k raters`= vc$vp/(vc$vp+(vc$vr+vc$ve)/k),
  `ICC(3,1) consist, 1`   = vc$vp/(vc$vp+vc$ve),
  `ICC(3,k) consist, k`   = vc$vp/(vc$vp+vc$ve/k))
knitr::kable(round(t(t(icc_from_vc(vc, nr))),3), col.names="coefficient",
             caption=sprintf("Intraclass correlations (k = %d raters).", nr))
```

- A **single LLM rater** has only *moderate* absolute agreement with the field
  (ICC(2,1) = 0.61). One model's implicit personality theory is a noisy instrument.
- The **12-rater ensemble** is *excellent* (ICC(2,k) = 0.95) — the reliable measurement
  object is the **cohort consensus**, not any single model. This is precisely why rgb's
  aggregate matches the human matrix above every individual model.
- Consistency exceeds agreement (0.68 > 0.61 single; 0.96 > 0.95 average): removing the
  leniency main effect recovers reliability, quantifying the calibration confound rgb
  flagged — **~0.07 of single-rater reliability is lost purely to differing scale use.**

## 2. D-study: how many raters do you need?

```{r dstudy}
ks <- 1:20
d <- tibble(k=ks,
            Relative =vc$vp/(vc$vp+vc$ve/ks),
            Absolute =vc$vp/(vc$vp+(vc$vr+vc$ve)/ks)) %>%
  pivot_longer(-k, names_to="coefficient", values_to="G")
ggplot(d, aes(k, G, colour=coefficient)) +
  geom_line(linewidth=1) + geom_point(size=1.4) +
  geom_hline(yintercept=c(.8,.9), linetype=3, colour="grey55") +
  geom_vline(xintercept=nr, linetype=2, colour="grey55") +
  annotate("text", x=nr+.3, y=.55, label="cohort n=12", hjust=0, size=3, colour="grey40") +
  scale_x_continuous(breaks=c(1,5,10,12,15,20)) +
  scale_colour_manual(values=c(Relative="#4C72B0", Absolute="#DD8452"), name=NULL) +
  labs(title="Generalizability vs number of raters (Spearman–Brown / D-study)",
       x="number of LLM raters averaged", y="G coefficient") +
  theme_judge()
crossings <- sapply(c(.8,.9), function(t) min(which(vc$vp/(vc$vp+(vc$vr+vc$ve)/ks) >= t)))
cat(sprintf("Raters needed for absolute G≥0.8: %d; ≥0.9: %d.\n", crossings[1], crossings[2]))
```

Dependability climbs steeply: ~4 raters for absolute G ≥ 0.8, ~8 for ≥ 0.9. The
cohort of 12 is comfortably in the dependable range — but the curve also says a
*handful* of well-chosen models would nearly suffice, so the ensemble's value is real
but saturating.

## 3. Rater quality is multidimensional — and the flat raters are *not* the bad ones

The tempting move is to score each rater by its correlation with the leave-one-out mean
of the others and call the low scorers "bad." Doing so on the **raw** ratings is
misleading, because raw agreement is inflated by the shared prevalence marginal (every
rater, even a flat one, reproduces the base-rate structure). So I score raters on three
distinct axes: **information** (SD of ratings — differentiation), **between-rater
agreement on structure** (leave-one-out correlation on the *double-centered*,
prevalence-free similarity), and **within-rater consistency** (agreement of a rater's two
conditional directions, §5).

```{r rater-quality}
loo_raw <- sapply(seq_len(nr), function(r) cor(X[,r], rowMeans(X[,-r])))
Sdc <- sapply(MODELS, function(m) upper_vec(double_center(build_matrix(sym, m, "sym"))))
loo_dc <- sapply(seq_len(nr), function(r) cor(Sdc[,r], rowMeans(Sdc[,-r])))
rq <- tibble(model=MODELS, display=DISPLAY[MODELS], family=FAMILY[MODELS],
             `raw`=loo_raw, `structural`=loo_dc,
             ev_sd=summ$ev_sd[match(MODELS,summ$model)])
rql <- rq %>% pivot_longer(c(raw, structural), names_to="metric", values_to="loo") %>%
  mutate(display=factor(display, levels=rev(rq$display[order(rq$structural)])))
ggplot(rql, aes(loo, display, colour=family, shape=metric)) +
  geom_line(aes(group=display), colour="grey80") +
  geom_point(size=2.6) + scale_colour_manual(values=FAM_PAL, name=NULL) +
  scale_shape_manual(values=c(raw=1, structural=16), name=NULL) +
  labs(title="Agreement of each rater with the leave-one-out consensus",
       subtitle="open = raw ratings (prevalence-inflated) · filled = double-centered structure",
       x="correlation with mean of the other 11 raters", y=NULL) +
  xlim(0.7,1) + theme_judge()
```

The ranking **inverts the naive expectation**. On the structural (marginal-free) metric
the lowest agreers are the high-variance *idiosyncratic* models — **`Gemma-4-31B` (0.76)
and `Aya` (0.83)** — while the flat non-discriminators **`FalconMamba` (0.92) and
`Llama3.2-3B` (0.89) agree with the consensus *above average*.** A flat rater is a
*compressed but faithful* image of the crowd (it tracks the shared structure, just with
small amplitude); it is **low-information, not low-agreement, and not noisy.** So the
"down-weight the non-responsive raters" intuition from Report I's typology is, on the
evidence, **wrong** — a correction worth stating plainly.

```{r trim-effect}
disc  <- setdiff(MODELS, c("Llama","FalconMamba"))     # flat / low-information
idio  <- setdiff(MODELS, c("Gemma4","Aya"))            # idiosyncratic / low-agreement
vcF <- gtheory(as.matrix(sym[, ..disc])); vcI <- gtheory(as.matrix(sym[, ..idio]))
tibble(Ensemble=c("All 12 raters","drop 2 flat (Llama3.2, FalconMamba)",
                  "drop 2 idiosyncratic (Gemma-4, Aya)"),
       `ICC(2,1) single`=round(c(icc_from_vc(vc,12)[1], icc_from_vc(vcF,10)[1], icc_from_vc(vcI,10)[1]),3),
       `pair var (true)`=round(c(vc$vp, vcF$vp, vcI$vp),3),
       `residual`       =round(c(vc$ve, vcF$ve, vcI$ve),3)) %>%
  knitr::kable(caption="Both trims raise single-rater ICC — by opposite mechanisms.")
```

Both trims *modestly* raise single-rater reliability, but for **opposite reasons**, which
is the real lesson: dropping the **flat** raters raises reliability by **increasing pair
(true-score) variance** (0.91→1.10 — the remaining raters differentiate more), whereas
dropping the **idiosyncratic** raters raises it by **cutting the residual** (0.43→0.36 —
less pair×rater disagreement). Neither is "removing noise"; they trade off different
variance components. The 12-rater *ensemble* reliability is essentially untouched either
way — the consensus is robust. Report V follows this up with a full consensus-construction
study (competence- and confidence-weighting), whose punchline is previewed here: it buys
almost nothing.

## 4. Which pairs are reliably vs idiosyncratically judged?

Centering each rater (removing leniency), the **between-rater SD of each pair** is its
contestedness. Relating it to the consensus similarity level and to judgment entropy:

```{r pair-reliab}
Xc <- sweep(X, 2, colMeans(X))                 # remove rater leniency
pair_sd <- sqrt(rowMeans((Xc - rowMeans(Xc))^2) * nr/(nr-1))
cons <- rowMeans(X)
syh <- fread_gz("sym_hent.csv.gz")
pair_ent <- rowMeans(as.matrix(syh[, ..MODELS]))
pr <- data.table(i=sym$i, j=sym$j, cons=cons, disagree=pair_sd, ent=pair_ent)
set.seed(1)
ggplot(pr[sample(.N, 40000)], aes(cons, disagree)) +
  geom_hex(bins=45) + scale_fill_viridis_c(trans="log10", name="pairs") +
  labs(title="Rater disagreement vs consensus similarity",
       subtitle="agreement is highest at the extremes; mid-similarity pairs are contested",
       x="consensus similarity (mean over raters)", y="between-rater SD (contestedness)") +
  theme_judge() + theme(legend.position="right")
cat(sprintf("corr(disagreement, |consensus−midpoint|) = %.2f;  corr(disagreement, entropy) = %.2f\n",
    cor(pr$disagree, abs(pr$cons-4)), cor(pr$disagree, pr$ent)))
```

Disagreement is an **inverted-U in similarity**: raters concur that clear synonyms are
similar and clear antonyms are not, and **contest the middle** — exactly where a pair's
co-occurrence is genuinely theory-laden. Contestedness also tracks per-judgment entropy:
the pairs raters *disagree* on are the pairs each rater is *individually unsure* about.
That double signature (between-rater SD ↔ within-rater entropy) is reassuring — the
uncertainty is in the construct, not a rater artefact.

```{r pair-examples}
pr[, `:=`(a_i=ADJ[i+1L], a_j=ADJ[j+1L])]
agree <- pr[cons>3 & cons<5][order(disagree)][1:8, .(pair=paste(a_i,"~",a_j), cons=round(cons,2), disagree=round(disagree,2))]
contest <- pr[order(-disagree)][1:8, .(pair=paste(a_i,"~",a_j), cons=round(cons,2), disagree=round(disagree,2))]
knitr::kable(list(agree, contest), caption="Most-agreed mid-similarity pairs (left) vs most-contested pairs (right).")
```

## 5. Within-rater reliability: do the two directions agree?

A single rater gives two directed observations per pair, `B[i,j]` and `B[j,i]`. Their
correlation is a **within-model** reliability of the symmetric score (a 2-observation
internal consistency), Spearman–Brown-stepped-up to the averaged score:

```{r within}
dir <- fread_gz("dir_ev.csv.gz")
# recover B[i,j] and B[j,i] by joining forward/backward keys
setkey(dir, i, j)
fwd <- dir[sym[, .(i,j)], on=.(i,j)]
bwd <- dir[sym[, .(j,i)], on=.(i=j, j=i)]
wr <- sapply(MODELS, function(m){
  r <- cor(fwd[[m]], bwd[[m]]); c(r=r, sb=2*r/(1+r))
})
wtab <- tibble(model=MODELS, display=DISPLAY[MODELS],
               `dir-dir r`=round(wr["r",],3), `Spearman–Brown`=round(wr["sb",],3),
               `ev_sd`=round(summ$ev_sd[match(MODELS,summ$model)],2)) %>% arrange(desc(`dir-dir r`))
knitr::kable(wtab, caption="Within-model agreement of the two conditional directions.")
```

The two directions correlate only **moderately (0.47–0.78; no rater reaches 0.8)** — so
a *single* directed judgment is a noticeably noisy estimate of the underlying similarity,
and averaging the two directions is a genuine reliability gain (Spearman–Brown steps
0.6 → ~0.75). This mirrors Report II: ~1/5 of each rater's differentiated variance is
direction-inconsistent. Crucially, the ordering is a **third, distinct axis** of rater
quality — it does **not** track the flat/discriminating split: the flat `FalconMamba` is
the *most* internally consistent rater (0.78 — a smooth, stable map), while the flat
`Llama3.2-3B` (0.51) and the high-variance `Gemma3-4B`/`Aya` are the *least*. So within-,
between-rater, and information axes each rank the raters differently; "flat" implies
neither low agreement (§3) nor low within-rater consistency. Only `Llama3.2-3B` is weak
on *both* variance and direction-consistency — the one rater whose scant signal is
largely noise.

## 6. Structural agreement (marginal-free)

Repeating the headline inter-rater agreement on the **double-centered** similarity
(prevalence + generosity swept out) — the strict "same covariance shape" test:

```{r structural}
Sdc <- sapply(MODELS, function(m){
  M <- build_matrix(sym, m, "sym"); upper_vec(double_center(M))
})
Craw <- cor(X); Cdc <- cor(Sdc)
tibble(Object=c("Raw symmetric ratings","Double-centered similarity"),
       `mean pairwise r`=round(c(mean(Craw[upper.tri(Craw)]), mean(Cdc[upper.tri(Cdc)])),3),
       `ICC(3,k) k=12` = c(round(icc_from_vc(vc,12)[4],3), NA)) %>%
  knitr::kable(caption="Agreement drops but stays substantial once shared marginals are removed.")
```

Stripping the shared prevalence marginal lowers mean pairwise agreement from 0.76 to a
still-substantial level — the raters agree on real *structure*, not merely on shared base
rates. (The residual is the material Report IV factors.)

## Verdict

1. **A single LLM's implicit personality theory is a moderately reliable instrument**
   (ICC(2,1) ≈ 0.61); the **dependable measurement is the cohort consensus**
   (ICC(2,k) ≈ 0.95). Treat "the JUDGE similarity" as an ensemble quantity.
2. **~4 good raters reach G ≥ 0.8**; the ensemble's reliability saturates well before 12.
3. **The consensus is robust; don't reflexively trim.** The flat raters *agree* with the
   crowd and don't inject noise — they are low-information, not low-quality. The genuine
   (mild) outliers are the high-variance *idiosyncratic* models (`Gemma-4-31B`, `Aya`).
   Trimming either extreme nudges single-rater ICC via opposite variance mechanisms, but
   the ensemble reliability is essentially invariant and competence-weighting buys almost
   nothing (Report V) — the naive "down-weight the flat raters" intuition is unsupported.
4. **Disagreement lives in the mid-similarity band** and coincides with within-rater
   uncertainty — the unreliability is construct-located (theory-laden co-occurrence),
   which is the *right* place for it to be.
5. **~0.07 single-rater reliability is pure leniency confound** — removable, and a reason
   to prefer consistency-type coefficients (or to center raters) in any cross-model use.
```
