---
title: "JUDGE channel — IV. Internal structure & validity evidence"
subtitle: "Dimensionality, structural invariance across raters, and what it does (and doesn't) license"
author: "ecb"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 9
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 110)
source("_common.R"); suppressMessages(library(psych))
```

## Frame, and a caution held throughout

In the Messick / *Standards* sense, **internal-structure evidence** asks whether the
relationships among the parts of a measure conform to the construct it claims. For the
JUDGE similarity that is two questions:

1. **Dimensionality & pattern** — does the symmetrized similarity organize into
   interpretable trait dimensions?
2. **Structural invariance** — do the 12 raters *agree* on that structure (is it a
   property of the measurement, not of one model)?

I keep one caution front and center (ecb's steer): **recovering an interpretable — even
Big-Five-shaped — structure is consistent with measuring personality description, but is
not by itself proof of validity.** A live alternative is that the model encodes the
*lexical/evaluative co-occurrence of trait words* (the "concept of personality," not a
personality; rgb's `representation_not_intention`). Internal structure can establish that
the instrument is *coherent and reproducible*; it cannot, alone, adjudicate what the
coherent thing *is*. Criterion evidence against humans is a separate — and, per ecb,
inferentially weaker — strand, deferred to a follow-up.

The measurement object is the **cohort-consensus** symmetrized similarity (Report III:
the dependable unit), built over the 10 discriminating raters (the 2 range-restricted
raters add little *differentiated* structure and, since they agree with the crowd,
including them shifts the consensus negligibly — Report V), transformed by
profile-correlation — the same
second-order transform rgb used, chosen because the human reference matrix is itself a
profile correlation over people (kept consistent so any later comparison is
apples-to-apples).

```{r build}
sym <- fread_gz("sym_ev.csv.gz")
disc <- setdiff(MODELS, c("Llama","FalconMamba"))
Sbar <- Reduce(`+`, lapply(disc, function(m) build_matrix(sym, m, "sym"))) / length(disc)
Rcons <- profile_cor(Sbar)
pca_load <- function(R, k){
  e <- eigen(R, symmetric = TRUE)
  L <- e$vectors[,1:k,drop=FALSE] %*% diag(sqrt(pmax(e$values[1:k],0)), k)
  stats::varimax(L)$loadings[]
}
```

## 1. Dimensionality: one dominant evaluative factor, then a trait ladder

```{r scree, fig.height=4.2}
ev <- eigen(Rcons, symmetric = TRUE, only.values = TRUE)$values
scr <- tibble(k = 1:20, eigen = ev[1:20])
ggplot(scr, aes(k, eigen)) + geom_line(colour="grey60") + geom_point(size=2, colour="#4C72B0") +
  geom_hline(yintercept = 1, linetype = 3, colour = "grey60") +
  scale_x_continuous(breaks = seq(0,20,5)) +
  labs(title = "Consensus JUDGE structure — scree (profile-correlation eigenvalues)",
       x = "component", y = "eigenvalue") + theme_judge()
cat(sprintf("PC1 explains %.0f%% of common variance; PC1/PC2 = %.2f. A single strong general factor.\n",
    100*ev[1]/sum(pmax(ev,0)), ev[1]/ev[2]))
```

There is a **strong general factor** — PC1 alone accounts for ~53% of common variance,
3.75× PC2. Substantively this is the model's **evaluative halo**: a good/competent ↔
bad/incompetent axis that pervades every judgment (rgb W15 §2's "PC1 over-collapse";
also present, if less dominant, in the human self-report — human PC1/PC2 ≈ 5.15 per rgb
W16 §9). It is **double-edged for validity**: it is a coherent, reproducible axis (a
plus), *and* it is a discriminant-validity threat — much of the "personality structure"
is one evaluative dimension (§4). After PC1 the scree steps down through a small number
of differentiated factors before a scree tail; a 5-factor solution captures the
interpretable trait structure.

## 2. The five-factor solution resembles the Big Five (read carefully)

```{r factors}
Lc <- pca_load(Rcons, 5)
top_adj <- function(f, n = 10){
  o <- order(-abs(Lc[,f]))[1:n]
  paste0(ADJ[o], " (", sprintf("%+.2f", Lc[o,f]), ")", collapse = ", ")
}
labs <- c("Competence / evaluation (halo)", "Antagonism  (low Agreeableness)",
          "Negative affect  (Neuroticism)", "Playful–uninhibited  (low Conscientiousness / Extraversion)",
          "Traditional–predictable  (low Openness)")
tibble(Factor = paste0("F", 1:5), `Tentative label` = labs,
       `Top-loading adjectives` = sapply(1:5, top_adj, n = 7)) %>%
  knitr::kable(caption = "Varimax(5) of the consensus JUDGE similarity. Labels are interpretive.")
```

Four of the five differentiated factors map recognizably onto **Agreeableness (reversed:
antagonism), Neuroticism, Openness (reversed), and a Conscientiousness/Extraversion
blend**, alongside the dominant **evaluative-competence** factor. This is *internal*
structure — extracted from the models' own judgments, with no human key imposed — so its
Big-Five resemblance is meaningful evidence that the JUDGE instrument organizes trait
words along personality-like lines.

**But note the two hedges the caution demands.** (a) The first and largest factor is
evaluative, not a clean Big-Five trait — an *implicit-personality-theory* halo, which is
exactly what a lexical-co-occurrence account also predicts. (b) The C and E poles are
blended rather than cleanly separated. So "resembles the Big Five" is honest; "recovers
the Big Five" would overclaim.

## 3. Structural invariance: the raters agree on the structure

The stronger internal-structure evidence is **agreement across raters** — if each model
independently yields the same factor pattern, the structure is a property of the
measurement, not an artefact of one model. Aligning each rater's varimax(5) solution to
the consensus (greedy factor matching, sign-corrected) and taking Tucker's congruence:

```{r invariance, fig.height=4.5}
align_cong <- function(Lm, Lc){
  C <- abs(psych::factor.congruence(Lm, Lc))     # 5x5 |congruence|
  best <- numeric(ncol(Lc)); used <- c()
  ord <- order(-apply(C,2,max))
  for(f in ord){ cand <- setdiff(order(-C[,f]), used); pick <- cand[1]; best[f] <- C[pick,f]; used <- c(used,pick) }
  best
}
congmat <- sapply(disc, function(m){
  Lm <- pca_load(profile_cor(build_matrix(sym, m, "sym")), 5)
  align_cong(Lm, Lc)
})
rownames(congmat) <- paste0("F",1:5)
per_model <- colMeans(congmat)
inv <- tibble(model = disc, display = DISPLAY[disc], family = FAMILY[disc],
              mean_cong = per_model) %>% arrange(desc(mean_cong)) %>%
  mutate(display = factor(display, levels = rev(display)))
ggplot(inv, aes(mean_cong, display, colour = family)) +
  geom_segment(aes(x=0.6, xend=mean_cong, yend=display), colour="grey85") +
  geom_point(size=3) + scale_colour_manual(values = FAM_PAL, name = NULL) +
  geom_vline(xintercept = 0.9, linetype = 3, colour="grey55") +
  labs(title="Structural invariance: congruence of each rater's factor solution to the consensus",
       x="mean Tucker congruence to consensus (5 factors)", y=NULL) +
  xlim(0.6, 1) + theme_judge()
cat(sprintf("Mean congruence to consensus across raters: %.2f (Tucker ≥0.85 conventionally 'fair', ≥0.95 'identical').\n",
    mean(per_model)))
```

Mean congruence to the consensus is **0.85** — comfortably in the "fair-to-good"
structural-agreement band, with the Qwen family, Phi4 and Llama-8B near-identical (≥0.90)
and Gemma-4-31B the weakest (consistent with its status as rgb's lone cohort outlier).
**The raters largely agree on the latent structure**, which is genuine
internal-structure reliability: the five-factor organization is not one model's quirk.

```{r per-factor}
tibble(Factor = paste0("F",1:5), Label = labs,
       `Mean congruence` = round(rowMeans(congmat),2),
       `Min across raters` = round(apply(congmat,1,min),2)) %>%
  knitr::kable(caption="Which factors replicate across raters. The evaluative/antagonism/affect factors are the most invariant.")
```

The evaluative, antagonism and negative-affect factors are the most reproducible; the
low-Openness factor is the least invariant — the same weak-Openness signal rgb found in
the *representation*, here at the structural-reliability level.

## 4. The general factor as a discriminant-validity threat

How much of the apparent structure is *just* the evaluative halo? Partialling the first
(general) component out of the consensus and re-extracting shows what differentiated
trait structure survives:

```{r residual}
e <- eigen(Rcons, symmetric=TRUE)
Rres <- Rcons - e$values[1]*tcrossprod(e$vectors[,1])   # remove PC1
evr <- eigen(Rres, symmetric=TRUE, only.values=TRUE)$values
Lres <- pca_load(Rres, 4)
congres <- sapply(disc, function(m){
  Rm <- profile_cor(build_matrix(sym, m, "sym"))
  em <- eigen(Rm, symmetric=TRUE); Rmr <- Rm - em$values[1]*tcrossprod(em$vectors[,1])
  mean(align_cong(pca_load(Rmr,4), Lres))
})
cat(sprintf("After removing the general factor: 4 residual factors still replicate across raters at mean congruence %.2f.\n", mean(congres)))
cat(sprintf("Residual top eigenvalues: %s\n", paste(round(head(evr,5),1), collapse=", ")))
```

Reassuringly, differentiated trait structure **survives** removing the general factor and
**still replicates across raters** — so the instrument is not *only* an evaluative halo;
there is reproducible discriminant structure beneath it. But the halo's dominance (53% of
common variance) is a real limit on discriminant validity: for any downstream use,
trait-specific JUDGE scores should be taken **net of the general evaluative factor**, or
much of what looks like (say) an Agreeableness signal will be halo.

## 5. Response-process evidence (recap)

Internal structure is buttressed by the **response-process** evidence from Report I: for
the discriminating raters the per-judgment entropy traces a sensible inverted-U in
similarity (certain at the poles, graded in the middle), and Report III showed
between-rater disagreement co-locates with within-rater uncertainty. The judgments behave
like a graded construct measurement, not like noise or a degenerate heuristic — a
necessary (not sufficient) condition for the structural evidence to mean anything.

## What this establishes — and what it doesn't

**Established (internal-structure + structural-invariance evidence):**

- The consensus JUDGE similarity is **coherent and low-rank** — one dominant evaluative
  factor plus a small ladder of differentiated trait factors.
- That structure is **reproducible across raters** (mean congruence 0.85; 0.85 again on
  the general-factor-removed residual) — it is a property of the *measurement*, not one
  model. This is real structural reliability, and the strongest validity-relevant claim
  the internal evidence supports.
- The differentiated factors **resemble** Agreeableness, Neuroticism, Openness and a
  C/E blend.

**Not established (and not claimable from this evidence):**

- That the models "have" personalities, or that the structure reflects anything beyond
  **learned lexical/evaluative co-occurrence of trait descriptors** — the general
  evaluative factor is equally predicted by that deflationary account.
- That the structure matches *human* personality structure. That is criterion evidence,
  handled separately and — per ecb — with care: **a match to humans would be consistent
  with validity but is not itself validity**, and a mismatch would not by itself
  disconfirm a coherent, reliable measurement of the models' own implicit personality
  theory.

The honest headline: **as a measurement of the models' implicit personality theory, the
JUDGE instrument is internally coherent, low-dimensional, and structurally reliable
across raters — with an evaluative halo that is simultaneously its strongest factor and
its main threat to discriminant validity.**
```
