---
title: "JUDGE channel — V. Is the crowd flat? Consensus construction & rater quality"
subtitle: "A reactive study: does weighting or trimming the 12-rater ensemble help?"
author: "ecb"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 9
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 110)
source("_common.R")
```

## Why this report exists

This one is **reactive**. Report I's response-style typology suggested an obvious move —
down-weight the two flat, non-discriminating raters (`Llama3.2-3B`, `FalconMamba`) when
forming the cohort consensus, since they carry little differentiated signal. Report III
then found the **opposite** of what that intuition predicts: those flat raters *agree*
with the consensus perfectly well; the raters that actually pull away are the
high-variance *idiosyncratic* models (`Gemma-4-31B`, `Aya`). That tension is worth
resolving directly: **what is the right way to build the JUDGE consensus, and does any
weighting or trimming beat the flat mean?** (rgb's W16 §8b used a flat 12-model average
and flagged that a weighted crowd "might" be better — this tests it.)

```{r load}
sym <- fread_gz("sym_ev.csv.gz"); syh <- fread_gz("sym_hent.csv.gz")
summ <- read_csv_("model_summary.csv") %>% as_tibble()
X  <- as.matrix(sym[, ..MODELS])                    # raw symmetric ratings
H  <- as.matrix(syh[, ..MODELS])                    # symmetric entropy
# double-centered (structural) per-rater vectors — the marginal-free signal
Sdc <- sapply(MODELS, function(m) upper_vec(double_center(build_matrix(sym, m, "sym"))))
nr <- length(MODELS)
```

## 1. Rater-quality anatomy: three axes that disagree

```{r anatomy}
dir <- fread_gz("dir_ev.csv.gz"); setkey(dir,i,j)
fwd <- dir[sym[,.(i,j)], on=.(i,j)]; bwd <- dir[sym[,.(j,i)], on=.(i=j,j=i)]
dirdir <- sapply(MODELS, function(m) cor(fwd[[m]], bwd[[m]]))
loo_dc <- sapply(seq_len(nr), function(r) cor(Sdc[,r], rowMeans(Sdc[,-r])))
anat <- tibble(Model=DISPLAY[MODELS], Family=FAMILY[MODELS],
               `Information (SD)`=round(summ$ev_sd[match(MODELS,summ$model)],2),
               `Between-rater agmt (struct. loo)`=round(loo_dc,3),
               `Within-rater consistency (dir–dir)`=round(dirdir,3)) %>%
  arrange(desc(`Between-rater agmt (struct. loo)`))
knitr::kable(anat, caption="Three orthogonal axes of rater quality. High on one ≠ high on another.")
```

The three quality axes **rank the raters differently** (pairwise rank correlations are
weak): information (differentiation), agreement with the crowd, and internal consistency
are genuinely distinct. `FalconMamba` is *low* information but *high* on both agreement
and consistency; `Aya` is *high* information but *low* agreement; `Gemma-4-31B` is
middling information but lowest agreement. There is no single "good rater" axis to weight by — which already warns that a one-number
competence weight will be crude.

## 2. Does weighting the crowd help? (split-crowd stability)

The clean, external-data-free criterion for consensus quality is **stability**:
partition the raters into two disjoint halves, build a consensus from each, and correlate
them. A better consensus recipe is a more *reproducible* one. I evaluate on the
**double-centered structure** (where differentiation matters; raw ratings are
prevalence-saturated and hide all these differences).

```{r stability}
comp_w <- pmax(loo_dc, 0)^2                          # competence weight (agreement²)
split_stab <- function(M, w=NULL, nrep=200){
  set.seed(11); k<-ncol(M); half<-k%/%2; rs<-numeric(nrep)
  for(b in 1:nrep){ p<-sample(k); a<-p[1:half]; bb<-p[(half+1):(2*half)]
    wa<-if(is.null(w)) rep(1,length(a)) else w[a]
    wb<-if(is.null(w)) rep(1,length(bb)) else w[bb]
    rs[b]<-cor(M[,a]%*%wa/sum(wa), M[,bb]%*%wb/sum(wb)) }
  c(mean=mean(rs), sd=sd(rs))
}
schemes <- tibble(
  Scheme = c("Flat mean (all 12)","Competence-weighted (agreement²)",
             "Drop 2 flat raters","Drop 2 idiosyncratic raters"),
  res = list(split_stab(Sdc), split_stab(Sdc, comp_w),
             split_stab(Sdc[, setdiff(MODELS,c("Llama","FalconMamba"))]),
             split_stab(Sdc[, setdiff(MODELS,c("Gemma4","Aya"))]))
) %>% mutate(`Split-half stability`=round(map_dbl(res, "mean"),4),
             `±SD`=round(map_dbl(res,"sd"),4)) %>% select(-res)
knitr::kable(schemes, caption="Consensus reproducibility under different construction recipes (double-centered structure).")
```

**Weighting and trimming barely move the needle.** Competence-weighting and dropping the
two idiosyncratic raters each lift split-half stability by ~0.003; dropping the two flat
raters slightly *lowers* it (~0.007) — but that is mostly the mechanical cost of a smaller
ensemble (each split-half loses a rater), not evidence the trim removed noise. All recipes
land within ±0.01 of the flat mean (~0.95). The **flat average is already near-optimal** —
the crowd's reproducibility is not being dragged down by any rater. This vindicates rgb's
flat aggregate and closes the Report I intuition: the non-responsive raters are not worth
excluding.

## 3. Per-pair confidence weighting

A subtler idea: weight each rater's judgment of a *specific pair* by its confidence
(inverse entropy on that pair), so a rater's diffuse judgments count less exactly where
it is unsure. Does entropy-weighted pooling produce a more stable consensus?

```{r conf-weight}
# confidence weight per (pair, rater) = 1/(0.2 + entropy); build weighted vs flat consensus,
# compare split-half stability on the *centered* structure.
Wc <- 1/(0.2 + H)
stab_conf <- function(cols, weighted, nrep=150){
  set.seed(12); k<-length(cols); half<-k%/%2; rs<-numeric(nrep)
  Xc <- X[,cols,drop=FALSE]; Wcc <- Wc[,cols,drop=FALSE]
  for(b in 1:nrep){ p<-sample(k); a<-p[1:half]; bb<-p[(half+1):(2*half)]
    mk <- function(idx){ if(weighted) rowSums(Xc[,idx]*Wcc[,idx])/rowSums(Wcc[,idx]) else rowMeans(Xc[,idx]) }
    ca<-mk(a); cb<-mk(bb)
    # center to remove marginals before correlating structure
    rs[b]<-cor(ca-mean(ca), cb-mean(cb)) }
  mean(rs)
}
tibble(Pooling=c("Flat per-pair mean","Confidence (inverse-entropy) weighted"),
       `Split-half stability (raw pairs, centered)` =
         round(c(stab_conf(seq_len(nr),FALSE), stab_conf(seq_len(nr),TRUE)),4)) %>%
  knitr::kable(caption="Per-pair confidence weighting vs flat pooling.")
```

Confidence weighting is **also a wash** — a defensible idea that the data do not reward,
because the raters whose judgments are diffuse (high entropy) are largely the same raters
whose judgments are low-amplitude, so inverse-entropy weighting mostly re-derives the
flat mean. The consensus is robust to how you pool it.

## 4. How many raters does the consensus need? (growth curve)

```{r growth}
set.seed(3)
grow <- map_dfr(2:nr, function(k){
  reps <- replicate(30, {
    s <- sample(nr, k)
    # stability of this k-subset's consensus vs the full-cohort consensus
    cons_k <- rowMeans(Sdc[, s, drop=FALSE])
    cor(cons_k, rowMeans(Sdc))
  })
  tibble(k=k, stab=mean(reps), lo=quantile(reps,.1), hi=quantile(reps,.9))
})
ggplot(grow, aes(k, stab)) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill="#4C72B0", alpha=.15) +
  geom_line(colour="#4C72B0", linewidth=1) + geom_point(size=1.6, colour="#4C72B0") +
  geom_hline(yintercept=.95, linetype=3, colour="grey55") +
  scale_x_continuous(breaks=2:12) +
  labs(title="Consensus convergence: agreement of a k-rater subset with the full cohort",
       x="number of raters in the subset", y="corr. with full 12-rater consensus (structural)") +
  theme_judge()
cat(sprintf("A random %d-rater subset already reproduces the full consensus at r>0.95.\n",
    min(grow$k[grow$stab>0.95])))
```

The consensus **converges fast**: a random ~3 raters reproduce the full-cohort structural
pattern at r > 0.95, and ~5 at r > 0.98. This "reproduce *this* cohort's mean" curve
saturates a little faster than the absolute-reliability D-study of Report III (which needs
~6 raters for G ≥ 0.9) — they answer different questions, but both say the practically-useful
ensemble is *modest* and which specific models you include matters little.

## Verdict

1. **The flat 12-model mean is already near-optimal.** No weighting (competence or
   confidence) or trimming scheme meaningfully beats it on reproducibility. rgb's flat
   aggregate is the right, robust object.
2. **The Report I "down-weight the flat raters" intuition is unsupported.** Low
   *information* ≠ low *quality*: the flat raters agree with and don't destabilize the
   consensus. If anything, the mild outliers are the high-variance idiosyncratic models
   (`Gemma-4-31B`, `Aya`) — and even trimming those buys ~nothing in stability.
3. **Rater quality is genuinely multidimensional** (information / agreement / internal
   consistency are ~orthogonal), which is *why* a single competence weight can't help —
   there is no one axis to weight by.
4. **A handful of raters suffice.** The structural consensus reproduces from even ~3
   raters, and Report III's D-study needs only ~6 for a dependable (G ≥ 0.9) measurement;
   the ensemble's value is real but bounded, and model identity within the crowd is largely
   fungible.

The measurement-design takeaway: **treat the JUDGE similarity as an unweighted
ensemble of ≥6 discriminating raters, symmetrized and double-centered.** Simplicity is
not costing anything here.
```
