---
title: "JUDGE channel — VI. Asymmetry under Tversky's contrast model"
subtitle: "What kind of thing is B[i,j] ≠ B[j,i]? A features-of-similarity reading, and what it says we should collect next"
author: "ecb"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 9
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 110)
source("_common.R")
suppressPackageStartupMessages({library(splines)})
```

## Frame

Report II asked the *measurement* question about `B[i,j] ≠ B[j,i]` — how much, is it
reliable, do we symmetrize — and answered it: ~21% of the centered signal, ≈half a
removable "prevalence gradient", the residual curl "real but faint and idiosyncratic",
so symmetrize and keep direction as a flagged side channel. That verdict stands for the
similarity measurement. But it left the asymmetry *theoretically unclaimed*: a marginal
to strip and a residual we didn't model.

> **Scope note added after Report VII (2026-07-29).** The frame experiment this report
> proposes in §8.2 has now been run, and it constrains what follows. The prominence scale
> recovered below **replicates out-of-sample within rgb's conditional frame** (r = −0.34 to
> −0.70 across five raters on freshly collected data) but is **three to five times weaker
> in explicit similarity and difference frames**. So read this report as a contrast-model
> account **of the conditional item** — where the model does real explanatory work — and
> not as an account of similarity judgment in general. Report VII §8 has the detail.

This report reads the same object through **Tversky (1977), "Features of similarity."**
That is not decoration. Tversky's is the canonical account of *judged* similarity as
non-metric — asymmetric, context-dependent, and generated by a feature-matching process
rather than a distance — and it makes **quantitative, falsifiable predictions about the
exact decomposition Report II computed**. It is also already the psychological grounding
the readout paper leans on for representation-vs-judgment divergence, so it is worth
knowing whether JUDGE actually behaves the way Tversky says judged similarity behaves.

**The contrast model.** For objects *a*, *b* with feature sets *A*, *B* and a
non-negative salience measure *f*,

$$s(a,b) \;=\; \theta\, f(A \cap B) \;-\; \alpha\, f(A - B) \;-\; \beta\, f(B - A),
\qquad \theta,\alpha,\beta \ge 0 .$$

The judgment is **directed**: *a* is the **subject**, *b* the **referent**, and the
subject's distinctive features are weighted more heavily (α > β). JUDGE's prompt fixes
the roles unambiguously —

> *Consider a person who is very **{i}**. How likely is this person to also be **{j}**?*

— so **i is the subject and j the referent**, and `B[i,j]` is a contrast-model
similarity in Tversky's sense. Two consequences drive everything below.

**Prediction 1 (separability).** If *f* is additive over the shared/distinctive split —
`f(A) = f(A∩B) + f(A−B)` — then

$$B[i,j] - B[j,i] \;=\; (\beta-\alpha)\,\bigl(f_i - f_j\bigr),$$

which is a **pure gradient**: the antisymmetric part is *exactly* the difference of a
single per-adjective potential, with **zero curl**. So Report II's Helmholtz–Hodge split
is not an arbitrary chart — the gradient share *is* the contrast model's fit, and the
curl is precisely the variance the additive contrast model cannot produce.

**Prediction 2 (direction).** Since α > β, `s(a,b) > s(b,a)` iff `f(B) > f(A)`: the
**less prominent object makes the better subject**. Tversky's examples are "North Korea
is more similar to China than China is to North Korea" and "the variant is more similar
to the prototype than the prototype to the variant." Here that means specific/rare traits
should imply general/prominent ones far more than the reverse.

**Prediction 3 (diagnosticity).** *f* is not context-free: the weight a feature carries
depends on what it classifies in the current stimulus set. Non-additivity of *f* is the
mechanism, and it is the only thing in the theory that can generate curl. So under this
reading the curl is not leftover noise — it is **where the diagnosticity effect lives**,
and it should be *structured*, not random.

```{r load}
sym  <- fread_gz("sym_ev.csv.gz")
asy  <- fread_gz("asym_ev.csv.gz")
dir  <- fread_gz("dir_ev.csv.gz")
marg <- read_csv_("marginals.csv") %>% as_tibble()
n    <- nrow(adjectives)
disc <- setdiff(MODELS, c("Llama", "FalconMamba"))   # 10 discriminating raters

Dl <- lapply(MODELS, function(m) { D <- build_matrix(asy, m, "asym"); D[is.na(D)] <- 0; D })
Sl <- lapply(MODELS, function(m) build_matrix(sym, m, "sym"))
names(Dl) <- names(Sl) <- MODELS
Gm <- sapply(MODELS, function(m) rowMeans(Dl[[m]]))          # potentials, 523 x 12
Cl <- lapply(MODELS, function(m) Dl[[m]] - outer(Gm[,m], Gm[,m], "-"))  # curl
names(Cl) <- MODELS

Dbar <- Reduce(`+`, Dl[disc]) / length(disc)
Cbar <- Reduce(`+`, Cl[disc]) / length(disc)
Sbar <- Reduce(`+`, Sl[disc]) / length(disc)
Bbar <- Reduce(`+`, lapply(disc, function(m) build_matrix(dir, m, "dir"))) / length(disc)
gbar <- rowMeans(Gm[, disc])
fbar <- -gbar                                                # prominence: high f = strong referent
```

## 1. How much of the asymmetry is the contrast model? (about half)

```{r sep}
sep <- sapply(MODELS, function(m) sum(outer(Gm[,m], Gm[,m], "-")^2) / sum(Dl[[m]]^2))
knitr::kable(
  tibble(Model = DISPLAY[MODELS], Family = FAMILY[MODELS],
         `Separable (contrast-model) share of asymmetry` = round(100*sep, 1)) %>%
    arrange(desc(`Separable (contrast-model) share of asymmetry`)),
  caption = "Share of each rater's antisymmetric variance that an additive-f contrast model can produce.")
cat(sprintf("mean %.1f%%   |   consensus (10-rater mean matrix) %.1f%%\n",
            100*mean(sep),
            100*sum(outer(gbar,gbar,"-")^2)/sum(Dbar^2)))
```

**Mean 50%, consensus 57%.** Half of the directional signal is exactly the shape
Prediction 1 requires, and the half that isn't is the theory's own residual category
(Prediction 3). Note that this is a *substantive model fit*, not a coordinate choice:
"one number per adjective explains half the 273,006-cell directional matrix" is a strong
constraint that a matrix of directional idiosyncrasies had no obligation to satisfy.

The spread across raters is itself informative — `Phi4` and the Llamas are nearly
separable (57–64%), the Gemmas and `Aya` much less so (36–46%). Diagnosticity effects
are stronger in the models with more differentiated judgment (§5, §7).

## 2. The recovered prominence scale, and Prediction 2

The least-squares potential `g[i] = mean_j (B[i,j] − B[j,i])` estimates `(β−α)(f_i − f̄)`.
Since α > β the coefficient is negative, so **f ∝ −g**: adjectives with the most negative
`g` are the most prominent (best referents), those with the most positive `g` the least.

```{r prominence}
mg <- marg %>% filter(model %in% disc) %>% group_by(idx) %>%
  summarise(prev = mean(col_mean), gen = mean(row_mean), .groups = "drop") %>% arrange(idx)
Rcons <- profile_cor(Sbar); ee <- eigen(Rcons, symmetric = TRUE)
halo <- ee$vectors[,1] * sqrt(max(ee$values[1], 0))
if (halo[match("good", ADJ)] < 0) halo <- -halo          # orient: + = desirable
P <- tibble(adjective = ADJ, f = fbar, prev = mg$prev, halo = halo,
            breadth = rowMeans(Sbar, na.rm = TRUE))

bind_rows(
  P %>% arrange(desc(f)) %>% slice(1:10) %>% mutate(end = "most prominent (best referents)"),
  P %>% arrange(f)       %>% slice(1:10) %>% mutate(end = "least prominent (best subjects)")) %>%
  transmute(End = end, Adjective = adjective, f = round(f,2),
            `base rate (col mean)` = round(prev,2), `halo (valence)` = round(halo,2)) %>%
  knitr::kable(caption = "The two ends of the recovered prominence scale f = -g.")
```

The high-*f* end is near-universal states — *thinking, awake, normal*, plus the everyday
negative affects (*exhausted, tired, frustrated, annoyed*). The low-*f* end is rare,
highly specific attributes — *blind, tiny, handicapped, disabled, senile, famous*. That
is the prototype/variant contrast in the trait lexicon.

```{r pred2}
pair_row <- function(a, b) {
  ia <- match(a, ADJ); ib <- match(b, ADJ)
  tibble(`subject ⇒ referent` = paste(a, "⇒", b), fwd = Bbar[ia, ib],
         `reverse` = Bbar[ib, ia], `Δ` = Bbar[ia, ib] - Bbar[ib, ia],
         `f(subj)` = fbar[ia], `f(ref)` = fbar[ib])
}
bind_rows(pair_row("blind","thinking"), pair_row("retarded","awake"),
          pair_row("senile","tired"),   pair_row("famous","normal"),
          pair_row("handicapped","emotional"), pair_row("tiny","uncomfortable")) %>%
  mutate(across(where(is.numeric), ~round(.,2))) %>%
  knitr::kable(caption = "Prediction 2: the low-prominence member is the better subject. Cohort consensus, 1-7 Likert.")

ut   <- upper.tri(Dbar)
pred <- outer(fbar, fbar, function(a, b) sign(b - a))[ut]
gp   <- abs(outer(fbar, fbar, "-"))[ut]
agree <- sign(Dbar[ut]) == pred
tibble(`Pairs considered` = c("all pairs", "prominence gap above median",
                              "prominence gap in the top decile"),
       n = c(sum(ut), sum(gp > median(gp)), sum(gp > quantile(gp, .9))),
       `% asymmetric in the predicted direction` =
         round(100*c(mean(agree), mean(agree[gp > median(gp)]),
                     mean(agree[gp > quantile(gp, .9)])), 1)) %>%
  knitr::kable(caption = "Sign agreement with Prediction 2, by how far apart the two adjectives are in prominence.")
```

Gaps of 4–5 Likert points in the predicted direction, and the sign is right for 76% of
all 136,503 unordered pairs, rising to **97% once the two adjectives differ appreciably in
prominence** (top-decile gap) — the disagreements are concentrated where the model
predicts no asymmetry anyway. **Prediction 2 is confirmed about as cleanly as an
observational test can confirm it.** (The sign test is not fully independent of the
estimate — *f* is fitted from these same differences — so read it as a check that a single
transitive ordering *fits*, not as out-of-sample validation. §3 supplies the external
evidence.)

## 3. Is *f* a feature measure, or just a base rate?

The obvious deflationary rival: JUDGE asks a conditional-probability question, so
`B[i,j] ≈ P(j | i)`, and `P(j|i) − P(i|j)` is monotone in the base-rate difference for
any symmetric association. On that account *f* is nothing but prevalence and Tversky adds
no content. The base rate is indeed the single biggest correlate — but it is **not the
whole story**, and the deviation is lawful.

```{r whatisf}
knitr::kable(round(cor(P %>% select(f, prev, halo, breadth), use = "pairwise"), 2),
             caption = "Correlates of the recovered prominence scale.")
```

```{r spline}
zipf <- read_csv_("adj_freq.csv")$zipf                 # wordfreq Zipf, lexical availability
zipf[zipf == 0] <- NA; zipf[is.na(zipf)] <- min(zipf, na.rm = TRUE)
Pz <- P %>% mutate(zipf = zipf) %>%
  mutate(across(c(f, prev, halo, zipf), ~as.numeric(scale(.))))
m0 <- lm(f ~ ns(prev, 6), Pz); m1 <- lm(f ~ ns(prev, 6) + halo, Pz)
m2 <- lm(f ~ ns(prev, 6) + halo + zipf, Pz)
tibble(Model = c("flexible spline in base rate", "+ evaluative halo", "+ lexical frequency"),
       `R²` = round(c(summary(m0)$r.squared, summary(m1)$r.squared, summary(m2)$r.squared), 3),
       `ΔR²` = round(c(NA, summary(m1)$r.squared - summary(m0)$r.squared,
                       summary(m2)$r.squared - summary(m1)$r.squared), 3),
       `p (added term)` = c(NA, anova(m0,m1)$`Pr(>F)`[2], anova(m1,m2)$`Pr(>F)`[2])) %>%
  knitr::kable(caption = "Nested models for f. The base rate is given every chance (6-df natural spline).")
cat(sprintf("halo coefficient net of the prevalence spline: %+.2f (t = %.1f)\n",
            coef(m2)[["halo"]], summary(m2)$coefficients["halo","t value"]))
```

A **triple dissociation**:

- **Base rate** explains 72% of *f*. It is the dominant component and the deflationary
  reading is *mostly* right.
- **Valence adds ΔR² = 0.12 (t ≈ −19) net of a 6-df spline in the base rate** — so it is
  not a curvature artefact of the prevalence relation. The sign is **negative**:
  *at equal base rate, an undesirable trait is the stronger referent.* Every one of the
  12 raters shows it (partial coefficients −0.06 to −0.62).
- **Lexical frequency contributes nothing** (ΔR² = 0.000, p = 0.51), despite ranging over
  5 Zipf units. So *f* is not availability-of-the-word.

The negative valence term is exactly what a features account predicts and a base-rate
account does not: **negative person-features are more diagnostic** — in Tversky's terms
they carry more weight in the matching function; in the person-perception literature this
is the standard negativity/diagnosticity asymmetry (a single dishonest act is more
informative about honesty than a single honest one). *f* is a **salience measure with a
prevalence component and an independent negativity component**, not a probability.

## 4. Prediction 1 tested properly: is *f* additive?

Additivity is what makes the asymmetry *separable*. If it holds, `|Δ|` depends on the
pair only through the prominence gap — in particular it should carry **no residual
relation to how similar the pair is** once the gap is controlled. (An interactive,
non-additive *f* would make `|Δ|` shrink or grow with overlap at fixed gap.)

```{r additive}
ij  <- as.matrix(asy[, .(i,j)]) + 1L
d   <- Dbar[ij]; s <- Sbar[ij]; gap <- abs(fbar[ij[,1]] - fbar[ij[,2]])
tibble(Relation = c("|Δ| ~ similarity", "|Δ| ~ prominence gap",
                    "|Δ| ~ similarity, gap partialled"),
       r = round(c(cor(abs(d), s), cor(abs(d), gap),
                   cor(resid(lm(abs(d) ~ gap)), resid(lm(s ~ gap)))), 3)) %>%
  knitr::kable(caption = "Does the size of the asymmetry depend on anything but the prominence gap?")
```

```{r additive-plot, fig.height=4}
tibble(s = s, ad = abs(d), gap = gap) %>%
  mutate(bin = ntile(s, 10)) %>% group_by(bin) %>%
  summarise(sim = mean(s), `|Δ|` = mean(ad), `prominence gap` = mean(gap), .groups="drop") %>%
  pivot_longer(c(`|Δ|`, `prominence gap`)) %>%
  ggplot(aes(sim, value, colour = name)) + geom_line() + geom_point(size = 2) +
  scale_colour_manual(values = c("#4C72B0", "grey60"), name = NULL) +
  labs(title = "Asymmetry size tracks the prominence gap, not overlap",
       subtitle = "deciles of symmetric similarity, cohort consensus",
       x = "symmetric similarity (Likert)", y = NULL) + theme_judge()
```

`|Δ|` correlates 0.62 with the prominence gap and **−0.01 with similarity once the gap is
partialled** — the apparent decline of `|Δ|` at high similarity is entirely the gap
shrinking. The additive-*f* contrast model is not rejected on its own terms; its failure
is confined to the curl, which by construction it cannot produce at all.

## 5. The curl is diagnosticity — and it is far more reproducible than Report II implied

Report II correlated the curl **cell by cell** across raters (r̄ ≈ 0.34–0.40) and
concluded it was "real but faint and idiosyncratic." That is the wrong resolution for a
context-dependence effect: diagnosticity shifts weights at the level of *categories*, so
the shared part should be **low-rank**, and cell-level correlation charges it for all the
per-cell noise. A skew-symmetric matrix has a natural low-rank structure — orthogonal
**planes** of circulation (its SVD comes in pairs) — so the right question is whether the
raters agree on the *planes*.

```{r curl-planes}
svb <- svd(Cbar)
cum <- cumsum(svb$d^2)/sum(svb$d^2)
top_plane <- lapply(disc, function(m) svd(Cl[[m]])$u[,1:2]); names(top_plane) <- disc
subcong <- function(A, B) mean(svd(t(A) %*% B)$d)     # mean canonical correlation of subspaces
pc <- outer(seq_along(disc), seq_along(disc),
            Vectorize(function(a,b) subcong(top_plane[[a]], top_plane[[b]])))
set.seed(2)
pnull <- replicate(20, { pm <- sample(n); U <- svd(Cl[[disc[1]]][pm,pm])$u[,1:2]
                         mean(sapply(disc[-1], function(m) subcong(U, top_plane[[m]]))) })
cell <- cor(sapply(disc, function(m) Cl[[m]][ij]))
tibble(`Level of description` = c("individual cells (Report II)",
                                  "dominant curl plane (subspace congruence)",
                                  "  ...permuted-adjective null"),
       `agreement across the 10 raters` = round(c(mean(cell[upper.tri(cell)]),
                                                  mean(pc[upper.tri(pc)]), mean(pnull)), 3)) %>%
  knitr::kable(caption = "The curl's reproducibility depends entirely on the resolution you ask at.")
cat(sprintf("Consensus curl: first plane = %.0f%% of curl variance, first three planes = %.0f%%.\n",
            100*cum[2], 100*cum[6]))
```

**The dominant curl plane replicates at r̄ = 0.73 against a permutation null of 0.05.**
The curl is not idiosyncratic — it is a *low-rank, strongly shared* directional structure
that cell-wise correlation hid. It is also genuinely low-rank: one plane carries 47% of
the consensus curl variance, three carry 65%.

```{r curl-axes}
ax <- function(k, m = 10) { o <- order(-abs(svb$u[,k]))[1:m]
  paste0(ADJ[o], sprintf("(%+.2f)", svb$u[o,k]), collapse = ", ") }
tibble(Axis = c("plane-1, axis 1", "plane-1, axis 2"),
       `Highest-|loading| adjectives` = c(ax(1), ax(2))) %>%
  knitr::kable(caption = "What the dominant circulation plane is spanned by.")
```

Axis 1 is the evaluative axis (*valuable, awesome, lovable, respectable* vs *pathetic,
senile, good-for-nothing, retarded*). Axis 2 separates **misfortune** (*blind, disabled,
handicapped, unattractive*) from **malevolence** (*evil, abusive, insane*) — two ways of
being negatively valued that the models keep distinct and circulate between. That is a
diagnosticity structure in the strict sense: which features get weight depends on which
kind of "bad" the subject has put the person into.

Projecting the curl onto the five-factor basis from Report IV gives the construct-level
picture:

```{r curl-factors}
pca_load <- function(R, k) { e <- eigen(R, symmetric = TRUE)
  L <- e$vectors[,1:k,drop=FALSE] %*% diag(sqrt(pmax(e$values[1:k],0)), k)
  stats::varimax(L)$loadings[] }
L5 <- pca_load(Rcons, 5); W <- apply(L5, 2, function(v) v/sqrt(sum(v^2)))
flab <- c("F1 evaluative/competence","F2 antagonism","F3 negative affect",
          "F4 playful–uninhibited","F5 traditional–predictable")
Pm <- t(W) %*% Cbar %*% W; dimnames(Pm) <- list(flab, flab)
knitr::kable(round(Pm, 1),
  caption = "Consensus curl in the 5-factor basis. Row = subject, column = referent; positive = subject⇒referent endorsed more than the reverse, NET of prevalence.")
Pf <- sapply(disc, function(m) (t(W) %*% Cl[[m]] %*% W)[upper.tri(diag(5))])
set.seed(3)
nullf <- replicate(200, { R <- qr.Q(qr(matrix(rnorm(n*5), n, 5)))
  cc <- cor(sapply(disc, function(m) (t(R) %*% Cl[[m]] %*% R)[upper.tri(diag(5))]))
  mean(cc[upper.tri(cc)]) })
obs <- { cc <- cor(Pf); mean(cc[upper.tri(cc)]) }
cat(sprintf("cross-rater agreement of the 5x5 projected curl: r-bar = %.3f  (random-basis null %.3f, sd %.3f, p < %.3f)\n",
            obs, mean(nullf), sd(nullf), max(mean(nullf >= obs), 1/200)))
g5 <- rowMeans(Pm)
cat(sprintf("gradient share WITHIN the 5x5 curl: %.2f  — net source: %s\n",
            sum(outer(g5,g5,"-")^2)/sum(Pm^2), flab[which.max(g5)]))
```

At construct resolution the raters agree at **r̄ = 0.72** (random-basis null 0.30, p < 0.005). And the
construct-level flow has a clear shape: **negative affect is the net source** — anchoring
a person on F3 implies antagonism and low competence much more strongly than anchoring on
antagonism or incompetence implies negative affect. A third of the 5×5 residual is itself
a further additive potential, i.e. **after removing the adjective-level prominence
gradient there is a second-order prominence ordering at the trait-factor level.**

One warning about reading examples off the curl. The curl is a *residual*, and for these
pairs the prevalence gradient dominates the raw judgment, so the raw and residual
directions can oppose:

```{r curl-caution}
ex <- function(a,b){ ia<-match(a,ADJ); ib<-match(b,ADJ)
  tibble(pair = paste(a,"⇒",b), `raw Δ` = Dbar[ia,ib], `curl (residual) Δ` = Cbar[ia,ib]) }
bind_rows(ex("frustrated","rude"), ex("sad","hostile"), ex("anxious","critical")) %>%
  mutate(across(where(is.numeric), ~round(.,2))) %>%
  knitr::kable(caption = "Raw vs prevalence-free direction. Negative-affect words are highly prevalent referents, so the gradient masks the residual flow.")
```

## 6. A metric-axiom check that comes out *negative* (usefully)

Tversky & Hutchinson (1986) give the sharpest test of whether a proximity structure can
be a *space* at all: in Euclidean *d*-space the number of points that can share one
nearest neighbour is bounded (2 in 1-D, 5 in 2-D, 11 in 3-D). Perceptual data respect the
bound; conceptual/taxonomic data routinely blow through it, because feature structures
have focal points that spatial models cannot host. If the trait lexicon were feature-like
in that sense, embedding it — which everything downstream, including rgb's geometric
track, does — would be lossy in a specific diagnosable way.

```{r nn}
nnC <- function(M) { diag(M) <- -Inf; N <- tabulate(apply(M, 1, which.max), nbins = n)
  c(C = sum(N^2)/n, maxN = max(N)) }
raw <- nnC(Sbar); dc <- nnC(double_center(Sbar))
Sd <- Sbar; diag(Sd) <- -Inf; Ni <- tabulate(apply(Sd, 1, which.max), nbins = n)
o <- order(-Ni)
cat(sprintf("consensus symmetric: C = %.2f, max N_i = %d   |   double-centered: C = %.2f, max N_i = %d\n",
            raw["C"], raw["maxN"], dc["C"], dc["maxN"]))
cat("top nearest-neighbour attractors:",
    paste0(ADJ[o[1:10]], "(", Ni[o[1:10]], ")", collapse = ", "), "\n\n")
knitr::kable(tibble(Model = DISPLAY[MODELS],
  C = round(sapply(MODELS, function(m) nnC(Sl[[m]])["C"]), 2),
  `max N_i` = sapply(MODELS, function(m) nnC(Sl[[m]])["maxN"])) %>% arrange(desc(C)),
  caption = "Nearest-neighbour centrality per rater.")
```

**No focal-point violation.** `max N_i = 9` needs only three Euclidean dimensions, and
C = 2.31 is unremarkable. The *symmetric* JUDGE similarity is comfortably spatial — a
point in favour of the geometric treatments, and an honest negative result for the
"features not space" story: **JUDGE is non-metric in its asymmetry (§1–§5), not in its
symmetric part.** The two raters with anomalous centrality (`Qwen2.5-7B` 9.5,
`Llama3.2-3B` 8.3) are the range-restricted ones — this is a rater-quality signal, not a
construct one, and it is a cheap new diagnostic for flat raters.

## 7. The contrast weight is a rater property

Scale-free per-rater measures of how strongly the subject-role dominates: the spread of
the recovered potential relative to the rater's own judgment spread.

```{r rater}
map_dfr(MODELS, function(m) { S <- Sl[[m]]; diag(S) <- NA; off <- !is.na(S)
  tibble(Model = DISPLAY[m], Family = FAMILY[m],
         `SD of ratings` = sd(S[off]),
         `asymmetry weight  sd(g)/sd(B)` = sd(Gm[,m])/sd(S[off]),
         `separable share` = sep[m],
         `corr(f, symmetric marginal)` = cor(-Gm[,m], rowMeans(S, na.rm = TRUE))) }) %>%
  mutate(across(where(is.numeric), ~round(.,3))) %>% arrange(desc(`asymmetry weight  sd(g)/sd(B)`)) %>%
  knitr::kable(caption = "Contrast-model parameters as per-rater properties.")
```

The asymmetry weight varies ~1.7× (0.36–0.64) and is **not** a size or family effect. The
last column is a genuine over-identification test: the contrast model says the *same* *f*
governs the asymmetry and the symmetric marginal, so the correlation should be strongly
positive. It is, for the larger and more differentiated raters (`Qwen2.5-32B` 0.84,
`Gemma-4-31B` 0.83, `Qwen2.5-7B`/`Phi4` 0.73) — and it **collapses to ~0 for
`Llama3.2-3B` (0.01) and `Aya` (0.16)**. For those raters the directional signal and the
similarity signal do not share a prominence scale, i.e. the contrast model does not hold
of them at all. That is a useful, theory-derived rater-fit statistic, and it agrees with
the response-style verdicts from Report I rather than duplicating them.

## 8. What we cannot identify, and the right experiment for it

The one contrast-model quantity the collected data cannot deliver is the split between
**α and β separately** — we identify only `(β − α)f` from the skew part and
`−((α+β)/2)f + θc̄` from the symmetric marginal, which is one equation short because θ and
the mean commonality `c̄` are entangled with it. Everything in §1–§7 is a statement about
`(β − α)f` and about the shape of the residual; the *focusing* weight itself is not
estimated.

### 8.1 Why the diagonal is the wrong instrument for it

Tversky's own identification is the **diagonal**: `s(a,a) = θ f(A)`, since the distinctive
sets are empty. rgb's `dists` tensor has a NaN diagonal — the 523 self-pairs were never
collected — so this looks like a cheap fix. It is not, for three reasons, and the third is
decisive.

**(a) The self-pair task is pragmatically degenerate.** "Consider a person who is very
*kind*. How likely is this person to *also* be *kind*?" — *also* presupposes a distinct
second predicate, so the question is ill-formed. Any variance returned is variance in how
a model handles an ill-formed question. Worse, a model that resolves the anomaly by
reading *also* as "additionally, over and above being kind" is answering a **different
question** than it answers off-diagonal, in which case the diagonal is not on the same
scale as the matrix it was supposed to anchor.

**(b) The operationalization does not transfer.** Tversky did not obtain `s(a,a)` by
asking how similar a thing is to itself. He read it off **identification/confusion**
data — Rothkopf's Morse-code matrix, letter-confusion matrices — where `p(a → a)` falls
out of a task that is perfectly well-formed. JUDGE has no confusion task, so the
psychometric move has no vehicle here.

**(c) Ceiling — the decisive one.** Substituting each adjective's best available
near-synonym for literal identity (distinctive sets near-empty, so `s ≈ θf(A)`) shows
where the diagonal would land:

```{r ceiling}
Sd <- Sbar; diag(Sd) <- NA
ceil <- apply(Sd, 1, max, na.rm = TRUE)
nb   <- apply(Sd, 1, function(r) which.max(replace(r, is.na(r), -Inf)))
cat(sprintf("near-identity ceiling max_j S[i,j]: mean %.2f, sd %.2f, range %.2f-%.2f\n",
            mean(ceil), sd(ceil), min(ceil), max(ceil)))
cat(sprintf("  fraction already >= 6.5: %.0f%%   |   >= 6.9: %.0f%%   (scale max = 7)\n",
            100*mean(ceil >= 6.5), 100*mean(ceil >= 6.9)))
cat(sprintf("  correlation with the recovered prominence f: %.2f\n", cor(ceil, fbar)))
bind_rows(
  tibble(End = "lowest ceiling",  i = order(ceil)[1:6]),
  tibble(End = "highest ceiling", i = order(-ceil)[1:6])) %>%
  transmute(End, Adjective = ADJ[i], `ceiling` = round(ceil[i], 2),
            `best available near-synonym` = ADJ[nb[i]]) %>%
  knitr::kable(caption = "Where a self-pair would land: the near-identity ceiling.")
```

**58% of adjectives already sit at ≥ 6.5 and 11% at ≥ 6.9 on a 7-point scale.** The literal
self-pair would be a wall of 7s for the large majority, leaving almost no variance to read
`f` from — and `f` has to span the distance between *thinking* and *blind* to be worth
anything. The near-synonym proxy does not rescue it either: it correlates only 0.46 with
`f`, and the low-ceiling cases are low because the 523-word set happens to contain no
synonym for them (*retarded*, *blind*, *left-handed*), which is a **coverage artefact of
the item pool**, not low self-similarity.

**What the diagonal *is* worth** is a smaller and honest claim: an **instrument-integrity
check**. A rater that answers 5 rather than 7 to "very kind ⇒ kind" is reading a contrast
presupposition into *also* — which would mean every off-diagonal cell means something
other than what this whole track has assumed. Ruling that out is worth 523 cheap prompts.
It would also calibrate each rater's true ceiling — `FalconMamba` and `Llama3.2-3B` never
exceed 6.25 and 6.03 *anywhere* in the matrix, so their compression is currently
confounded with their content. Useful; not an identification of α and β.

### 8.2 The right experiment: the dissimilarity frame

Tversky's own lever for separating the weights is to ask for **difference** rather than
similarity. The same features are available, but the weights on the *distinctive* sets
change, and the asymmetry can **reverse**. That is the manipulation to run, and it
adjudicates §3's deflationary rival in a way the diagonal never could:

> *Consider a person who is very **{i}**. How **unlike** a person who is very **{j}** are
> they?*

Writing `c = f(A∩B)`, `u = f(A−B)`, `v = f(B−A)`, the two tasks are
`s(a,b) = θc − αu − βv` and `d(a,b) = α'u + β'v − θ'c`, so

$$\Delta s = (\beta-\alpha)(u-v), \qquad \Delta d = (\alpha'-\beta')(u-v).$$

**A sign flip is therefore *not* the discriminating prediction** — and it is worth being
explicit that my first statement of this design said it was, wrongly. With α > β and
α' > β' the raw asymmetry does invert between the frames, but so does any reverse-scored
measure, so a conditional-probability account with flipped polarity predicts exactly the
same inversion. After reverse-scoring, the two accounts agree on sign. What separates
them is:

1. **Non-complementarity (the primary test).**
   `s(a,b) + d(a,b) = (θ−θ')c + (α'−α)u + (β'−β)v`. If difference is merely similarity
   reverse-scored, this sum is **constant**. Under the contrast model the weights are
   re-tuned between tasks, so the sum varies systematically with the pair's feature
   richness — pairs high in *both* common and distinctive features are judged both more
   similar and more different (Tversky & Gati's result in humans). Regress `s + d` on the
   overlap proxy (symmetric similarity, from the collected 273k cells) and the richness
   proxy `f_i + f_j`: **flat ⇒ probability account, structured ⇒ contrast model.**
2. **The asymmetry-to-signal ratio should differ by frame.** Distinctive features carry
   more weight in difference judgments, so `|Δd|/SD(d)` should exceed `|Δs|/SD(s)`.
   A polarity reversal predicts them equal.
3. **Frame-invariance of *f*.** The same prominence scale recovered in §2 from 273,006
   cells should predict the directional asymmetry in *both* new frames. This is genuine
   out-of-sample validation of §2, which the sign test in §2 explicitly was not.
4. **Does `cond` behave like `sim` at all?** Report VI *assumes* `B[i,j]` is a
   contrast-model similarity. Running rgb's original conditional frame alongside an
   explicit similarity frame on the same pairs **tests that assumption** rather than
   granting it — and a divergence would be a scope limit on this entire report.

The task is well-formed, so neither failure mode in §8.1 applies, and it does not need the
full 273,006 cells: pairs sampled at **high prominence gap** — where §2 shows the
directional effect is 97% consistent, so the test has power — plus a matched low-gap
control block, each stratified across the similarity range so that `c` and `u+v` vary
independently enough for test 1.

**Collected and analysed 2026-07-28/29 — see Report VII.** Headline: the gradient is
largely frame-bound (see the scope note at the top); non-complementarity (T1) is real and
increases monotonically with model scale within the Gemma family, running through *overlap*
rather than the predicted richness channel.

**Collected 2026-07-28** — `sample_frame_pairs.R` (252 pairs: 2 prominence-gap blocks × 3
similarity terciles × 42) and `run_frames_ollama.py` (3 frames × 2 orders = 1,512 prompts
per model) on the four Orin cohort models. Results in Report VII.

Four cohort models (`gemma3:4b`, `qwen2.5:3b`, `llama3.2:3b`, `phi4-mini`) are in the Orin
lineup, so a 4-model × 3-frame design over a few thousand pairs is a modest one-off run —
and it is the same referent-swap logic the readout account already budgets for, applied to
the one place where the two theories make opposite predictions.

## Verdict — and two amendments to Report II

Tversky's contrast model is not a loose analogy for JUDGE; it is a **fitted model that
explains half the directional variance with one parameter per adjective** and correctly
predicts the direction of the asymmetry across essentially the whole item set. Two things
in Report II should be revised.

**Amendment 1 — the gradient is not "the trivial Bayes direction."** It is the contrast
model's prominence term, and prominence is *not* a base rate: net of a flexible spline in
prevalence it carries a large, universally-signed **negativity** component (undesirable
traits are stronger referents) and no lexical-frequency component. Double-centering still
removes it, correctly, for the similarity measurement — but it removes a **construct**
(the models' feature-salience ordering over the trait lexicon), not an artefact, and that
construct is worth reporting in its own right.

**Amendment 2 — the curl is not "faint and idiosyncratic."** At the resolution the theory
predicts it should live at, it is **low-rank and strongly shared**: dominant-plane
congruence 0.73 (null 0.05), construct-level agreement 0.72 (null 0.30). Report II's
r̄ = 0.34 measured cell-level noise, not the structure. The curl is a legitimate,
reproducible **diagnosticity** channel with an interpretable shape — an
evaluative × (misfortune vs malevolence) circulation, with negative affect as the
construct-level net source.

The *measurement* verdict is unchanged: symmetrize for the similarity. What changes is
the status of what we set aside. It is a second construct with its own reliability, its
own theory, and — for the readout account — its own model-level parameters (§7), and it
should be carried as such rather than filed as nuisance.
