---
title: "Gemma3 12B GFC-30 TIRT Analysis"
subtitle: "Persona recovery, trait structure, and neutral model placement"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(thurstonianIRT)
library(tidyverse)
library(jsonlite)

traits <- c("A", "C", "E", "N", "O")
trait_labels <- c(A = "Agreeableness", C = "Conscientiousness",
                  E = "Extraversion", N = "Neuroticism", O = "Openness")
```

# Introduction

This report analyzes Gemma3 12B's responses to the Okada et al. (2026) GFC-30
(Graded Forced-Choice Big Five) instrument, administered under 400 synthetic
personas with known ground-truth trait profiles. We fit a Thurstonian IRT (TIRT)
model using the `thurstonianIRT` R package (Stan backend, cumulative family) and
evaluate:

1. **Ground-truth recovery**: How well do TIRT trait estimates correlate with
   the z-scores used to generate each persona?
2. **Trait correlation structure**: Does TIRT recover the van der Linden (2010)
   inter-trait correlations used to simulate the personas?
3. **Response behavior**: How does Gemma3 use the 7-point bipolar scale?
4. **Neutral model placement**: Where does Gemma3's "default personality" fall
   on the persona-fitted trait scale, tested under two prompt conditions
   ("respondent" frame vs. "bare" honest instruction)?

# 1. TIRT Fit Summary

The TIRT model was fitted using `thurstonianIRT::fit_TIRT_stan()` with 2 chains
x 1000 iterations (500 warmup), cumulative family. The 30 GFC pairs produce 30
ordinal indicators mapped onto 5 latent traits via fixed loadings from the TIRT
design matrix.

```{r load-fit}
fit <- readRDS("gemma3_gfc30_synthetic_tirt_fit.rds")

# Load instrument metadata + build blocks (reused for neutral scoring)
inst <- fromJSON("../instruments/okada_gfc30.json")
pairs <- inst$pairs

block_df <- tibble(
  block = rep(seq_len(nrow(pairs)), each = 2),
  item = character(nrow(pairs) * 2),
  trait = character(nrow(pairs) * 2),
  sign = numeric(nrow(pairs) * 2)
)
for (i in seq_len(nrow(pairs))) {
  left_name <- paste0(pairs$left$trait[i],
                       ifelse(pairs$left$keying[i] == "+", "p", "n"), i)
  right_name <- paste0(pairs$right$trait[i],
                        ifelse(pairs$right$keying[i] == "+", "p", "n"), i)
  idx <- (i - 1) * 2
  block_df$item[idx + 1] <- left_name
  block_df$trait[idx + 1] <- pairs$left$trait[i]
  block_df$sign[idx + 1] <- ifelse(pairs$left$keying[i] == "+", 1, -1)
  block_df$item[idx + 2] <- right_name
  block_df$trait[idx + 2] <- pairs$right$trait[i]
  block_df$sign[idx + 2] <- ifelse(pairs$right$keying[i] == "+", 1, -1)
}
blocks <- set_blocks_from_df(block_df)

# Load ground truth
personas_json <- fromJSON("../instruments/synthetic_personas.json")
ground_truth <- tibble(
  persona_id = personas_json$personas$persona_id,
  A_true = personas_json$personas$z_scores$A,
  C_true = personas_json$personas$z_scores$C,
  E_true = personas_json$personas$z_scores$E,
  N_true = personas_json$personas$z_scores$N,
  O_true = personas_json$personas$z_scores$O
)

# Wide CSV for persona ID ordering
wide <- read_csv("gemma3-12b_gfc30_synthetic_wide.csv", show_col_types = FALSE)
complete_mask <- complete.cases(wide[, -1])
persona_ids <- wide$persona_id[complete_mask]
n_complete <- sum(complete_mask)

cat(sprintf("Complete respondents: %d / %d\n", n_complete, nrow(wide)))
```

```{r extract-scores}
scores_long <- as_tibble(predict(fit))

scores_wide <- scores_long %>%
  pivot_wider(names_from = trait,
              values_from = c(estimate, se, lower_ci, upper_ci),
              names_glue = "{trait}_{.value}") %>%
  mutate(persona_id = persona_ids) %>%
  left_join(ground_truth, by = "persona_id")
```

# 2. Ground-Truth Recovery

The central question: when we tell an LLM to respond as a person with known Big
Five trait levels (e.g., "very extraverted, moderately agreeable, low
neuroticism"), does the GFC+TIRT pipeline recover those trait levels?

We compare two scoring methods:

- **TIRT**: Thurstonian IRT trait estimates from the Stan model --- a latent
  variable approach that disentangles the forced-choice ipsativity.
- **Simple scoring**: Mean endorsement per trait across all pairs involving that
  trait. This is mechanically ipsative (trait scores sum to zero within each
  respondent), but serves as a baseline.

## Recovery Correlations

```{r recovery-table}
# Simple scoring
raw <- fromJSON("gemma3-12b_gfc30_synthetic.json", flatten = TRUE)
all_results <- as_tibble(raw$results)

trait_evidence <- all_results %>%
  filter(!is.na(response_argmax)) %>%
  mutate(
    resp_val = coalesce(as.numeric(response_ev), as.numeric(response_argmax)),
    pref = resp_val - 4,
    right_endorsement = pref * ifelse(right_keying == "+", 1, -1),
    left_endorsement = -pref * ifelse(left_keying == "+", 1, -1)
  )

simple_scores <- bind_rows(
  trait_evidence %>% select(persona_id, trait = right_trait,
                            endorsement = right_endorsement),
  trait_evidence %>% select(persona_id, trait = left_trait,
                            endorsement = left_endorsement)
) %>%
  group_by(persona_id, trait) %>%
  summarise(score = mean(endorsement, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = trait, values_from = score) %>%
  left_join(ground_truth, by = "persona_id")

recovery <- tibble(trait = traits) %>%
  rowwise() %>%
  mutate(
    tirt_r = cor(scores_wide[[paste0(trait, "_estimate")]],
                 scores_wide[[paste0(trait, "_true")]],
                 use = "complete.obs"),
    simple_r = cor(simple_scores[[trait]],
                   simple_scores[[paste0(trait, "_true")]],
                   use = "complete.obs"),
    trait_name = trait_labels[trait]
  ) %>%
  ungroup()

knitr::kable(recovery %>% select(Trait = trait_name, TIRT = tirt_r,
                                  Simple = simple_r),
             digits = 3, caption = "Ground-truth recovery (Pearson r)")
```

TIRT substantially outperforms simple scoring on E, N, and O --- the traits
where ipsative contamination is most damaging to simple scoring. However, A shows
a sign flip in TIRT (negative correlation), and C recovery is near zero. This is
a known issue with the `thurstonianIRT` Stan implementation on small-N data with
only 2 chains: the posterior may not mix well enough to resolve rotational
ambiguity in the latent space.

Simple scoring, despite its ipsativity constraint, provides a reasonable baseline
for individual trait recovery (r = 0.25--0.46).

## Scatterplots: TIRT Scores vs Ground Truth

```{r scatter-recovery, fig.height=8}
scatter_data <- scores_wide %>%
  pivot_longer(cols = matches("^[ACENO]_(estimate|true)$"),
               names_to = c("trait", ".value"),
               names_pattern = "^(.)_(.+)$")

ggplot(scatter_data, aes(x = true, y = estimate)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  geom_text(data = recovery, aes(label = sprintf("r = %.2f", tirt_r),
                                  x = -Inf, y = Inf),
            hjust = -0.1, vjust = 1.3, size = 4, inherit.aes = FALSE) +
  facet_wrap(~ trait, scales = "free", labeller = labeller(
    trait = trait_labels)) +
  labs(x = "Ground-truth z-score", y = "TIRT estimate",
       title = "TIRT Score Recovery: Gemma3 12B on GFC-30 (400 synthetic personas)") +
  theme_minimal()
```

The scatterplots reveal the heterogeneity in recovery quality. E, N, and O show
clear linear trends. A's sign flip is visible (negative slope), and C is
essentially flat --- TIRT failed to separate high-C from low-C personas.

## Simple Scoring Recovery

```{r scatter-simple, fig.height=8}
simple_long <- simple_scores %>%
  pivot_longer(cols = all_of(traits), names_to = "trait",
               values_to = "simple_est") %>%
  left_join(
    ground_truth %>%
      pivot_longer(-persona_id, names_to = "trait", values_to = "true",
                   names_pattern = "(.)_true"),
    by = c("persona_id", "trait")
  )

ggplot(simple_long, aes(x = true, y = simple_est)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "coral") +
  geom_text(data = recovery, aes(label = sprintf("r = %.2f", simple_r),
                                  x = -Inf, y = Inf),
            hjust = -0.1, vjust = 1.3, size = 4, inherit.aes = FALSE) +
  facet_wrap(~ trait, scales = "free", labeller = labeller(
    trait = trait_labels)) +
  labs(x = "Ground-truth z-score", y = "Simple score (mean endorsement)",
       title = "Simple Scoring Recovery: Gemma3 12B on GFC-30") +
  theme_minimal()
```

Simple scoring shows more uniform (if modest) recovery across all five traits.
The compressed vertical range reflects the ipsative constraint.

# 3. Estimated Trait Correlations

A key advantage of TIRT over simple scoring is its ability to recover the true
inter-trait correlation structure, free from the mechanical negative correlations
imposed by forced-choice ipsativity. The ground-truth correlations come from the
van der Linden (2010) meta-analytic matrix used to generate the personas.

```{r trait-cors}
tirt_cor <- scores_wide %>%
  select(all_of(paste0(traits, "_estimate"))) %>%
  cor(use = "complete.obs")
colnames(tirt_cor) <- traits
rownames(tirt_cor) <- traits

gt_cor <- ground_truth %>%
  select(all_of(paste0(traits, "_true"))) %>%
  cor()
colnames(gt_cor) <- traits
rownames(gt_cor) <- traits

simple_cor <- simple_scores %>%
  select(all_of(traits)) %>%
  cor(use = "complete.obs")
```

```{r cor-heatmaps, fig.height=5, fig.width=14}
make_cor_long <- function(mat, label) {
  as.data.frame(as.table(mat)) %>%
    rename(Var1 = 1, Var2 = 2, value = 3) %>%
    mutate(source = label)
}

cor_all <- bind_rows(
  make_cor_long(gt_cor, "Ground Truth"),
  make_cor_long(tirt_cor, "TIRT"),
  make_cor_long(simple_cor, "Simple")
) %>%
  mutate(source = factor(source, levels = c("Ground Truth", "TIRT", "Simple")))

ggplot(cor_all, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", value)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~ source) +
  labs(title = "Trait Correlations: Ground Truth vs TIRT vs Simple Scoring",
       x = NULL, y = NULL, fill = "r") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0))
```

The TIRT inter-trait correlations are implausibly high (e.g., C-E = 0.92, E-N =
-0.98). This is consistent with poor posterior mixing in the 2-chain run ---
the latent trait dimensions are not well-separated. Simple scoring shows the
expected ipsative pattern: pervasive negative correlations driven by the
forced-choice constraint.

Neither method recovers the moderate ground-truth correlations (|r| ~ 0.2--0.4).
This is a known challenge for TIRT with limited data and chains. A 4-chain refit
with LKJ priors on the correlation matrix would likely improve this.

# 4. Response Distribution

```{r response-dist, fig.height=5}
all_results %>%
  filter(!is.na(response_argmax)) %>%
  mutate(response = as.integer(response_argmax)) %>%
  count(response) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = factor(response), y = pct)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%", pct)), vjust = -0.3, size = 3) +
  labs(x = "Response (1=strongly left, 7=strongly right)",
       y = "Percent",
       title = "GFC-30 Response Distribution: Gemma3 12B (all 12,000 prompts)") +
  theme_minimal()
```

```{r response-by-trait, fig.height=6}
# Response distribution broken down by the trait pair
pair_traits <- all_results %>%
  filter(!is.na(response_argmax)) %>%
  mutate(response = as.integer(response_argmax),
         pair_label = paste0(left_trait, " vs ", right_trait))

pair_traits %>%
  count(left_trait, right_trait, response) %>%
  group_by(left_trait, right_trait) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  mutate(pair = paste0(left_trait, "-", right_trait)) %>%
  ggplot(aes(x = factor(response), y = pct)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  facet_wrap(~ pair, scales = "free_y") +
  labs(x = "Response", y = "Percent",
       title = "Response Distribution by Trait Pair",
       subtitle = "Lower = prefers left trait; higher = prefers right trait") +
  theme_minimal() +
  theme(strip.text = element_text(size = 8))
```

The overall distribution is strongly centered on 4 ("About the same"), which
is expected for GFC with desirability-matched pairs --- neither item is
obviously more "correct." The per-pair breakdown reveals which trait
comparisons produce more spread, offering a window into which trait contrasts
Gemma3 is most sensitive to under persona manipulation.

```{r entropy-by-persona, fig.height=5}
# Distribution of per-persona mean entropy
persona_entropy <- all_results %>%
  filter(!is.na(response_entropy)) %>%
  group_by(persona_id) %>%
  summarise(mean_entropy = mean(response_entropy, na.rm = TRUE),
            mean_ev = mean(as.numeric(response_ev), na.rm = TRUE),
            .groups = "drop")

ggplot(persona_entropy, aes(x = mean_entropy)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(persona_entropy$mean_entropy),
             color = "firebrick", linetype = "dashed") +
  labs(x = "Mean entropy (nats) across 30 items",
       y = "Count (personas)",
       title = "Response Certainty Across Personas",
       subtitle = sprintf("Mean = %.3f, max possible = %.3f (ln 7)",
                           mean(persona_entropy$mean_entropy), log(7))) +
  theme_minimal()
```

Gemma3 12B shows relatively low entropy (mean ~`r round(mean(persona_entropy$mean_entropy), 2)` nats out of a maximum of `r round(log(7), 2)`), indicating confident
responses. This peaked distribution is consistent with the model strongly
committing to particular response options rather than spreading probability
across the scale.

# 5. Neutral Model Placement

A key goal is to place Gemma3 12B's "default personality" on the trait scale
defined by the persona-fitted TIRT model. We administered GFC-30 to the model
without any persona description, under two conditions:

- **Respondent**: Includes "YOU ARE THE RESPONDENT" role assignment + honest
  instruction ("be as honest as possible, present yourself as you really are")
- **Bare**: Honest instruction only, no role assignment

Both conditions give 30 responses (one per GFC pair). We score them two ways:
simple scoring (for interpretability) and post-hoc TIRT scoring using
`predict(fit, newdata=...)`, which applies the persona-fitted item parameters to
estimate where the neutral respondent falls on each latent trait dimension.

## Simple Scoring

```{r neutral-simple}
score_neutral_simple <- function(json_path, label) {
  raw <- fromJSON(json_path, flatten = TRUE)
  results <- as_tibble(raw$results)

  results %>%
    filter(!is.na(response_argmax)) %>%
    mutate(
      resp_val = coalesce(as.numeric(response_ev), as.numeric(response_argmax)),
      pref = resp_val - 4,
      right_endorsement = pref * ifelse(right_keying == "+", 1, -1),
      left_endorsement = -pref * ifelse(left_keying == "+", 1, -1)
    ) %>%
    {bind_rows(
      select(., trait = right_trait, endorsement = right_endorsement),
      select(., trait = left_trait, endorsement = left_endorsement)
    )} %>%
    group_by(trait) %>%
    summarise(score = mean(endorsement, na.rm = TRUE), .groups = "drop") %>%
    mutate(mode = label)
}

neutral_simple_respondent <- score_neutral_simple(
  "gemma3-12b_gfc30_neutral-respondent.json", "respondent")
neutral_simple_bare <- score_neutral_simple(
  "gemma3-12b_gfc30_neutral-bare.json", "bare")
neutral_simple <- bind_rows(neutral_simple_respondent, neutral_simple_bare)

knitr::kable(neutral_simple %>% pivot_wider(names_from = mode,
                                             values_from = score),
             digits = 3,
             caption = "Gemma3 12B neutral simple scores by prompt mode")
```

## TIRT Scoring (Post-Hoc)

The post-hoc TIRT scoring uses `predict(fit, newdata=...)`, which takes the
posterior median item parameters from the persona-fitted model and estimates
trait scores for new response patterns. This places the neutral respondent on
the same latent scale as the 400 personas, avoiding the ipsativity problem of
simple scoring.

```{r neutral-tirt}
build_neutral_pairwise <- function(json_path) {
  raw <- fromJSON(json_path, flatten = TRUE)
  results <- as_tibble(raw$results)

  pw <- data.frame(person = 1L)
  for (i in seq_len(nrow(pairs))) {
    left_name <- paste0(pairs$left$trait[i],
                         ifelse(pairs$left$keying[i] == "+", "p", "n"), i)
    right_name <- paste0(pairs$right$trait[i],
                          ifelse(pairs$right$keying[i] == "+", "p", "n"), i)
    col_name <- paste0(left_name, right_name)
    block_result <- results %>% filter(block == i)
    pw[[col_name]] <- as.integer(block_result$response_argmax[1])
  }
  pw
}

pw_respondent <- build_neutral_pairwise("gemma3-12b_gfc30_neutral-respondent.json")
pw_bare <- build_neutral_pairwise("gemma3-12b_gfc30_neutral-bare.json")
pw_neutral <- bind_rows(pw_respondent, pw_bare)
pw_neutral$person <- 1:2

tirt_neutral <- make_TIRT_data(
  data = pw_neutral,
  blocks = blocks,
  direction = "larger",
  format = "pairwise",
  family = "cumulative",
  range = c(1, 7)
)

neutral_tirt_scores <- as_tibble(predict(fit, newdata = tirt_neutral))

neutral_tirt_wide <- neutral_tirt_scores %>%
  mutate(mode = rep(c("respondent", "bare"), each = length(traits))) %>%
  select(mode, trait, estimate, se)

knitr::kable(neutral_tirt_wide %>%
               pivot_wider(names_from = mode,
                           values_from = c(estimate, se),
                           names_glue = "{mode}_{.value}"),
             digits = 3,
             caption = "Gemma3 12B neutral TIRT scores (post-hoc, posterior median item params)")
```

## Neutral Placement vs Persona Distribution

The violin plots below show the distribution of persona TIRT scores grouped by
ground-truth tercile (low/medium/high on each trait). The horizontal lines show
where Gemma3's neutral responses fall. If the model has a "default personality,"
these lines should not be centered --- they should be displaced toward whichever
end of the trait the model's RLHF training favors.

```{r persona-terciles}
persona_tirt <- scores_long %>%
  mutate(persona_id = rep(persona_ids, each = length(traits)))

gt_long <- ground_truth %>%
  pivot_longer(-persona_id, names_to = "trait", values_to = "z_true",
               names_pattern = "(.)_true") %>%
  group_by(trait) %>%
  mutate(tercile = case_when(
    z_true <= quantile(z_true, 1/3) ~ "Low",
    z_true >= quantile(z_true, 2/3) ~ "High",
    TRUE ~ "Medium"
  )) %>%
  ungroup()

persona_scored <- persona_tirt %>%
  left_join(gt_long, by = c("persona_id", "trait"))
```

```{r placement-plot-tirt, fig.height=7, fig.width=10}
neutral_plot_data <- neutral_tirt_wide %>%
  select(mode, trait, estimate)

ggplot() +
  geom_violin(data = persona_scored %>%
                mutate(tercile = factor(tercile,
                                        levels = c("Low", "Medium", "High"))),
              aes(x = tercile, y = estimate, fill = tercile),
              alpha = 0.5, scale = "width") +
  geom_boxplot(data = persona_scored %>%
                 mutate(tercile = factor(tercile,
                                          levels = c("Low", "Medium", "High"))),
               aes(x = tercile, y = estimate),
               width = 0.15, outlier.size = 0.5) +
  geom_hline(data = neutral_plot_data,
             aes(yintercept = estimate, linetype = mode, color = mode),
             linewidth = 1) +
  scale_fill_manual(values = c(Low = "#4575b4", Medium = "#ffffbf",
                                High = "#d73027"),
                    name = "Persona\ntercile") +
  scale_color_manual(values = c(respondent = "black", bare = "darkgreen"),
                     name = "Neutral\nmode") +
  scale_linetype_manual(values = c(respondent = "solid", bare = "dashed"),
                        name = "Neutral\nmode") +
  facet_wrap(~ trait, scales = "free_y", labeller = labeller(
    trait = trait_labels)) +
  labs(y = "TIRT trait estimate",
       x = "Ground-truth tercile",
       title = "Gemma3 12B: Neutral TIRT Placement vs Persona Distributions",
       subtitle = "Lines = model default (respondent/bare); violins = persona TIRT scores by ground-truth tercile") +
  theme_minimal() +
  theme(legend.position = "right")
```

### Density overlay: neutral vs persona distribution

An alternative view: where does the neutral point fall in the overall persona
TIRT score distribution for each trait?

```{r density-placement, fig.height=8}
ggplot(persona_tirt, aes(x = estimate)) +
  geom_density(fill = "grey70", alpha = 0.5) +
  geom_vline(data = neutral_plot_data %>% filter(mode == "respondent"),
             aes(xintercept = estimate), color = "black",
             linetype = "solid", linewidth = 1) +
  geom_vline(data = neutral_plot_data %>% filter(mode == "bare"),
             aes(xintercept = estimate), color = "darkgreen",
             linetype = "dashed", linewidth = 1) +
  facet_wrap(~ trait, scales = "free", ncol = 1, labeller = labeller(
    trait = trait_labels)) +
  labs(x = "TIRT trait estimate",
       y = "Density",
       title = "Neutral Placement on Persona TIRT Score Distributions",
       subtitle = "Black solid = respondent frame; green dashed = bare frame") +
  theme_minimal()
```

### Neutral as percentile of persona distribution

```{r neutral-percentile}
percentile_df <- neutral_plot_data %>%
  group_by(mode) %>%
  group_modify(~ {
    .x %>%
      rowwise() %>%
      mutate(
        percentile = mean(
          persona_tirt$estimate[persona_tirt$trait == trait] <= estimate
        ) * 100
      ) %>%
      ungroup()
  })

knitr::kable(percentile_df %>%
               select(mode, trait, estimate, percentile) %>%
               pivot_wider(names_from = mode,
                           values_from = c(estimate, percentile),
                           names_glue = "{mode}_{.value}"),
             digits = 2,
             caption = "Neutral placement as percentile of persona TIRT distribution")
```

A percentile near 50 means the model's default is indistinguishable from the
average synthetic persona. Percentiles far from 50 reveal the model's
"personality" --- traits where RLHF training has pushed the model toward one
pole.

## Per-Item Comparison: Respondent vs Bare

The two neutral conditions differ only in whether "YOU ARE THE RESPONDENT" is
prepended. If this framing activates a different response mode, we should see
systematic item-level differences.

```{r item-comparison, fig.height=5}
resp_items <- fromJSON("gemma3-12b_gfc30_neutral-respondent.json",
                       flatten = TRUE)$results %>%
  as_tibble() %>%
  mutate(mode = "respondent",
         ev = as.numeric(response_ev),
         block = as.integer(block))

bare_items <- fromJSON("gemma3-12b_gfc30_neutral-bare.json",
                       flatten = TRUE)$results %>%
  as_tibble() %>%
  mutate(mode = "bare",
         ev = as.numeric(response_ev),
         block = as.integer(block))

item_compare <- bind_rows(resp_items, bare_items) %>%
  select(block, left_trait, right_trait, mode, ev) %>%
  pivot_wider(names_from = mode, values_from = ev)

ggplot(item_compare, aes(x = respondent, y = bare)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_text(aes(label = block), nudge_y = 0.1, size = 2.5) +
  labs(x = "EV (respondent frame)", y = "EV (bare frame)",
       title = "Per-Item EV: Respondent vs Bare Neutral (Gemma3 12B)",
       subtitle = "Points on diagonal = identical responses; labeled by block number") +
  coord_equal(xlim = c(1, 7), ylim = c(1, 7)) +
  theme_minimal()

cat(sprintf("Correlation between modes: r = %.3f\n",
            cor(item_compare$respondent, item_compare$bare, use = "complete.obs")))
cat(sprintf("Mean absolute difference: %.3f\n",
            mean(abs(item_compare$respondent - item_compare$bare), na.rm = TRUE)))
```

```{r item-diff, fig.height=5}
item_compare %>%
  mutate(diff = bare - respondent,
         pair_label = paste0("B", block, ": ", left_trait, "-", right_trait)) %>%
  ggplot(aes(x = reorder(pair_label, diff), y = diff)) +
  geom_col(aes(fill = diff > 0), alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "darkgreen", "FALSE" = "black")) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = NULL, y = "EV difference (bare - respondent)",
       title = "Item-Level Differences Between Neutral Modes",
       subtitle = "Positive = bare gives higher EV (more toward right statement)") +
  theme_minimal()
```

# 6. Summary

```{r summary-table}
cat("=== Gemma3 12B GFC-30 TIRT Summary ===\n\n")
cat(sprintf("Personas: %d complete / %d total\n", n_complete, nrow(wide)))
cat(sprintf("Prompts: %d\n", nrow(all_results)))
cat(sprintf("Valid responses: %d (%.1f%%)\n",
            sum(!is.na(all_results$response_argmax)),
            100 * mean(!is.na(all_results$response_argmax))))

cat("\nGround-truth recovery (Pearson r):\n")
for (i in seq_along(traits)) {
  cat(sprintf("  %s: TIRT=%.3f, Simple=%.3f\n",
              trait_labels[traits[i]],
              recovery$tirt_r[i], recovery$simple_r[i]))
}

cat("\nNeutral TIRT placement:\n")
for (m in c("respondent", "bare")) {
  scores_m <- neutral_tirt_wide %>% filter(mode == m)
  cat(sprintf("  %s: %s\n", m,
              paste(scores_m %>%
                      mutate(s = sprintf("%s=%.2f (SE=%.2f)", trait, estimate, se)) %>%
                      pull(s), collapse = ", ")))
}
```

## Key Findings

1. **GFC breaks the Likert ceiling effect**: Gemma3 uses the full 7-point scale
   (centered on 4), unlike the Likert "assistant shape" where responses cluster
   at 5/5.

2. **TIRT improves over simple scoring for E, N, O** (r = 0.53--0.68 vs
   0.35--0.44), but struggles with A (sign flip) and C (near-zero recovery).
   The 2-chain Stan run likely has insufficient posterior exploration.

3. **Inter-trait correlations are not recovered** by either method: TIRT
   produces implausibly high correlations, simple scoring produces ipsative
   artifacts. A 4-chain refit with informative priors is the next step.

4. **Neutral model placement** uses post-hoc TIRT scoring against persona-fitted
   item parameters, providing a principled trait-scale placement of the model's
   default response pattern.

## Next Steps

- **Refit with 4 chains** and LKJ prior on trait correlations for better
  posterior exploration.
- **Multi-model comparison**: Phi4-mini TIRT fit is in progress. Placing
  multiple models on the same trait scale enables cross-model personality
  comparison.
- **Measurement invariance**: Test whether item parameters hold across models
  (configural -> metric -> scalar invariance).
