---
title: "EDA: Logprob Distributions from Llama-2 BFI Pilot"
author: "Ethan Brown"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(tidyverse)
library(patchwork)
```

## Load Data

```{r load}
df <- read_csv("results_ollama-run_llama2-7b-chat_0.csv")

# Impute NA logprobs with floor value (min observed across all logprob columns).
# NAs occur when an option's probability was too low to appear in top_logprobs.
logprob_floor <- min(df %>% select(starts_with("logprob_")), na.rm = TRUE)
n_imputed <- sum(is.na(df %>% select(starts_with("logprob_"))))
cat(sprintf("Imputing %d NA logprobs with floor value %.2f\n",
            n_imputed, logprob_floor))

df <- df %>%
  mutate(across(starts_with("logprob_"), ~ replace_na(.x, logprob_floor)))

# Convert logprobs to probabilities
df <- df %>%
  mutate(across(starts_with("logprob_"), exp, .names = "prob_{col}")) %>%
  rename_with(~ str_replace(., "prob_logprob_", "prob_"), starts_with("prob_"))

glimpse(df)
```

## Overview

```{r overview}
# Response distribution
ggplot(df, aes(x = model_output)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Model Chosen Responses",
       x = "Likert Response", y = "Count") +
  theme_minimal()
```

```{r valid_mass}
ggplot(df, aes(x = valid_prob_mass)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  geom_vline(xintercept = mean(df$valid_prob_mass, na.rm = TRUE),
             linetype = "dashed", color = "red") +
  labs(title = "Valid Probability Mass (proportion on Likert options vs. nonsense)",
       subtitle = sprintf("Mean = %.3f", mean(df$valid_prob_mass, na.rm = TRUE)),
       x = "Valid Prob Mass", y = "Count") +
  theme_minimal()
```

## Probability Distributions Across Options

How does the model distribute probability across Likert options?

```{r prob_distributions}
# Reshape to long format
probs_long <- df %>%
  select(starts_with("prob_"), model_output, item_id, item_preamble_id,
         response_scale_id, valid_prob_mass) %>%
  pivot_longer(cols = starts_with("prob_"),
               names_to = "option",
               values_to = "prob") %>%
  mutate(option = as.integer(str_extract(option, "\\d+")))

# Distribution of probability by option, faceted by chosen response
ggplot(probs_long, aes(x = factor(option), y = prob)) +
  geom_boxplot(aes(fill = factor(option)), outlier.size = 0.5) +
  facet_wrap(~ paste("Chose:", model_output), nrow = 1) +
  labs(title = "Probability by Option, Grouped by Model's Chosen Response",
       subtitle = "GRM-friendly: probability should peak at chosen and decay ordinally",
       x = "Likert Option", y = "Probability") +
  scale_fill_brewer(palette = "RdYlBu", direction = -1, guide = "none") +
  theme_minimal()
```

## Ordinality Analysis

For a graded response model, we'd expect the probability distribution to be
roughly unimodal around the chosen response, with probability decreasing
as we move away. Let's test this.

```{r ordinality_metrics}
# For each response, compute whether probabilities are unimodal
# and whether the mode matches the chosen response
ordinality <- df %>%
  rowwise() %>%
  mutate(
    # Probabilities as a vector
    probs = list(c(prob_1, prob_2, prob_3, prob_4, prob_5)),
    # Mode of the probability distribution
    prob_mode = which.max(c(prob_1, prob_2, prob_3, prob_4, prob_5)),
    # Does the logprob mode match the chosen response?
    mode_matches_choice = prob_mode == as.integer(model_output),
    # Distance from mode to choice
    mode_choice_distance = abs(prob_mode - as.integer(model_output)),
    # Is the distribution unimodal? Check that probs decrease
    # monotonically on each side of the mode
    left_monotone = {
      p <- c(prob_1, prob_2, prob_3, prob_4, prob_5)
      m <- prob_mode
      if (any(is.na(p))) NA
      else if (m == 1) TRUE
      else all(diff(p[1:m]) >= 0)
    },
    right_monotone = {
      p <- c(prob_1, prob_2, prob_3, prob_4, prob_5)
      m <- prob_mode
      if (any(is.na(p))) NA
      else if (m == 5) TRUE
      else all(diff(p[m:5]) <= 0)
    },
    unimodal = left_monotone & right_monotone
  ) %>%
  ungroup()

cat(sprintf("Mode matches chosen response: %d/%d (%.1f%%)\n",
            sum(ordinality$mode_matches_choice),
            nrow(ordinality),
            100 * mean(ordinality$mode_matches_choice, na.rm = TRUE)))
cat(sprintf("Unimodal distributions: %d/%d (%.1f%%)\n",
            sum(ordinality$unimodal),
            nrow(ordinality),
            100 * mean(ordinality$unimodal, na.rm = TRUE)))
```

```{r ordinality_viz}
ggplot(ordinality, aes(x = factor(prob_mode), fill = mode_matches_choice)) +
  geom_bar(position = "stack") +
  labs(title = "Logprob Mode vs. Chosen Response",
       x = "Logprob Mode (highest-probability option)",
       fill = "Matches Choice") +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  theme_minimal()
```

```{r unimodality_by_choice}
ggplot(ordinality, aes(x = factor(model_output), fill = unimodal)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Unimodal Distributions by Chosen Response",
       subtitle = "Unimodal = probs decrease monotonically from mode on both sides",
       x = "Chosen Response", y = "Proportion",
       fill = "Unimodal") +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()
```

## Adjacent Category Ratios (GRM Diagnostic)

In a graded response model, the log-odds of choosing category $k$ vs. $k-1$
should be a linear function of the latent trait. Let's look at adjacent
log-odds ratios to see if they behave orderly.

```{r adjacent_logodds}
adj_logodds <- df %>%
  mutate(
    # Adjacent log-odds ratios (log p_k - log p_{k-1})
    logodds_2v1 = logprob_2 - logprob_1,
    logodds_3v2 = logprob_3 - logprob_2,
    logodds_4v3 = logprob_4 - logprob_3,
    logodds_5v4 = logprob_5 - logprob_4
  )

adj_long <- adj_logodds %>%
  select(starts_with("logodds_"), model_output, item_id) %>%
  pivot_longer(cols = starts_with("logodds_"),
               names_to = "boundary",
               values_to = "log_odds_ratio") %>%
  mutate(boundary = factor(boundary,
                           levels = c("logodds_2v1", "logodds_3v2",
                                      "logodds_4v3", "logodds_5v4"),
                           labels = c("2 vs 1", "3 vs 2", "4 vs 3", "5 vs 4")))

ggplot(adj_long, aes(x = boundary, y = log_odds_ratio)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(title = "Adjacent Log-Odds Ratios Across Category Boundaries",
       subtitle = "GRM expects ordered thresholds; large positive = higher categories favored",
       x = "Category Boundary", y = "Log-Odds Ratio (log p_k - log p_{k-1})") +
  theme_minimal()
```

```{r adjacent_by_choice}
ggplot(adj_long, aes(x = boundary, y = log_odds_ratio,
                     fill = factor(model_output))) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_wrap(~ paste("Chose:", model_output), nrow = 1) +
  labs(title = "Adjacent Log-Odds Ratios by Chosen Response",
       x = "Category Boundary", y = "Log-Odds Ratio") +
  scale_fill_brewer(palette = "RdYlBu", direction = -1, guide = "none") +
  theme_minimal()
```

## Correlations Between Option Logprobs

If the response process is ordinal, we'd expect strong positive correlations
between adjacent options' logprobs, and negative correlations between
distant options.

```{r logprob_cors}
cor_mat <- df %>%
  select(logprob_1:logprob_5) %>%
  cor(use = "pairwise.complete.obs")

# Tidy for plotting
cor_long <- cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("opt_a") %>%
  pivot_longer(-opt_a, names_to = "opt_b", values_to = "r") %>%
  mutate(across(c(opt_a, opt_b), ~ str_extract(., "\\d+")))

ggplot(cor_long, aes(x = opt_a, y = opt_b, fill = r)) +
  geom_tile() +
  geom_text(aes(label = round(r, 2)), size = 4) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "tomato",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Matrix of Option Log-Probabilities",
       subtitle = "Ordinal structure: adjacent options should correlate positively",
       x = "Option", y = "Option", fill = "r") +
  coord_fixed() +
  theme_minimal()
```

## Mean Probability Profiles by Item

Do different BFI items elicit different response profiles?

```{r item_profiles, fig.height=8}
item_probs <- probs_long %>%
  group_by(item_id, option) %>%
  summarise(mean_prob = mean(prob, na.rm = TRUE), .groups = "drop")

ggplot(item_probs, aes(x = option, y = mean_prob, group = item_id)) +
  geom_line(alpha = 0.4, color = "steelblue") +
  geom_point(alpha = 0.4, color = "steelblue", size = 1) +
  labs(title = "Mean Probability Profile by Item (across preambles)",
       subtitle = "Each line = one BFI item; ordinal items should show peaked curves",
       x = "Likert Option", y = "Mean Probability") +
  theme_minimal()
```

## Summary: GRM vs. Nominal

```{r grm_summary}
# Proportion of responses where adjacent logprobs are ordered
# (i.e., categories below the mode have decreasing probs,
#  categories above the mode have decreasing probs)
cat("=== GRM Suitability Summary ===\n\n")
cat(sprintf("Mode matches chosen response: %.1f%%\n",
            100 * mean(ordinality$mode_matches_choice, na.rm = TRUE)))
cat(sprintf("Unimodal logprob distributions: %.1f%%\n",
            100 * mean(ordinality$unimodal, na.rm = TRUE)))
cat(sprintf("Mean valid prob mass: %.3f\n",
            mean(df$valid_prob_mass, na.rm = TRUE)))

# Check if adjacent log-odds are ordered across boundaries
threshold_order <- adj_logodds %>%
  mutate(ordered = logodds_2v1 < logodds_3v2 &
                   logodds_3v2 < logodds_4v3 &
                   logodds_4v3 < logodds_5v4)
cat(sprintf("Monotonically ordered thresholds: %.1f%%\n",
            100 * mean(threshold_order$ordered, na.rm = TRUE)))

cat("\n")
cat("High unimodality + ordered thresholds -> GRM appropriate.\n")
cat("Low unimodality or disordered thresholds -> consider nominal model.\n")
```
