---
title: "IRT Analysis: GRM & NRM on Full BFI Run (Llama-2 7B Chat)"
author: "Ethan Brown"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6
)
library(mirt, lib.loc = "/home/ethanbrown/R/library")
library(tidyverse)
```

## Overview

We fit unidimensional **Graded Response Models (GRM)** and **Nominal Response Models (NRM)**
to each Big Five trait from the full BFI run on Llama-2 7B Chat (55,000 prompts:
44 items x 250 preambles x 5 postambles = 1,250 SPIDs).

Each model is fit in two ways:

1. **Top-1 (standard)**: Using the logprob argmax as the categorical response
2. **Soft-evidence**: Using the full logprob distribution via our modified mirt fork

The soft-evidence approach replaces the one-hot categorical likelihood with a
weighted sum: $\ell_{ij} = \sum_k \pi_{ijk} \cdot \log P(Y=k \mid \theta_i, a_j, b_j)$.

## 1. Data Preparation

```{r load}
df <- read_csv("results_bfi-full_llama2-7b-chat_0.csv", show_col_types = FALSE)
```

```{r scale_defs}
bfi_scales <- list(
  `BFI-EXT` = list(
    items = c("bf1","bf6","bf11","bf16","bf21","bf26","bf31","bf36"),
    reverse = c("bf6","bf21","bf31")
  ),
  `BFI-AGR` = list(
    items = c("bf2","bf7","bf12","bf17","bf22","bf27","bf32","bf37","bf42"),
    reverse = c("bf2","bf12","bf27","bf37")
  ),
  `BFI-CON` = list(
    items = c("bf3","bf8","bf13","bf18","bf23","bf28","bf33","bf38","bf43"),
    reverse = c("bf8","bf18","bf23","bf43")
  ),
  `BFI-NEU` = list(
    items = c("bf4","bf9","bf14","bf19","bf24","bf29","bf34","bf39"),
    reverse = c("bf9","bf24","bf34")
  ),
  `BFI-OPE` = list(
    items = c("bf5","bf10","bf15","bf20","bf25","bf30","bf35","bf40","bf41","bf44"),
    reverse = c("bf35","bf41")
  )
)
```

### Response distribution

The logprob argmax only ever produces categories **1, 3, and 5** — never 2 or 4.
This extreme category suppression means mirt remaps our 5-point scale to 3 effective categories.

```{r response_dist, fig.height=4}
df <- df %>% mutate(response = as.integer(logprob_argmax))

df %>%
  count(response) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = factor(response), y = pct)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = sprintf("%.1f%%", pct * 100)), vjust = -0.3, size = 3.5) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Distribution of Logprob Argmax Responses",
       subtitle = "Categories 2 and 4 never appear — strong ceiling effect on category 5",
       x = "Logprob Argmax Response", y = "Proportion") +
  theme_minimal()
```

### Building the person-by-item matrix

```{r build_matrices}
resp_wide <- df %>%
  select(spid, item_id, response) %>%
  pivot_wider(names_from = item_id, values_from = response) %>%
  arrange(spid)

spid_ids <- resp_wide$spid
N <- nrow(resp_wide)
all_items <- sort(unique(df$item_id))
J <- length(all_items)
K_raw <- 5L
```

**`r N` SPIDs** (persons) x **`r J` items**, fully crossed (no missingness).

### Building the soft-evidence array

We convert 5-category logprob distributions to probabilities and collapse to 3 categories
matching the observed data: $p(\text{low}) = p_1 + p_2$, $p(\text{mid}) = p_3$,
$p(\text{high}) = p_4 + p_5$.

```{r soft_evidence_array}
lp_data <- df %>%
  select(spid, item_id, logprob_1, logprob_2, logprob_3, logprob_4, logprob_5) %>%
  arrange(spid, item_id)

lp_mat <- lp_data %>% select(starts_with("logprob_")) %>% as.matrix()

# Impute NAs with floor
lp_floor <- min(lp_mat, na.rm = TRUE)
n_na <- sum(is.na(lp_mat))
lp_mat[is.na(lp_mat)] <- lp_floor

# Exponentiate and renormalize
prob_mat <- exp(lp_mat)
prob_mat <- prob_mat / rowSums(prob_mat)

# Reshape to N x J x 5
pi_array_5 <- array(prob_mat, dim = c(J, N, K_raw))
pi_array_5 <- aperm(pi_array_5, c(2, 1, 3))

# Collapse to 3 categories
K <- 3L
pi_array <- array(0, dim = c(N, J, K))
pi_array[, , 1] <- pi_array_5[, , 1] + pi_array_5[, , 2]  # low
pi_array[, , 2] <- pi_array_5[, , 3]                        # mid
pi_array[, , 3] <- pi_array_5[, , 4] + pi_array_5[, , 5]  # high
```

```{r soft_evidence_summary, fig.height=4}
# Summarize the collapsed distributions
cat_probs <- tibble(
  Category = c("Low (1+2)", "Mid (3)", "High (4+5)"),
  Mean = c(mean(pi_array[,,1]), mean(pi_array[,,2]), mean(pi_array[,,3])),
  SD   = c(sd(pi_array[,,1]),   sd(pi_array[,,2]),   sd(pi_array[,,3]))
) %>%
  mutate(Category = fct_inorder(Category))

cat_probs %>%
  ggplot(aes(x = Category, y = Mean)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = Mean - SD, ymax = pmin(Mean + SD, 1)),
                width = 0.3) +
  geom_text(aes(label = sprintf("%.1f%% (sd=%.1f%%)", Mean*100, SD*100)),
            vjust = -0.5, size = 3.5) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(title = "Mean Soft-Evidence Probabilities (3-category collapsed)",
       subtitle = sprintf("Imputed %d NA logprobs with floor value %.2f", n_na, lp_floor),
       x = NULL, y = "Mean Probability") +
  theme_minimal()
```

The soft evidence is heavily dominated by the "high" category (~80%), with limited
between-person variation. This foreshadows the discrimination collapse we observe below.


## 2. Helper Functions

```{r helpers}
# Build softmat matching mirt's internal itemloc structure
build_softmat_from_model <- function(mod, pi_sub, items, dat_top1) {
  K_vec <- extract.mirt(mod, "K")
  itemloc <- extract.mirt(mod, "itemloc")
  ncols <- sum(K_vec)
  J_sub <- length(items)
  N_local <- dim(pi_sub)[1]
  softmat <- matrix(0, nrow = N_local, ncol = ncols)

  for (jj in 1:J_sub) {
    cols <- itemloc[jj]:(itemloc[jj + 1] - 1)
    if (K_vec[jj] == 3) {
      softmat[, cols] <- pi_sub[, jj, ]
    } else if (K_vec[jj] == 2) {
      # Collapse: low+mid -> cat1, high -> cat2
      softmat[, cols[1]] <- pi_sub[, jj, 1] + pi_sub[, jj, 2]
      softmat[, cols[2]] <- pi_sub[, jj, 3]
    }
  }
  softmat
}

# Extract fit stats into a tidy row
extract_fit <- function(mod, label) {
  if (is.null(mod)) return(NULL)
  tibble(
    Model = label,
    logLik = mod@Fit$logLik,
    AIC = mod@Fit$AIC,
    BIC = mod@Fit$BIC,
    npar = mod@Fit$df - mod@Fit$df  # placeholder
  )
}
```


## 3. Model Fitting

We loop over each Big Five trait, fitting four models:

- **GRM (top-1)**: Standard graded response model on categorical argmax responses
- **GRM (soft)**: Soft-evidence graded response model using logprob distributions
- **NRM (top-1)**: Standard nominal response model on categorical responses
- **NRM (soft)**: Soft-evidence nominal response model

```{r fit_models, results='hide'}
results <- list()

for (scale_id in names(bfi_scales)) {
  items <- bfi_scales[[scale_id]]$items
  rev_items <- bfi_scales[[scale_id]]$reverse
  J_sub <- length(items)
  item_idx <- match(items, all_items)

  # Top-1 response matrix
  dat_top1 <- resp_wide %>% select(all_of(items)) %>% as.data.frame()
  for (ri in rev_items) dat_top1[[ri]] <- 6L - dat_top1[[ri]]

  # Soft-evidence sub-array (reverse-keyed)
  pi_sub <- pi_array[, item_idx, , drop = FALSE]
  for (ri in rev_items) {
    ri_idx <- which(items == ri)
    pi_sub[, ri_idx, ] <- pi_sub[, ri_idx, K:1]
  }

  # 1. GRM top-1
  mod_grm_top1 <- tryCatch(
    mirt(dat_top1, 1, itemtype = "graded", verbose = FALSE, TOL = 1e-4),
    error = function(e) NULL
  )

  # Build softmat from GRM's K structure
  softmat <- if (!is.null(mod_grm_top1)) {
    build_softmat_from_model(mod_grm_top1, pi_sub, items, dat_top1)
  }

  # 2. GRM soft
  mod_grm_soft <- if (!is.null(softmat)) {
    tryCatch(
      mirt(dat_top1, 1, itemtype = "graded", verbose = FALSE, TOL = 1e-4,
           technical = list(softevidence = softmat)),
      error = function(e) NULL
    )
  }

  # 3. NRM top-1
  mod_nrm_top1 <- tryCatch(
    mirt(dat_top1, 1, itemtype = "nominal", verbose = FALSE, TOL = 1e-4),
    error = function(e) NULL
  )

  # Build softmat for NRM (may differ in K)
  softmat_nrm <- if (!is.null(mod_nrm_top1)) {
    build_softmat_from_model(mod_nrm_top1, pi_sub, items, dat_top1)
  }

  # 4. NRM soft
  mod_nrm_soft <- if (!is.null(softmat_nrm)) {
    tryCatch(
      mirt(dat_top1, 1, itemtype = "nominal", verbose = FALSE, TOL = 1e-4,
           technical = list(softevidence = softmat_nrm)),
      error = function(e) NULL
    )
  }

  results[[scale_id]] <- list(
    grm_top1 = mod_grm_top1, grm_soft = mod_grm_soft,
    nrm_top1 = mod_nrm_top1, nrm_soft = mod_nrm_soft,
    items = items, reverse = rev_items,
    dat_top1 = dat_top1, softmat = softmat, pi_sub = pi_sub
  )
}
```

```{r save_rds, include=FALSE}
saveRDS(results, "fit_irt_bfi_full_results.rds")
```


## 4. Results: Top-1 (Standard IRT)

### Fit comparison: GRM vs NRM

```{r top1_summary}
top1_summary <- names(results) %>%
  map_dfr(function(s) {
    r <- results[[s]]
    tibble(
      Scale = s,
      J = length(r$items),
      GRM_logLik = if (!is.null(r$grm_top1)) r$grm_top1@Fit$logLik else NA_real_,
      GRM_AIC = if (!is.null(r$grm_top1)) r$grm_top1@Fit$AIC else NA_real_,
      GRM_BIC = if (!is.null(r$grm_top1)) r$grm_top1@Fit$BIC else NA_real_,
      NRM_logLik = if (!is.null(r$nrm_top1)) r$nrm_top1@Fit$logLik else NA_real_,
      NRM_AIC = if (!is.null(r$nrm_top1)) r$nrm_top1@Fit$AIC else NA_real_,
      NRM_BIC = if (!is.null(r$nrm_top1)) r$nrm_top1@Fit$BIC else NA_real_
    ) %>%
      mutate(
        Delta_AIC = GRM_AIC - NRM_AIC,
        Preferred_AIC = ifelse(Delta_AIC < 0, "GRM", "NRM"),
        Preferred_BIC = ifelse(GRM_BIC < NRM_BIC, "GRM", "NRM")
      )
  })

top1_summary %>%
  select(Scale, J, GRM_AIC, NRM_AIC, Delta_AIC, Preferred_AIC, Preferred_BIC) %>%
  knitr::kable(digits = 1, caption = "Top-1 model comparison: GRM vs NRM per trait")
```

```{r top1_aic_plot, fig.height=4}
top1_summary %>%
  select(Scale, GRM = GRM_AIC, NRM = NRM_AIC) %>%
  pivot_longer(-Scale, names_to = "Model", values_to = "AIC") %>%
  ggplot(aes(x = Scale, y = AIC, fill = Model)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(GRM = "steelblue", NRM = "coral")) +
  labs(title = "AIC Comparison: GRM vs NRM (Top-1 Responses)",
       subtitle = "Lower is better. GRM preferred for CON, NEU, OPE; NRM for EXT, AGR",
       y = "AIC") +
  theme_minimal()
```

**GRM is preferred for 3 of 5 scales** (CON, NEU, OPE by both AIC and BIC).
NRM is preferred for EXT (strongly) and AGR (AIC only; BIC favors GRM).
The NRM preference for EXT/AGR may reflect non-ordinal response patterns for those traits
under Llama-2's persona prompting.


### GRM discrimination parameters

```{r grm_top1_params}
grm_params <- names(results) %>%
  map_dfr(function(s) {
    r <- results[[s]]
    if (is.null(r$grm_top1)) return(NULL)
    coefs <- coef(r$grm_top1, simplify = TRUE)$items
    as_tibble(coefs, rownames = "item") %>%
      mutate(Scale = s, .before = 1)
  })

grm_params %>%
  select(Scale, item, a1, d1, d2) %>%
  knitr::kable(digits = 2, caption = "GRM item parameters (top-1)")
```

```{r grm_discrim_plot, fig.height=5}
grm_params %>%
  ggplot(aes(x = reorder(item, a1), y = a1, fill = Scale)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~Scale, scales = "free_y") +
  labs(title = "GRM Discrimination Parameters (a1) by Item",
       subtitle = "Negative a1 = reverse-keyed items correctly loading in opposite direction",
       x = NULL, y = "Discrimination (a1)") +
  theme_minimal() +
  theme(legend.position = "none")
```

The discrimination parameters are large (|a1| = 3-37), reflecting the near-deterministic
response patterns. Signed a1 correctly captures reverse-keying direction:
negative-loading items correspond exactly to the reverse-keyed item set.


## 5. Results: Soft-Evidence IRT

### The discrimination collapse

```{r soft_params}
soft_params <- names(results) %>%
  map_dfr(function(s) {
    r <- results[[s]]
    if (is.null(r$grm_soft)) return(NULL)
    coefs <- coef(r$grm_soft, simplify = TRUE)$items
    as_tibble(coefs, rownames = "item") %>%
      mutate(Scale = s, .before = 1)
  })

soft_params %>%
  select(Scale, item, a1, d1, d2) %>%
  knitr::kable(digits = 3, caption = "Soft-evidence GRM item parameters — note a1 = 0 everywhere")
```

**All discrimination parameters collapse to zero.** The soft-evidence GRM cannot
recover a latent trait dimension because the logprob distributions are too homogeneous
across persons.

### Why soft evidence fails here

```{r soft_diagnostics, fig.height=5}
# Show the between-person variation in soft evidence for a sample scale (EXT)
ext_pi <- results[["BFI-EXT"]]$pi_sub
ext_items <- results[["BFI-EXT"]]$items

# For each person x item, compute entropy of the 3-cat distribution
entropy_data <- expand_grid(
  person = 1:N,
  item_idx = 1:length(ext_items)
) %>%
  mutate(
    item = ext_items[item_idx],
    p_low  = map2_dbl(person, item_idx, ~ ext_pi[.x, .y, 1]),
    p_mid  = map2_dbl(person, item_idx, ~ ext_pi[.x, .y, 2]),
    p_high = map2_dbl(person, item_idx, ~ ext_pi[.x, .y, 3])
  ) %>%
  rowwise() %>%
  mutate(
    entropy = {
      p <- c(p_low, p_mid, p_high)
      p <- p[p > 0]
      -sum(p * log(p))
    },
    normed_entropy = entropy / log(3)
  ) %>%
  ungroup()

entropy_data %>%
  ggplot(aes(x = p_high)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~item, nrow = 2) +
  labs(title = "Distribution of p(high) Across SPIDs — BFI Extraversion Items",
       subtitle = "Near-zero between-person variation: most SPIDs have p(high) ≈ 0.75–0.85",
       x = "p(high) = p4 + p5", y = "Count") +
  theme_minimal()
```

```{r entropy_hist, fig.height=4}
entropy_data %>%
  ggplot(aes(x = normed_entropy)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  geom_vline(xintercept = mean(entropy_data$normed_entropy),
             linetype = "dashed", color = "red") +
  labs(title = "Normalized Entropy of 3-Category Soft Evidence (EXT items)",
       subtitle = sprintf("Mean = %.3f — distributions are peaked, not uniform (1.0)",
                          mean(entropy_data$normed_entropy)),
       x = "Normalized Entropy (0=certain, 1=uniform)", y = "Count") +
  theme_minimal()
```

The problem is clear: soft evidence requires **between-person distributional variation**
to estimate item discrimination. With Llama-2's extreme ceiling effect, virtually all
SPIDs produce similar probability profiles (~5% low, ~15% mid, ~80% high). The d-parameters
capture these item-level base rates, but a1 goes to zero because there is nothing to
discriminate *between* persons.

### Soft-evidence fit comparison

```{r soft_summary}
soft_summary <- names(results) %>%
  map_dfr(function(s) {
    r <- results[[s]]
    tibble(
      Scale = s,
      GRM_soft_AIC = if (!is.null(r$grm_soft)) r$grm_soft@Fit$AIC else NA_real_,
      NRM_soft_AIC = if (!is.null(r$nrm_soft)) r$nrm_soft@Fit$AIC else NA_real_,
      Delta_AIC = GRM_soft_AIC - NRM_soft_AIC,
      Preferred = ifelse(Delta_AIC < 0, "GRM", "NRM")
    )
  })

soft_summary %>%
  knitr::kable(digits = 1, caption = "Soft-evidence model comparison (all degenerate)")
```

GRM is consistently preferred over NRM for soft evidence, but only because GRM uses
fewer parameters to describe the same intercept-only model. Both have a1 ≈ 0.


## 6. Comparing Top-1 vs Soft-Evidence GRM

```{r compare_top1_soft}
comparison <- names(results) %>%
  map_dfr(function(s) {
    r <- results[[s]]
    if (is.null(r$grm_top1) || is.null(r$grm_soft)) return(NULL)
    coefs_t <- coef(r$grm_top1, simplify = TRUE)$items
    coefs_s <- coef(r$grm_soft, simplify = TRUE)$items
    shared <- intersect(colnames(coefs_t), colnames(coefs_s))
    diff_mat <- coefs_s[, shared] - coefs_t[, shared]
    a_cor <- cor(coefs_t[, "a1"], coefs_s[, "a1"])

    tibble(
      Scale = s,
      mean_abs_diff = mean(abs(diff_mat), na.rm = TRUE),
      max_abs_diff = max(abs(diff_mat), na.rm = TRUE),
      a1_correlation = a_cor
    )
  })

comparison %>%
  knitr::kable(digits = 3,
               caption = "Parameter differences: soft-evidence vs top-1 GRM")
```

The correlation between top-1 and soft-evidence discrimination is near zero or negative —
confirming the soft-evidence models are not recovering the same latent structure.
This is expected given the a1=0 collapse.


## 7. Summary

### Key findings

| Finding | Detail |
|---------|--------|
| **Top-1 GRM works well** | Strong discrimination, correct reverse-keying direction |
| **GRM preferred for 3/5 traits** | CON, NEU, OPE by AIC/BIC |
| **NRM preferred for 2/5 traits** | EXT (strongly), AGR (AIC only) |
| **Soft evidence collapses** | All a1 = 0; no latent dimension recovered |
| **Root cause** | Ceiling effect: ~80% probability on "high", minimal between-person variation |
| **Category suppression confirmed** | Categories 2 and 4 *never* appear as argmax across 55K prompts |

### Implications

1. **Standard (top-1) IRT is viable** for LLM personality measurement, even with only
   3 effective response categories. The preamble-induced trait variation creates meaningful
   person-level differences that IRT can recover.

2. **Soft-evidence IRT needs models with more response variability.** The approach is
   theoretically sound (validated with simulated data in `mirt/tests/test_softevidence.R`),
   but Llama-2's near-deterministic responses leave too little distributional signal.

3. **Multi-model comparison** may be more informative for soft evidence — models with
   different ceiling tendencies would provide actual distributional variation.

4. **The 5-category collapse to 3** loses information. Future work could explore models
   that actually use categories 2 and 4 (perhaps with different prompt framing or
   less instruction-tuned models).

```{r session_info, echo=FALSE}
cat("Session info:\n")
cat(sprintf("  mirt version: %s\n", packageVersion("mirt")))
cat(sprintf("  R version: %s\n", R.version.string))
```
